---
title: 'ggplot charts'
subtitle: 'Theming'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "custom.css"]
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);" />
        </div>
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(palmerpenguins)
#data(package = 'palmerpenguins')
knitr::opts_chunk$set(echo = TRUE, comment = "")
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina=2, out.width="100%")
hrc <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/hrc/horizons_river_ecoli.csv")

xaringanExtra:::use_freezeframe()
```

## Theming

The entire layout of a `ggplot2` chart can be customised by the `theme`:

- Background colours
- Size of panel
- All text size, font, style, placement, justification, angle etc
- Space between facets, facet header styles

Basically everything you see on the plot can be customised.

The **easiest** way to customise the plot is to choose a pre-made theme. (Let someone else do the work!)

---

.left-code[
## Themes

```{r theme1, warning=FALSE, eval=FALSE}
ggplot(penguins) +
  geom_density(
    aes(x=body_mass_g,
        fill=species),
    alpha=0.5)
```

The standard theme is `theme_grey`.
]

.right-plot[
```{r, ref.label="theme1", warning=FALSE, echo=FALSE}
```
]

---

.left-code[
## Themes

```{r theme2, warning=FALSE, eval=FALSE}
ggplot(penguins) +
  geom_density(
    aes(x=body_mass_g,
        fill=species),
    alpha=0.5) +
  theme_bw()
```

]

.right-plot[
```{r, ref.label="theme2", warning=FALSE, echo=FALSE}
```
]

---

.left-code[
## Themes

```{r theme3, warning=FALSE, eval=FALSE}
ggplot(penguins) +
  geom_density(
    aes(x=body_mass_g,
        fill=species),
    alpha=0.5) +
  theme_light()
```
]

.right-plot[
```{r, ref.label="theme3", warning=FALSE, echo=FALSE}
```
]

---

.left-code[
## Themes

```{r theme4, warning=FALSE, eval=FALSE}
ggplot(penguins) +
  geom_density(
    aes(x=body_mass_g,
        fill=species),
    alpha=0.5) +
  theme_dark()
```
]

.right-plot[
```{r, ref.label="theme4", warning=FALSE, echo=FALSE}
```
]

---

.left-code[
## Themes

```{r theme5, warning=FALSE, eval=FALSE}
ggplot(penguins) +
  geom_density(
    aes(x=body_mass_g,
        fill=species),
    alpha=0.5) +
  theme_minimal()
```
]

.right-plot[
```{r, ref.label="theme5", warning=FALSE, echo=FALSE}
```
]

---

.left-code[
## Themes

```{r theme6, warning=FALSE, eval=FALSE}
ggplot(penguins) +
  geom_density(
    aes(x=body_mass_g,
        fill=species),
    alpha=0.5) +
  theme_minimal(base_size=16)
```

The `base_size` argument sets the default text size.
]

.right-plot[
```{r, ref.label="theme6", warning=FALSE, echo=FALSE}
```
]

---

.left-code[
## Themes

```{r theme7, warning=FALSE, eval=FALSE}
ggplot(penguins) +
  geom_density(
    aes(x=body_mass_g,
        fill=species),
    alpha=0.5) +
  theme_minimal() +
  theme(legend.position = 'bottom')
```

The `theme()` function is how you alter elements of an existing theme.
]

.right-plot[
```{r, ref.label="theme7", warning=FALSE, echo=FALSE}
```
]

---

class: middle, inverse

# Example of story telling with data

https://www.massey.ac.nz/~jcmarsha/stories/story_of_a_chart.html

---

## Key points

- Alter the look of plots with the various `theme_*()` functions, and use `theme()` to fine tune.
