---
title: 'ggplot charts'
subtitle: 'Combining different charts with patchwork'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "custom.css"]
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);" />
        </div>
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(palmerpenguins)
#data(package = 'palmerpenguins')
knitr::opts_chunk$set(echo = TRUE, comment = "")
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina=2, out.width="100%")
hrc <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/hrc/horizons_river_ecoli.csv")

xaringanExtra:::use_freezeframe()
```

## Combining different kinds of plot

Sometimes we want to combine different types of plot. e.g. perhaps we want
to compare the body mass distribution of penguins as well as show the correlation between bill length and depth.

We can do the separate plots and can combine them using the `patchwork` library

---

.pull-left[
```{r, warning=FALSE}
ggplot(penguins) +
  geom_density(
    aes(x=body_mass_g,
        fill=species),
    alpha=0.5)
```
]

.pull-right[
```{r, warning=FALSE}
ggplot(penguins) +
  geom_point(aes(x=bill_length_mm,
                 y=bill_depth_mm,
                 col=species),
               alpha=0.5)
```
]

---

## Combining plots with `patchwork`

With `patchwork` we first **save** the plots into separate variables, then use `+` or `/` to combine them together.

When you save a `ggplot2` plot, you can 'print' it out by just typing the name. So your code will look like:

```{r, eval=FALSE}
g1 = ggplot(data1) + ...
g2 = ggplot(data2) + ...

g1 # To look at the first plot
g2 # To look at the second plot

library(patchwork)
g1 + g2 # To put graphs side by side
g1 / g2 # To put graphs top and bottom
```

---

.left-code[
## Combining plots with `patchwork`

```{r patch1, eval=FALSE}
library(patchwork)
g1 = ggplot(penguins) +
  geom_density(
    aes(x=body_mass_g,
        fill=species),
    alpha=0.5)

g2 = ggplot(penguins) +
  geom_point(
    aes(x=bill_length_mm,
        y=bill_depth_mm,
        col=species),
    alpha=0.5)

g1 / g2
```
]

.right-plot[
```{r, ref.label="patch1", echo=FALSE, warning=FALSE}
```
]

---

## Combining plots with `patchwork`

Use `+` to put things side by side:

```{r, warning=FALSE, fig.dim=c(8,3), fig.retina=3}
g1 + g2
```

---

## Combining plots with `patchwork`

The `plot_layout()` function is useful to control things a little.

```{r, warning=FALSE, fig.dim=c(8,3), fig.retina=3}
g1 + g2 + plot_layout(guides = 'collect')
```

---

The `plot_annotation()` function allows adding titles, subtitles to the total layout.

```{r, warning=FALSE, fig.dim=c(8,3), fig.retina=3}
g1 + g2 + plot_layout(guides = 'collect') +
  plot_annotation(title = "Gentoo penguins are chonky and have long, narrow beaks")
```

---

Lots more can be done. See https://patchwork.data-imaginist.com/index.html for more!

```{r, echo=FALSE, warning=FALSE}
g0 = ggplot(penguins) + geom_density(aes(x=flipper_length_mm, fill=species), alpha=0.5)
```

```{r, warning=FALSE, fig.dim=c(8,3.5), fig.retina=3}
(g0 / g1 | g2) + plot_layout(guides = 'collect') +
  plot_annotation(title = "Gentoo penguins are chonky and have long, narrow beaks")
```

---

## Key points

- Use `patchwork` to layout multiple different charts on the same page.

- Think about which guides can be combined, and add overall titles with `plot_annotation`.
