---
title: 'ggplot charts'
subtitle: 'Small multiple plots'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "custom.css"]
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);" />
        </div>
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(palmerpenguins)
#data(package = 'palmerpenguins')
knitr::opts_chunk$set(echo = TRUE, comment = "")
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina=2, out.width="100%")
hrc <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/hrc/horizons_river_ecoli.csv")

xaringanExtra:::use_freezeframe()
```

## Small multiple plots

With `dplyr` we can use `group_by()` to do summarising and filtering by subgroups of the data.

We can do the same thing when we produce plots with `ggplot2`.

The idea is that often we want to see trends or patterns between some variables, but often those trends or patterns may vary by subgroup.

In that case we might want to plot them separately for each subgroup so
we can compare and contrast.

---

## Small multiple plots

One way to do this is to use `dplyr` to filter down to the subgroup that we want to compare and then produce separate plots.

e.g. suppose we want to compare two sites on the Manawatu river for their E.coli counts over time. The river at Site80 (down by Waitoetoe park, opposite where the Turitea Stream comes out) and the river at Site15 (by the old teacher's college in Hokowhitu):

```{r}
waitoetoe <- hrc |>
  filter(Site == "manawatu at u/s pncc stp",
         Date >= "2008-01-01") |>
  select(Site, Date, Value)

hokowhitu <- hrc |>
  filter(Site == "manawatu at teachers college",
         Date >= "2008-01-01") |>
  select(Site, Date, Value)
```

---

## Comparing sites

.pull-left[
```{r}
waitoetoe |>
  summarise(Median = median(Value),
            Max = max(Value))
```
]

.pull-right[
```{r}
hokowhitu |>
  summarise(Median = median(Value),
            Max = max(Value))
```
]

In terms of the median and maximum these sites are quite similar, which makes sense - they're only about 3km from each other.

We could do a plot of each one and compare.
---

.pull-left[
```{r}
ggplot(data=waitoetoe) +
  geom_line(aes(x=Date, y=Value)) +
  scale_y_log10(
    labels=scales::label_comma()
    )
```
]

.pull-right[
```{r}
ggplot(data=hokowhitu) +
  geom_line(aes(x=Date, y=Value)) +
  scale_y_log10(
    labels=scales::label_comma()
    )
```
]

---

## Combining plots

The problem here is that each uses different scale. So while the patterns look kinda similar, it's hard to really
know the level as the y-axis scales are a bit different.

We could change the axes to force them to be the same using the `limits` argument to `scale_y_log10()`, but there's
quite a lot of repetition.

And, if we want them on the same scale anyway, why do we need separate y-axis scales? We could just place them side
by side and remove the additional axis space.

---

.pull-left[
```{r}
ggplot(data=waitoetoe) +
  geom_line(aes(x=Date, y=Value)) +
  scale_y_log10(
    labels=scales::label_comma(),
    limits = c(1,20000)
    )
```
]

.pull-right[
```{r}
ggplot(data=hokowhitu) +
  geom_line(aes(x=Date, y=Value)) +
  scale_y_log10(
    labels=scales::label_comma(),
    limits = c(1,20000)
    )
```
]

---

## Combining plots

The better way to do it is to combine the two datasets and let `ggplot2` split out the two groups for us when it creates the plot.

Essentially we work the same way as we would with `dplyr` to summarise: leave the data in a single data.frame.

```{r}
two_sites <- hrc |>
  filter(Site %in% c("manawatu at u/s pncc stp", "manawatu at teachers college"),
         Date >= "2008-01-01")

two_sites |> group_by(Site) |>
  summarise(Median = median(Value), Max = max(Value))
```

---

.left-code[
## Combining plots

```{r facet1, eval=FALSE}
ggplot(data=two_sites) +
  geom_line(
    mapping = aes(x=Date, y=Value,
                  col=Site)
    ) +
  scale_y_log10(
    labels = scales::label_comma()
  )
```

We could use colour to distinguish the sites
]

.right-plot[
```{r, ref.label="facet1", echo=FALSE}
```
]

---

.left-code[
## Combining plots

```{r facet2, eval=FALSE}
ggplot(data=two_sites) +
  geom_line(
    mapping = aes(x=Date, y=Value)
    ) +
  scale_y_log10(
    labels = scales::label_comma()
  ) +
  facet_wrap(vars(Site)) #<<
```

But having separate plots is better.
]

.right-plot[
```{r, ref.label="facet2", echo=FALSE}
```
]

---

.left-code[
## Combining plots

```{r facet3, eval=FALSE}
ggplot(data=two_sites) +
  geom_line(
    mapping = aes(x=Date, y=Value),
    alpha=0.5
    ) +
  geom_smooth(
    mapping = aes(x=Date, y=Value)
  ) +
  scale_y_log10(
    labels = scales::label_comma()
  ) +
  facet_wrap(vars(Site))
```

The trend shows that the site at Waitoetoe park used to be
worse than at Hokowhitu, but now is more similar.

They're now both worse?
]

.right-plot[
```{r, ref.label="facet3", echo=FALSE}
```
]

---

## Combining plots

We can now easily generalise to more plots. e.g. what is happening with the turitea stream which flows into the Manawatu near Waitoetoe Park?

Unfortunately this was measured only since 2013, but nonetheless might be interesting

```{r}
three_sites <- hrc |> filter(
  Site %in% c("manawatu at u/s pncc stp",
              "manawatu at teachers college",
              "turitea at no1 dairy"),
  Date >= "2008-01-01"
)
three_sites |> group_by(Site) |>
  summarise(Median = median(Value), Max = max(Value))
```

A higher median but lower maximum.

---

.left-code[
## Combining plots

```{r facet4, eval=FALSE}
ggplot(data=three_sites) +
  geom_line(
    mapping = aes(x=Date, y=Value),
    alpha=0.5
    ) +
  geom_smooth(
    mapping = aes(x=Date, y=Value)
  ) +
  scale_y_log10(
    labels = scales::label_comma()
  ) +
  facet_wrap(vars(Site))
```

We can't tell if the Turitea stream is the
reason for the increased counts pre 2013 as
we don't have the measures then.

The counts post 2013 are higher in the Turitea
stream than in the river (more dilution in the river?)
]

.right-plot[
```{r, ref.label="facet4", echo=FALSE}
```
]

---

.left-code[
## Combining plots

```{r facet5, eval=FALSE}
ggplot(data=three_sites) +
  geom_line(
    mapping = aes(x=Date, y=Value),
    alpha=0.5
    ) +
  geom_smooth(
    mapping = aes(x=Date, y=Value)
  ) +
  scale_y_log10(
    labels = scales::label_comma()
  ) +
  facet_wrap(vars(Site), ncol=1) #<<
```

Can specify the number of rows or columns.

This allows us to see the correlation in the two Manawatu measures.

But is harder to assess overall levels.
]

.right-plot[
```{r, ref.label="facet5", echo=FALSE}
```
]

---

.left-code[
## Combining plots

```{r peng1, eval=FALSE}
ggplot(penguins) + 
  geom_density(aes(x=body_mass_g),
               fill='light blue') +
  facet_grid( #<<
    rows = vars(sex), #<<
    cols = vars(species) #<<
  ) #<<
```

If we have two variables that we'd like to use to split the data, we can use a grid of small plots, with rows representing one variable and columns the other.

The `NA` for `sex` though is a bit odd!
]

.right-plot[
```{r, ref.label="peng1", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Combining plots

```{r peng2, eval=FALSE}
penguins |> #<<
  filter(!is.na(sex)) |> #<<
ggplot() + 
  geom_density(aes(x=body_mass_g),
               fill='light blue') +
  facet_grid(
    rows = vars(sex),
    cols = vars(species)
  )
```

For densities, where the `x` axis is our comparator, we can
easily assess down columns.

i.e. Males are heavier than females for each species, less pronounced in Chinstrap?
]

.right-plot[
```{r, ref.label="peng2", echo=FALSE}
```
]

---

.left-code[
## Combining plots

```{r peng3, eval=FALSE}
penguins |>
  filter(!is.na(sex)) |>
ggplot() + 
  geom_density(aes(x=body_mass_g),
               fill='light blue') +
  facet_grid(
    rows = vars(species), #<<
    cols = vars(sex) #<<
  )
```

For densities, where the `x` axis is our comparator, we can
easily assess down columns.

Switching roles of variables means we can easily compare within
males and females separately.

e.g. we see that male Adelie and Chinstraps have similar weight distributions,
while females differ.
]

.right-plot[
```{r, ref.label="peng3", echo=FALSE}
```
]

---

.left-code[
## Combining plots

```{r peng4, eval=FALSE}
penguins |>
  filter(!is.na(sex)) |>
ggplot() + 
  geom_density(aes(x=body_mass_g,
                   fill=species),
               alpha=0.5) + #<<
  facet_wrap(vars(sex), #<<
             ncol=1) #<<
```

In this example the small number of groups allows overlapping
densities to be a little better perhaps.
]

.right-plot[
```{r, ref.label="peng4", echo=FALSE}
```
]

---

## Key points

- Use `facet_wrap()` or `facet_grid()` for when you want repeated graphs on subgroups.

- Think about rows versus columns: if the thing to compare is on the x-axis, then stacking the charts vertically is best.

- Sometimes using `fill` or `col` is a better option.
