---
title: 'Data Manipulation with dplyr'
subtitle: 'The pipe'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "custom.css"]
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);" />
        </div>
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(palmerpenguins)
#data(package = 'palmerpenguins')
knitr::opts_chunk$set(echo = TRUE, comment = "")
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina=2, out.width="100%")
```

## Data today

We'll use the data from Horizons Regional council on river quality again today.

```{r, echo=FALSE, message=FALSE}
hrc <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/hrc/horizons_river_ecoli.csv")
hrc
```

These data are available from Land, Air, Water Aoteoroa, https://www.lawa.org.nz/download-data/

---

## A pipeline of verbs

Each of the verbs in `dplyr` take a dataset in as the first parameter, and return a data frame out again.

Often, an analyses will consist of a bunch of verbs called one after the other, with each one feeding into the next.

```{r, eval=FALSE}
large_values <- filter(hrc, Value > 50000)
arrange(large_values, Site, Date)
```

The temporary dataset `large_values` is really only used here to feed to the `arrange()` verb. We don't really need it.

We could do:

```{r, eval=FALSE}
arrange(filter(hrc, Value > 50000), Site, Date)
```

But you have to read that 'inside out'!

---

class: middle, inverse

# The pipe `|>`

---

## The pipe `|>`

The pipe operator, `|>` is specifically designed to improve this. It takes the item on the left of the pipe and places it
as the first argument of the function on the right.

```{r, eval=FALSE}
hrc |> filter(Value > 5000)
```

Is equivalent to:

```{r, eval=FALSE}
filter(hrc, Value > 5000)
```

But, the pipe version is readable as "Take the dataset `hrc` and filter it so that rows with Value > 5000 remain"

To enter a pipe in RStudio, you can use Ctrl-Shift-M (Command-Shift-M on a Mac).

---

## The pipe `|>`

This extends naturally:

```{r, eval=FALSE}
hrc |>
  filter(Value > 5000) |>
  arrange(Site, Date)
```

reads "Take the dataset `hrc`, filter it so that rows with Value > 5000 remain, then arrange by Site and Date"

When constructing pipelines like this, make sure the `|>` symbol is at the end of intermediate lines.

.left-code[
Use this:

```{r, eval=FALSE}
hrc |>
  filter(Value > 5000) |>
  arrange(Site, Date)
```
]

.right-plot[
Not this:

```{r, eval=FALSE}
hrc
 |> filter(Value > 5000)
 |> arrange(Site, Date)
```
]

---

## The pipe `|>`

The pipe also gives us **code completion** by pressing the TAB key.

e.g. type:

```{r, eval=FALSE}
hrc |> select(sit
```

and press the Tab key brings up a list of symbols starting with `sit`, which includes `Site` and `SiteID`.

This allows you to make sure you get the spelling correct :)

---

## The magrittr pipe `%>%`

Prior to R version 4.1, the `|>` (base) pipe didn't exist. Instead, people used a pipe from the `magrittr` package which uses `%>%`.

Generally, these are basically the same. You will see `%>%` used 'in the wild' when googling - that's OK, it'll work just fine.

In RStudio, by default the keyboard combination Ctrl-Shift-M (Cmd-Shift-M on Mac) inserts the magrittr pipe `%>%`.

**We suggest you change this in RStudio**

See Tools->Global Options->Code and select the 'use native pipe operator, |>' option.

---

## Using the pipe in other places

We can use the pipe anytime to take something and insert it
as the first argument of a function.

e.g. `ggplot()` has `data` as it's first argument, so using
the pipe to pass a dataset in is useful.

```{r}
hrc |>
  filter(Value > 5000) |>
  ggplot() +
  geom_point(mapping=aes(x=Date, y=Value)) +
  scale_y_log10()
```

```{r}
hrc |> filter(str_detect(Site, "manawatu")) |>
  ggplot() +
  geom_boxplot(mapping=aes(y=Site, x=Value)) +
  scale_x_log10()
```

Remember to switch from `|>` to `+` !!

---

## Key ideas

- The pipe `|>` is useful for tidying up the order of our code.

- It takes the thing on the left and puts it into the function
on the right as the first argument.

- It works best when you KNOW what the function is expecting as
it's first argument.

- All `dplyr` verbs take data as first argument and return a
new version of the data so that piping works naturally.
