---
title: 'Data Manipulation with dplyr'
subtitle: 'select and mutate'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "custom.css"]
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);" />
        </div>
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(palmerpenguins)
#data(package = 'palmerpenguins')
knitr::opts_chunk$set(echo = TRUE, comment = "")
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina=2, out.width="100%")
```

## Data today

We'll use the data from Horizons Regional council on river quality again today.

```{r, echo=FALSE, message=FALSE}
hrc <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/hrc/horizons_river_ecoli.csv")
hrc
```

These data are available from Land, Air, Water Aoteoroa, https://www.lawa.org.nz/download-data/

---

## `dplyr`

Just like `ggplot2` is a grammar of graphics, `dplyr` is a grammar of data manipulation.

It consists of a set of verbs that help us solve the most common data manipulation challenges:

- `filter` to choose observations (rows) by value.
- `arrange` changes the order of rows.
- `select` to choose variables (columns) by name.
- `mutate` to add new variables based on existing variables.
- `summarise` to reduce multiple values down to a single summary.

All of these combine with `group_by` to do all the above operation on subgroups of the data.

---

## Selecting and renaming columns

Just like we can select rows with `filter`, we can choose columns with `select`.

This is useful for getting rid of columns we don't need, for rearranging columns, and for changing names to
something more convenient.

```{r}
hrc |> select(Site, Date, Value)
```

---

## Selecting columns with `select`

We can select a range of columns with a colon `:`

```{r}
hrc |> select(Council, Site:RawValue)
```

---

## Selecting columns with `select`

We can select by column number instead of column name if you like:

```{r}
hrc |> select(Council, 3:6)
```

Generally it's better to use names though, as the order of columns might be different to what you expect.

---

## Selecting columns with `select`

The `everything()` helper function is useful if you want one column first, then everything else:

```{r}
hrc |> select(Symbol, Value, everything())
```

---

## Selecting columns with `select`

The `starts_with()` helper is useful if you have a bunch of columns that start similarly:

```{r}
hrc |> select(Date, Value, starts_with("Site"))
```

---

## Selecting columns with `select`

You can rename a column just by changing it's name in the `select`:

```{r}
hrc |> select(Site, Date, EColi = Value)
```

---

## Dropping columns

To drop a column use `-` in front of the name:

```{r}
hrc |> select(-Indicator, -RawValue, -Symbol)
```

---

## Renaming columns with `rename`

If you want to rename a column but don't want to specify all the other columns, then `rename()` does this:

```{r}
hrc |> rename(Name = Site)
```

---

## Adding new columns: `mutate`

We can add new columns based on existing ones with `mutate`:

```{r}
hrc |> select(SiteID, Date, Value) |>
  mutate(Log10Value = log10(Value))
```

---

## Adding new columns: `mutate`

We can add new columns based on existing ones with `mutate`:

```{r, message=FALSE}
library(lubridate)
hrc |> select(SiteID, Date, Value) |>
  mutate(Year = year(Date), Month = month(Date, label=TRUE))
```

---

.left-code[
## Seasonality?
```{r, label='hrc_season', eval=FALSE}
hrc |>
  mutate(
    Month = month(Date, label=TRUE)
    ) |>
  ggplot() +
  geom_boxplot(
    aes(
      x = Month,
      y = Value)
    ) +
  scale_y_log10(
    labels = scales::label_comma()
  )
```

We can pipe the result of wrangling into
`ggplot()` as it takes a data.frame as first
argument.

Remember to switch from `|>` to `+` !!
]

.right-plot[
```{r, ref.label='hrc_season', echo=FALSE, eval=TRUE}
```
]

---

## Adding new columns: `mutate`

We can change existing columns by overwriting them:

```{r}
library(lubridate)
hrc |> select(SiteID, Date, Value) |>
  mutate(SiteID = as.numeric(SiteID))
```

---

## Adding new columns: `mutate`

We can change existing columns by overwriting them:

```{r}
hrc |> select(SiteID, Site, Date, Value) |>
  mutate(SiteID = as.numeric(SiteID),
         Site = str_to_title(Site))
```

---

## How I extracted value, symbol

.left-code-wide[
```{r, label="hrc_extract", eval=FALSE}
hrc |> select(SiteID, RawValue) |>
  mutate(
    Symbol = str_extract(RawValue, '[<>]'),
    Value = str_extract(RawValue, '[0-9]+'),
    Value = as.numeric(Value)
    ) |>
  arrange(Symbol)
```

The patterns `[<>]`, `[0-9]+` are **regular expressions**.

They're a way of specifying how to match parts of text.

Don't worry about these - you won't have to come up with them yourselves!
]

.right-out-narrow[
```{r, ref.label="hrc_extract", echo=FALSE, eval=TRUE}
```
]

---

## Key ideas

So far we have learnt:

- `filter()` to pick rows based on value.

- `arrange()` to change the order of rows based on value.

- `select()` to pick columns based on column name.

- `rename()` to rename columns while keeping others.

- `mutate()` to add new columns based on existing ones.
