---
title: 'Data Manipulation with dplyr'
subtitle: 'filter and arrange'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "custom.css"]
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);" />
        </div>
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(palmerpenguins)
#data(package = 'palmerpenguins')
knitr::opts_chunk$set(echo = TRUE, comment = "")
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina=2, out.width="100%")
```

## Data today

We'll use the data from Horizons Regional council on river quality again today.

```{r, echo=FALSE, message=FALSE}
hrc <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/hrc/horizons_river_ecoli.csv")
hrc
```

These data are available from Land, Air, Water Aoteoroa, https://www.lawa.org.nz/download-data/

---

## Data wrangling

You'll have noticed in that we have introduced (in passing) a bunch of code to wrangle data into a form that is usable for charting or tables.

Much of the data wrangling I do daily (and it is daily!) can be done using only a handful of functions, the majority taken from the package `dplyr`, with a few extras from other packages such as `tidyr`.

Essentially, the `dplyr` package provides a bunch of "verbs" (functions) for transforming data. It is similar to SQL if you have done any database work.

It is part of the `tidyverse` set of packages: https://www.tidyverse.org

---

class: middle, inverse

# `dplyr`

---

## `dplyr`

Just like `ggplot2` is a grammar of graphics, `dplyr` is a grammar of data manipulation.

It consists of a set of verbs that help us solve the most common data manipulation challenges:

- `filter` to choose observations (rows) by value.
- `arrange` changes the order of rows.
- `select` to choose variables (columns) by name.
- `mutate` to add new variables based on existing variables.
- `summarise` to reduce multiple values down to a single summary.

All of these combine with `group_by` to do all the above operation on subgroups of the data.

---

## `dplyr`

One of the nice things about `dplyr` is it abstracts the language of data processing across the implementation of where the data exist.

It works with `data.frame`'s loaded into memory, as well as remote databases (via SQL, or Apache Spark) using the same set of R commands.

The other nice thing is the consistency:
 - All verbs take in as their first argument a `data.frame`.
 - Then all other arguments describe what to do with the `data.frame`.
 - The result is then a new `data.frame`.

```{r eval=FALSE}
new_data <- do_something(old_data, other, stuff)
```
 
---

## Filtering rows with `filter`
 
Let's look again at the river data:
 
```{r}
hrc
```

---

## Filtering rows with `filter`
 
What if we just wanted the rows from 2017 onwards?
 
```{r}
filter(hrc, Date >= '2017-01-01')
```

---

## Filtering rows with `filter`
 
What if we wanted rows with large counts?

```{r}
filter(hrc, Value > 100000)
```

---

## Filtering rows with `filter`
 
What if we wanted really low counts before 2010?

```{r}
filter(hrc, Value < 100, Date <= '2010-12-31')
```

---

## Filtering rows with `filter`
 
What if we wanted the extreme low or high counts?

```{r}
filter(hrc, Value < 3 | Value > 100000)
```

---

## Logical expressions in R

The `filter` argument is a logical expression. i.e. it's something that evaluates to `TRUE` or `FALSE`.

For convenience, filter can take more than one logical expression, separated by a comma. In this case they're AND'd together (i.e. rows must satisfy both criteria), but anything within each expression is done first.

If we want to use OR rather than AND, we use the vertical bar `|` character.

i.e. `filter(hrc, Value < 3 | Value > 100000, SiteID == "00081")`

will return rows where the value is very small or very large from Site 00081.

---

## Logical expressions in R

Notice that for 'is equal to' we've been using `==`. The single `=` is for assignment of parameters instead. `dplyr` will warn you about this:

```{r, error=TRUE}
filter(hrc, Value = 4)
```

**This is rare** - usually error messages are rather more opaque!

Reading carefully, and searching the error on google can sometimes help.

---

## Logical expressions in R

Other operations are `!=` (not equal to) and `%in%` (in). e.g.

```{r}
filter(hrc, Value %in% c(3,5))
```

---

## Dealing with missing values

Notice the `Symbol` column is full of `NA` which means "missing". Maybe there are some that aren't missing? We might try:

```{r}
filter(hrc, Symbol != NA)
```

This seems like it might have worked?? It says there are no rows returned...

--

```{r}
filter(hrc, Symbol == NA)
```

Hmm, that doesn't seem right??

---

## Dealing with missing values

.left-code[
The key thing to remember is if a value is missing, then comparing it is senseless:

```{r}
2 == NA
NA == NA
```

Is 2 equal to something missing?

Maybe! We don't know! So the answer is missing.
]

--
.right-plot[
We use `is.na()` instead for this.

```{r}
is.na(2)
is.na(NA)
```
]

RStudio will remind you about this with a yellow warning in the margin, as it's a common problem.
---

## Dealing with missing values

Notice the `Symbol` column is full of `NA` which means "missing". Maybe there are some that aren't missing?

```{r}
filter(hrc, !is.na(Symbol))
```

Ok, it seems like we have a few (`r hrc |> filter(!is.na(Symbol)) |> nrow()`) numbers where the count actually means less than or greater than the value given, rather than equal to the value given.

---

## Arranging rows with `arrange`

Rather than pick rows out, we can sort the data instead.

```{r}
arrange(hrc, Value)
```

---

## Arranging rows with `arrange`

Selecting more than one variable to arrange on will first arrange on the first, then use later variables for ties:

```{r}
arrange(hrc, Value, Date)
```

---

## Arranging rows with `arrange`

To arrange descending rather than ascending, we can use `desc`

```{r}
arrange(hrc, Value, desc(Date))
```

Interestingly, no value of 1 after 2012... This suggests a change in the way things were measured.

---

## Combining `filter` and `arrange`

Let's look at post-2012 values:

```{r}
after2012 <- filter(hrc, Date >= '2013-01-01')
arrange(after2012, Value)
```

It seems some of the values recorded as 4 actually mean "less than 4".

---

## Combining `filter` and `arrange`

```{r}
large_values <- filter(hrc, Value > 50000)
arrange(large_values, Site, Date)
```

---

## Key ideas

Today we've learnt about:

- `filter()` to pick rows based on value.

- `arrange()` to change the order of rows based on value.
