---
title: 'Tidying data'
subtitle: 'Reordering groups'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "custom.css"]
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);" />
        </div>
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(palmerpenguins)
#data(package = 'palmerpenguins')
knitr::opts_chunk$set(echo = TRUE, comment = "")
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina=2, out.width="100%")
hrc <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/hrc/horizons_river_ecoli.csv")

covid1 <- read_csv("http://www.massey.ac.nz/~jcmarsha/data/covid_au_nz.csv")
```

## Reordering groups

A common problem when presenting data in the form of tables or charts is ensuring that
order is correct.

By default, if an order isn't specified, `ggplot2` will present things alphabetically. This is almost certainly not what you want.

You can reorder groups by specifying that they are a **factor** rather than a **character** type.

The `forcats` library is useful. In particular the functions:

- `as_factor()` uses the order in the data (i.e. first group to appear goes first).

- `fct_relevel()` allows reordering in any way you like.

---

.left-code[

## Reordering groups

```{r reorder1, eval=FALSE}
covid1 |>
  mutate(rate = cases/tests) |>
  ggplot() +
  geom_col(aes(x=month,
               y=rate,
               fill=country),
           position='dodge') +
  scale_y_continuous(
    labels = scales::label_percent()
    ) +
  labs(
    x=NULL,
    y="Cases per COVID-19 test",
    fill=NULL
    ) +
  theme(legend.position = 'top')
```
]

.right-plot[
```{r, ref.label="reorder1", echo=FALSE}
```
]

---

.left-code[
## Reordering groups

```{r reorder2, eval=FALSE}
covid1 |>
  mutate(rate = cases/tests) |>
  ggplot() +
  geom_col(aes(x=as_factor(month), #<<
               y=rate,
               fill=country),
           position='dodge') +
  scale_y_continuous(
    labels = scales::label_percent()
    ) +
  labs(
    x=NULL,
    y="Cases per COVID-19 test",
    fill=NULL
    ) +
  theme(legend.position = 'top')
```

The `as_factor()` command will use the order
present in the data.
]

.right-plot[
```{r, ref.label="reorder2", echo=FALSE}
```
]

---

.left-code[
## Reordering groups

```{r reorder2b, eval=FALSE}
covid1 |>
  mutate(rate = cases/tests) |>
  ggplot() +
  geom_col(aes(x=as_factor(month),
               y=rate,
               fill=country),
           position='dodge') +
  scale_y_continuous(
    labels = scales::label_percent()
    ) +
  labs(
    x=NULL,
    y="Cases per COVID-19 test",
    fill=NULL
    ) +
  theme(legend.position = 'top')
```

The `as_factor()` command will use the order
present in the data.
]

.right-plot[
```{r, ref.label="reorder2b", echo=FALSE}
```
]

---
.left-code[
## Reordering groups

```{r reorder3, eval=FALSE}
covid1 |>
  mutate(rate = cases/tests) |>
  mutate( #<<
    country=fct_relevel(country, #<<
                    "New Zealand", #<<
                    "Australia") #<<
  ) |> #<<
  ggplot() +
  geom_col(aes(x=as_factor(month),
               y=rate,
               fill=country),
           position='dodge') +
  scale_y_continuous(
    labels = scales::label_percent()
    ) + ...
```

The `fct_relevel()` command lets you specify an order.
]

.right-plot[
```{r, echo=FALSE}
covid1 |>
  mutate(rate = cases/tests) |>
  mutate(
    country=fct_relevel(country,
                    "New Zealand",
                    "Australia")
  ) |>
  ggplot() +
  geom_col(aes(x=as_factor(month),
               y=rate,
               fill=country),
           position='dodge') +
  scale_y_continuous(
    labels = scales::label_percent()
    ) +
  labs(
    x=NULL,
    y="Cases per COVID-19 test",
    fill=NULL
    ) +
  theme(legend.position = 'top')
```
]

---
.left-code[
## Reordering groups

```{r reorder4, eval=FALSE}
covid1 |>
  mutate(rate = cases/tests) |>
  mutate( #<<
    country=fct_reorder(country,#<<
                        rate) #<<
  ) |> #<<
  ggplot() +
  geom_col(aes(x=as_factor(month),
               y=rate,
               fill=country),
           position='dodge') +
  scale_y_continuous(
    labels = scales::label_percent()
    ) + ...
```

The `fct_reorder()` command reorders the groups
based on the (median) values of another column.

Reordering by `rate` puts New Zealand first.
]

.right-plot[
```{r, echo=FALSE}
covid1 |>
  mutate(rate = cases/tests) |>
  mutate(
    country = fct_reorder(country, rate)
  ) |>
  ggplot() +
  geom_col(aes(x=as_factor(month),
               y=rate,
               fill=country),
           position='dodge') +
  scale_y_continuous(
    labels = scales::label_percent()
    ) +
  labs(
    x=NULL,
    y="Cases per COVID-19 test",
    fill=NULL
    ) +
  theme(legend.position = 'top')
```
]

---

## Key points

- Reorder grouping variables with `as.factor()`, `as_factor()`, `fct_relevel()` or `fct_reorder()`.

- `as.factor` orders alphabetically.

- `as_factor` orders based on the data order.

- `fct_relevel` allows you to order things however you like.

- `fct_reorder` orders according to another variable.
