---
title: "297.101 Week 11: Multiple Regression and Factors"
subtitle: "Adjusted comparisons and predictions"
author: Dr. Nick Knowlton
format:
  revealjs:
    code-copy: hover
    embed-resources: true
    css: "knowlton.css"
    theme: sky
    slide-number: true
    incremental: true
    smaller: true
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-svg.js
    highlight-style: breezedark
    highlight-lines: true
    progress: true
    logo: graphics/660491-ls_black_whitoutunz.png
execute:
  output: true
  eval: true
  code-fold: true
editor:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)
library(knitr)
library(ggplot2)
theme_set(theme_bw(base_size = 15))

# Quarto may execute from the repository root or the lecture directory.
vehicle_file <- if (file.exists("labs/nz_vehicles_2025.csv")) {
  "labs/nz_vehicles_2025.csv"
} else {
  "../labs/nz_vehicles_2025.csv"
}

vehicles <- readr::read_csv(vehicle_file, show_col_types = FALSE) |>
  mutate(
    station_wagon = factor(
      if_else(body_type == "STATION WAGON", "Station wagon", "Other"),
      levels = c("Station wagon", "Other")
    ),
    body_type = factor(
      body_type,
      levels = c(
        "STATION WAGON", "HATCHBACK", "SALOON",
        "SPORTS CAR", "CONVERTIBLE"
      ),
      labels = c(
        "Station wagon", "Hatchback", "Sedan",
        "Sports car", "Convertible"
      )
    )
  )
```

## Learning outcomes

By the end of this lecture, you should be able to:

- fit a model with a numerical predictor and a categorical predictor;
- interpret an adjusted slope and a difference from a reference category;
- explain why an adjusted comparison can differ from a raw comparison;
- make predictions at a common value of the numerical predictor; and
- compare ordinary and adjusted R² across models fitted to the same response.

# Return to the vehicle problem {.center}

## One predictor was a start

Last week we related fuel use to engine size, power or mass, one at a time. A fitted line described the average trend, but vehicles with the same engine size can be quite different.

Does body type account for some of that difference? We can add it to the model while keeping engine size in the equation.

## What are we comparing?

Each row is one 2025 petrol vehicle specification. The response is combined test-cycle fuel use in L/100 km. Engine size is measured in litres; body type is a category.

These are associations among specifications. The model does not tell us the fuel use of every driver or show that changing a vehicle's body type would change its consumption.

```{r vehicle-counts}
vehicles |>
  count(body_type, name = "specifications")

nrow(vehicles)
```

## Start with one line

Fit fuel use against engine size and draw the fitted line over the data.

The slope gives the fitted change in L/100 km for a one-litre difference in engine size. It averages across all the body types in this dataset. The intercept is the fitted value at zero litres, well outside the vehicles we observed.

```{r engine-model}
engine_model <- lm(
  fuel_use_l_per_100km ~ engine_l,
  data = vehicles
)

summary(engine_model)

ggplot(vehicles, aes(engine_l, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.45) +
  geom_smooth(method = "lm", se = FALSE, colour = "#0072B2") +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)"
  )
```

## Add Station wagon or Other

First, reduce body type to two categories: Station wagon and Other. We fit one engine-size slope while allowing the groups to have different fitted heights.

The model asks how fuel use differs between these groups at the same engine size. That is a different comparison from putting all vehicles on one line.

```{r binary-model}
binary_model <- lm(
  fuel_use_l_per_100km ~ engine_l + station_wagon,
  data = vehicles
)

binary_plot_data <- vehicles |>
  mutate(fitted = predict(binary_model)) |>
  arrange(station_wagon, engine_l)

tidy(binary_model)

ggplot(binary_plot_data, aes(engine_l, fuel_use_l_per_100km)) +
  geom_point(aes(colour = station_wagon), alpha = 0.45) +
  geom_line(
    aes(y = fitted, colour = station_wagon, group = station_wagon),
    linewidth = 1
  ) +
  scale_colour_manual(
    values = c("Station wagon" = "#0072B2", "Other" = "#D55E00")
  ) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    colour = "Body type"
  )
```

## Two intercepts, one slope

For Station wagons, the fitted mean is

$$
\widehat{\mathrm{fuel\ use}} =
\hat\beta_0 + \hat\beta_1\,\mathrm{engine\ size}.
$$

For Other vehicles, it is

$$
\widehat{\mathrm{fuel\ use}} =
(\hat\beta_0+\hat\beta_2) + \hat\beta_1\,\mathrm{engine\ size}.
$$

Station wagon is the reference category. The coefficient \(\hat\beta_2\) is the fitted Other-minus-Station-wagon difference at the same engine size. Both groups share \(\hat\beta_1\), so the fitted lines are parallel.

## The reference category is coded as zero

Run `model.matrix(binary_model)` and inspect a few rows. R has made a column called `station_wagonOther`: zero for Station wagons and one for Other vehicles.

For a Station wagon, that term drops out. For an Other vehicle, its coefficient is added to the fitted value. The reference category is the first factor level we supplied; it is a coding choice, not a claim that Station wagons are the natural baseline.

```{r model-matrix}
head(model.matrix(binary_model))
```

## Two different comparisons

Compare the raw mean fuel use in each group with the fitted difference at a common engine size.

The raw means combine differences in body type with differences in the kinds of engines found in each group. The factor coefficient holds engine size fixed within the additive model. Neither comparison identifies a causal effect.

```{r raw-adjusted}
raw_group_means <- vehicles |>
  group_by(station_wagon) |>
  summarise(raw_mean = mean(fuel_use_l_per_100km), .groups = "drop")

raw_difference <- with(
  raw_group_means,
  raw_mean[station_wagon == "Other"] -
    raw_mean[station_wagon == "Station wagon"]
)

adjusted_difference <- tidy(binary_model) |>
  filter(term == "station_wagonOther") |>
  pull(estimate)

tibble(
  comparison = c(
    "Raw Other minus Station wagon",
    "Adjusted Other minus Station wagon at the same engine size"
  ),
  difference_l_per_100km = c(raw_difference, adjusted_difference)
)
```

## Keep the question, change the numerical predictor

Now fit fuel use against engine power, first alone and then with Station wagon or Other. Compare the two power slopes.

The simple slope summarises the association across all vehicles. The adjusted slope compares vehicles in the same group. The coefficient for Other compares groups at the same power, in L/100 km.

```{r power-binary}
power_model <- lm(
  fuel_use_l_per_100km ~ power_kw,
  data = vehicles
)

power_binary_model <- lm(
  fuel_use_l_per_100km ~ power_kw + station_wagon,
  data = vehicles
)

tibble(
  model = c("Power only", "Power adjusted for Station wagon/Other"),
  slope_per_10_kw = c(
    10 * coef(power_model)[["power_kw"]],
    10 * coef(power_binary_model)[["power_kw"]]
  )
)

tidy(power_binary_model)
```

## Use all five categories

Our two-group comparison hides differences among Hatchbacks, Sedans, Sports cars and Convertibles. Replace it with the full `body_type` factor.

Station wagon remains the reference. R fits one shared engine-size slope and four body-type differences. Each difference compares that category with Station wagons at the same engine size.

```{r body-model}
body_model <- lm(
  fuel_use_l_per_100km ~ engine_l + body_type,
  data = vehicles
)

tidy(body_model)
```

## What does the Sedan coefficient mean?

The `body_typeSedan` coefficient is the fitted Sedan-minus-Station-wagon difference in L/100 km for vehicles with the same engine size. Its sign gives the direction of that comparison.

It is not the Sedan mean minus the Station wagon mean. It is also not a separate engine-size slope: the additive model uses the same slope for every body type.

```{r body-model-summary}
summary(body_model)$coefficients
```

## Estimates have uncertainty

`Estimate` is the fitted coefficient. `Std. Error` describes how much that estimate would vary across comparable samples under the model. The t statistic is the estimate divided by its standard error; the p-value tests a zero coefficient.

For a body-type row, the null hypothesis is a zero adjusted difference from Station wagons. A small p-value is evidence against that null under the model. It does not measure how large or useful the difference is, and a large p-value does not establish equality. We will work with confidence intervals next week.

## Hold power fixed instead

Fit fuel use against power and all five body types. Compare the Sedan coefficient with the one from the engine-size model.

These models hold different numerical characteristics fixed. If their body-type coefficients differ, that does not make one of them automatically wrong. State which comparison answers the question you actually want to ask.

```{r power-body-model}
power_body_model <- lm(
  fuel_use_l_per_100km ~ power_kw + body_type,
  data = vehicles
)

body_sedan <- tidy(body_model) |>
  filter(term == "body_typeSedan") |>
  mutate(model = "Adjusted for engine size")

power_sedan <- tidy(power_body_model) |>
  filter(term == "body_typeSedan") |>
  mutate(model = "Adjusted for engine power")

bind_rows(body_sedan, power_sedan) |>
  select(model, estimate, std.error, p.value)
```

## Compare body types at 2.0 L

Use `predict(body_model, newdata = ...)` to obtain a fitted mean for each body type at an engine size of 2.0 L.

Every prediction uses the same engine size, so the differences among fitted values come from the body-type coefficients. These are model-based averages, not guarantees for individual vehicles.

```{r body-prediction-engine}
new_body_vehicles <- tibble(
  engine_l = 2,
  body_type = factor(
    levels(vehicles$body_type),
    levels = levels(vehicles$body_type)
  )
)

new_body_vehicles |>
  mutate(
    predicted_fuel_use = predict(body_model, newdata = new_body_vehicles)
  )
```

## Compare body types at 150 kW

Now use the power-and-body-type model to predict each body type at 150 kW. Compare its ranking with the 2.0 L engine-size comparison.

The rankings may differ because one model holds engine size fixed and the other holds power fixed. At 150 kW the Sports-car prediction is below the observed Sports-car power range, so it is an extrapolation. Do not treat its rank as a like-for-like comparison supported by these data.

```{r body-prediction-power}
new_body_power <- tibble(
  power_kw = 150,
  body_type = factor(
    levels(vehicles$body_type),
    levels = levels(vehicles$body_type)
  )
)

new_body_power |>
  mutate(
    predicted_fuel_use = predict(
      power_body_model,
      newdata = new_body_power
    )
  )
```

## Does adding body type improve the model?

Compare the simple engine-size model, the two-group model and the five-body-type model. All three have the same response in the same units and use the same vehicle rows.

Ordinary R² cannot fall when we add terms to a model fitted on the same observations. Adjusted R² accounts for the number of fitted coefficients, so it can fall if extra terms contribute too little. Neither measure establishes causation or performance on new vehicles.

```{r model-fit}
model_fit <- bind_rows(
  glance(engine_model) |> mutate(model = "Engine size"),
  glance(binary_model) |> mutate(model = "Engine size + Station wagon/Other"),
  glance(body_model) |> mutate(model = "Engine size + five body types")
) |>
  select(model, nobs, r.squared, adj.r.squared)

stopifnot(n_distinct(model_fit$nobs) == 1)

model_fit
```

## What have we assumed?

The additive model lets body types have different intercepts but gives them a common numerical slope. The lines are parallel because we specified that structure.

A model with different slopes would ask another question. We are not fitting that model in 297.101. For now, interpret each body-type difference at a shared value of engine size or power.

## Try the same reasoning with mass

In the workshop you will start with gross vehicle mass, add the two-group factor, then use all five body types. Watch what happens to the mass slope when you compare vehicles within body types.

Next week we ask how uncertain the coefficients and predictions are, and whether the residual patterns support the model we have fitted.
