---
title: "297.101 Week 11: Multiple Regression and Factors"
subtitle: "Adjusted comparisons and predictions"
author: Dr. Nick Knowlton
format:
  revealjs:
    code-copy: hover
    embed-resources: true
    css: "knowlton.css"
    theme: sky
    slide-number: true
    incremental: true
    smaller: true
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-svg.js
    highlight-style: breezedark
    highlight-lines: true
    progress: true
    logo: graphics/660491-ls_black_whitoutunz.png
execute:
  output: true
  eval: true
  code-fold: true
editor:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)
library(knitr)
library(ggplot2)
theme_set(theme_bw(base_size = 15))

# Quarto may execute from the repository root or the lecture directory.
donkey_file <- if (file.exists("data/donkey.csv")) {
  "data/donkey.csv"
} else {
  "../data/donkey.csv"
}

donkey <- readr::read_csv(donkey_file, show_col_types = FALSE) |>
  mutate(
    Sex = factor(Sex, levels = c("Female", "Male")),
    AgeGroup = factor(
      if_else(Age < 7, "Under 7", "7 and over"),
      levels = c("Under 7", "7 and over")
    )
  )
```

## Learning outcomes

By the end of this lecture, you should be able to:

- fit a model with a numerical predictor and categorical predictors;
- interpret an adjusted slope and a difference from a reference category;
- explain why an adjusted comparison can differ from a raw comparison;
- make predictions at common values of the other predictors; and
- compare ordinary and adjusted R² across models fitted to the same response.

# Back to the Moroccan donkeys {.center}

## One predictor was a start

Last week we fitted body weight against heart girth:

```r
Bodywt ~ Heartgirth
```

Heart girth explains a large amount of the variation in body weight, but
donkeys with the same heart girth can still have different body weights.

Can some of that remaining variation be explained by age or sex? We can add
those variables while keeping heart girth in the model.

## The data

Each row is one Moroccan donkey. We will keep body weight as the response and
heart girth as our numerical predictor.

This week we will also use two categorical predictors:

- **Age group:** Under 7 versus 7 and over
- **Sex:** Female versus Male

The age split is just a convenient teaching category. We are not claiming that
seven years is a biologically important threshold.

```{r donkey-counts}
donkey |>
  count(AgeGroup, Sex)
```

## Start with the model from Week 10

```{r heart-model}
heart_model <- lm(
  Bodywt ~ Heartgirth,
  data = donkey
)

summary(heart_model)

ggplot(donkey, aes(Heartgirth, Bodywt)) +
  geom_point(alpha = 0.45) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Heart girth (cm)",
    y = "Body weight (kg)"
  )
```

The slope gives the fitted change in mean body weight for a one-centimetre
difference in heart girth.

## Add age group

Now allow younger and older donkeys to have different fitted mean body weights
at the same heart girth.

```{r age-model}
age_model <- lm(
  Bodywt ~ Heartgirth + AgeGroup,
  data = donkey
)

age_plot_data <- donkey |>
  mutate(fitted = predict(age_model)) |>
  arrange(AgeGroup, Heartgirth)

tidy(age_model)

ggplot(age_plot_data, aes(Heartgirth, Bodywt)) +
  geom_point(aes(colour = AgeGroup), alpha = 0.45) +
  geom_line(
    aes(y = fitted, colour = AgeGroup, group = AgeGroup),
    linewidth = 1
  ) +
  labs(
    x = "Heart girth (cm)",
    y = "Body weight (kg)",
    colour = "Age group"
  )
```

## Two intercepts, one slope

For donkeys under 7, the fitted mean is

$$
\widehat{\mathrm{Bodywt}} =
\hat\beta_0 + \hat\beta_1\,\mathrm{Heartgirth}.
$$

For donkeys aged 7 and over, it is

$$
\widehat{\mathrm{Bodywt}} =
(\hat\beta_0+\hat\beta_2) +
\hat\beta_1\,\mathrm{Heartgirth}.
$$

**Under 7** is the reference category. The coefficient
\(\hat\beta_2\) is the fitted difference between the age groups at the
same heart girth.

Both groups share \(\hat\beta_1\), so the fitted lines are parallel.

## The reference category is coded as zero

Run `model.matrix(age_model)` and inspect a few rows.

R has made a column called `AgeGroup7 and over`:

- 0 for donkeys under 7
- 1 for donkeys aged 7 and over

For the reference group, that term drops out. For the other group, its
coefficient is added to the fitted value.

The reference category is a coding choice. It does not change the fitted
values.

```{r model-matrix}
head(model.matrix(age_model))
```

## Raw and adjusted comparisons are different

First compare the raw mean body weight in the two age groups.

Then compare the coefficient from the model that holds heart girth fixed.

```{r raw-adjusted-age}
raw_age_means <- donkey |>
  group_by(AgeGroup) |>
  summarise(raw_mean = mean(Bodywt), .groups = "drop")

raw_age_difference <- with(
  raw_age_means,
  raw_mean[AgeGroup == "7 and over"] -
    raw_mean[AgeGroup == "Under 7"]
)

adjusted_age_difference <- tidy(age_model) |>
  filter(term == "AgeGroup7 and over") |>
  pull(estimate)

tibble(
  comparison = c(
    "Raw: 7 and over minus Under 7",
    "Adjusted: 7 and over minus Under 7 at the same heart girth"
  ),
  difference_kg = c(raw_age_difference, adjusted_age_difference)
)
```

The raw means also contain differences in heart girth between the groups. The
factor coefficient compares the groups **at the same heart girth** within this
additive model.

## Add another factor: Sex

Now add Sex to the age-adjusted model.

```{r full-model}
full_model <- lm(
  Bodywt ~ Heartgirth + AgeGroup + Sex,
  data = donkey
)

tidy(full_model)
```

With Female as the reference category, `SexMale` is the fitted Male-minus-
Female difference for donkeys with the same heart girth and in the same age
group.

Nothing about adding Sex changes the meaning of the heart-girth slope: it is
still the fitted change in mean body weight for a one-centimetre difference in
heart girth, holding the other predictors fixed.

## What does "holding the others fixed" mean?

In the full model:

- the **Heartgirth** coefficient compares donkeys differing by 1 cm in heart
  girth but with the same age group and sex;
- the **AgeGroup** coefficient compares age groups at the same heart girth and
  sex;
- the **Sex** coefficient compares males and females at the same heart girth
  and age group.

That is what an adjusted coefficient is doing. Each coefficient answers its
comparison after accounting for the other terms in the model.

## One slope, several shifted lines

The full model is still additive. There is one common heart-girth slope.

```{r full-model-plot}
full_plot_data <- donkey |>
  mutate(fitted = predict(full_model)) |>
  arrange(AgeGroup, Sex, Heartgirth)

ggplot(full_plot_data, aes(Heartgirth, Bodywt)) +
  geom_point(aes(colour = Sex), alpha = 0.35) +
  geom_line(
    aes(
      y = fitted,
      colour = Sex,
      linetype = AgeGroup,
      group = paste(Sex, AgeGroup)
    ),
    linewidth = 1
  ) +
  labs(
    x = "Heart girth (cm)",
    y = "Body weight (kg)",
    colour = "Sex",
    linetype = "Age group"
  )
```

Four groups give four fitted lines, but all four lines have the same slope.
The factor terms only shift the fitted mean up or down.

## Estimates have uncertainty

```{r full-model-summary}
summary(full_model)$coefficients
```

`Estimate` is the fitted coefficient. `Std. Error` describes the sampling
uncertainty in that estimate under the model.

The t statistic is the estimate divided by its standard error, and the p-value
tests a zero coefficient.

For a factor row, the null hypothesis is a zero adjusted difference from the
reference category. A large p-value is not evidence that the groups are
identical. It means these data do not provide strong evidence for an adjusted
difference.

We will work more carefully with uncertainty next week.

## Predict comparable donkeys

Suppose we want fitted mean body weights for all four combinations of age group
and sex at a heart girth of 115 cm.

```{r common-predictions}
new_donkeys <- tidyr::expand_grid(
  Heartgirth = 115,
  AgeGroup = levels(donkey$AgeGroup),
  Sex = levels(donkey$Sex)
)

new_donkeys |>
  mutate(
    predicted_body_weight = predict(
      full_model,
      newdata = new_donkeys
    )
  )
```

Every prediction uses the same heart girth. The remaining differences come
from the fitted factor coefficients.

## Does every extra term improve the model?

Compare four models fitted to exactly the same response and the same donkeys.

```{r model-fit}
sex_model <- lm(
  Bodywt ~ Heartgirth + Sex,
  data = donkey
)

model_fit <- bind_rows(
  glance(heart_model) |>
    mutate(model = "Heart girth"),
  glance(age_model) |>
    mutate(model = "Heart girth + age group"),
  glance(sex_model) |>
    mutate(model = "Heart girth + sex"),
  glance(full_model) |>
    mutate(model = "Heart girth + age group + sex")
) |>
  select(model, nobs, r.squared, adj.r.squared)

stopifnot(n_distinct(model_fit$nobs) == 1)

model_fit
```

Ordinary R² cannot fall when we add terms to a model fitted to the same
observations.

Adjusted R² asks whether the improvement is large enough to justify estimating
another coefficient. Here, age group earns its place more clearly. Adding Sex
after age group increases ordinary R² a little, but adjusted R² is essentially
unchanged and is slightly lower.

## Why adjusted R² can go down

Adding a term gives the model another way to fit the observed data. So ordinary
R² can only stay the same or increase.

Adjusted R² applies a penalty for estimating more coefficients. If the new term
does not reduce the residual variation enough, the penalty is larger than the
gain and adjusted R² falls.

That does **not** prove the term has no scientific importance. It tells us that,
in this fitted model and sample, the extra term has added little explanatory
information.

## What have we assumed?

The additive model allows the groups to have different intercepts but gives all
groups the same heart-girth slope.

The lines are parallel because that is the model we specified.

A model with different slopes would contain an interaction. We are **not**
fitting interactions in 297.101. For this course, interpret the factor
differences at a shared value of the numerical predictor.

## The same structure in the workshop

The workshop uses the vehicle data you first saw in Week 10.

The dataset changes, but the model logic is the same:

1. start with a numerical predictor;
2. add a categorical predictor;
3. interpret factor coefficients relative to a reference category;
4. compare fitted values while holding the numerical predictor fixed; and
5. ask whether the extra terms add enough information to justify the more
   complicated model.

Next week we add uncertainty, residual diagnostics and model assumptions.
