---
title: "297.101 Week 12: Inference and Assumptions"
subtitle: "Uncertainty, intervals and model diagnostics"
author: Dr. Nick Knowlton
format:
  revealjs:
    code-copy: hover
    embed-resources: true
    css:
      - "knowlton.css"
      - "regression-lab.css"
      - "line-assumptions-lab.css"
    theme: sky
    slide-number: true
    incremental: true
    smaller: true
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-svg.js
    highlight-style: breezedark
    highlight-lines: true
    progress: true
    logo: graphics/660491-ls_black_whitoutunz.png
execute:
  output: true
  eval: true
  code-fold: true
editor:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)
library(ggplot2)
theme_set(theme_bw(base_size = 15))

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  out.width = "100%"
)

# Quarto may execute from the repository root or the lecture directory.
donkey_file <- if (file.exists("data/donkey.csv")) {
  "data/donkey.csv"
} else {
  "../data/donkey.csv"
}

donkey <- readr::read_csv(donkey_file, show_col_types = FALSE) |>
  mutate(
    Sex = factor(Sex, levels = c("Female", "Male")),
    AgeBand = factor(
      case_when(
        Age < 7 ~ "Under 7",
        Age < 10 ~ "7–9",
        TRUE ~ "10 and over"
      ),
      levels = c("7–9", "Under 7", "10 and over")
    )
  )

mod <- lm(Bodywt ~ Heartgirth, data = donkey)
```

```{ojs}
//| echo: false
//| output: false
import { predictionIntervalLab } from "./regression-lab.js"
import {
  linearityLab,
  independenceLab,
  normalityLab,
  equalVarianceLab,
} from "./line-assumptions-lab.js"
```

# Learning outcomes

By the end of this lecture, you should be able to:

- explain why fitted coefficients vary across repeated comparable samples;
- read uncertainty and hypothesis tests for a regression coefficient;
- distinguish a test of one factor coefficient from a joint test of the whole factor;
- distinguish an interval for a mean response from an interval for one new outcome;
- explain the assumptions behind linear-model inference;
- use residual plots to identify possible departures from those assumptions; and
- investigate influential observations and compare a reasonable alternative model.

# How much can we trust the fitted model? {.center}

Last week we used heart girth, age group and sex to make adjusted comparisons and fitted mean predictions for donkeys.
The coefficients are estimates from a *sample*, and the fitted means are predictions from a *model*.

. . .

Now we need to quantify uncertainty and check whether the model's structure is credible. We will use the donkey model for the worked examples. In the workshop, you will transfer the same reasoning to county-level cancer data.

## Why do estimates vary?

Last week we distinguished a population coefficient from the estimate calculated
from our sample. The population slope, $\beta$, is fixed but unknown. The fitted
slope, $\hat\beta$, depends on which donkeys we happened to observe.

If we repeated the same study with another comparable sample of donkeys, the
observed body weights would differ slightly and the fitted line would move. The
collection of $\hat\beta$ values we could obtain across repeated samples is
the **sampling distribution** of the slope estimator.

That sampling distribution is what coefficient uncertainty refers to. We are
not treating the population slope itself as random.

## From sampling distribution to standard error

Under the linear model, the least-squares estimator is centred on the population
coefficient:

$$
E(\hat\beta)=\beta.
$$

So across repeated comparable samples, least squares does not systematically
estimate the slope too high or too low.

The **standard error** estimates the spread of that sampling distribution. With
more information, usually from a larger and more informative sample, that spread
gets narrower.

This is the link we need for inference:

$$
\text{estimate} \rightarrow \text{standard error} \rightarrow
\text{t statistic} \rightarrow \text{p-value or confidence interval}.
$$

## A coefficient is an estimate

Fit body weight against heart girth and inspect the coefficient table. The slope is the estimated change in mean body weight, in kilograms, for a one-centimetre difference in heart girth.

The standard error describes sampling variation in that estimated slope under the model. It is not the spread of individual donkey weights around the fitted line. That spread is summarised separately by the residual standard error.

```{r coefficient-output}
summary(mod)$coefficients
sigma(mod)
```

## What does the slope p-value test?

For heart girth, the coefficient test starts from $H_0:\beta_{\mathrm{girth}}=0$. The t statistic compares the estimated slope with its standard error. The p-value describes how unusual a statistic at least as extreme would be if that null hypothesis and the model assumptions held.

A small p-value is evidence against a zero slope. It does not tell us the probability that the null is true, how accurate individual predictions are, or whether heart girth causes body weight.

## A range for the population slope

Use a 95% confidence interval for the heart-girth coefficient. In repeated comparable samples, the interval procedure would contain the population coefficient about 95% of the time.

For a two-sided 5% test of a zero coefficient, an interval that excludes zero agrees with rejecting that null. Interpret the endpoints in kilograms per centimetre. We do not say that there is a 95% probability that this particular fixed coefficient lies in the calculated interval.

```{r coefficient-interval}
#| echo: true
#| code-fold: false
confint(mod, parm = "Heartgirth")
```

## What if a factor has three levels?

Last week we used Under 7 versus 7 and over. Now split the older group into
7–9 and 10 and over. These are **teaching bands**, keeping last week's cutoff
at 7.

```{r age-band-counts, echo=FALSE}
donkey |>
  count(AgeBand) |>
  arrange(match(AgeBand, c("Under 7", "7–9", "10 and over"))) |>
  knitr::kable(col.names = c("Age band", "Donkeys"))
```

```{r age-band-model}
#| echo: true
#| code-fold: false
factor_model <- lm(Bodywt ~ Heartgirth + Sex + AgeBand, data = donkey)
```

With **7–9 as the reference**, three levels need two age-band coefficients.
Comparisons hold heart girth and sex fixed; the model keeps one common slope.

## What do these p-values test?

Each age-band row tests a zero population difference from **7–9**, at the
same heart girth and sex.

```{r age-band-coefficients, echo=FALSE}
age_band_terms <- tidy(factor_model) |>
  filter(startsWith(term, "AgeBand"))

age_band_terms |>
  transmute(
    Comparison = c("Under 7 versus 7–9", "10 and over versus 7–9"),
    `Difference (kg)` = sprintf("%.2f", estimate),
    `p-value` = format.pval(p.value, digits = 3, eps = 0.00001)
  ) |>
  knitr::kable(align = c("l", "r", "r"))
```

**The second comparison is inconclusive. Does that mean age band contributes
no information?**

. . .

No. Its large p-value does not establish equality or answer the question
about the whole factor. The Under 7 comparison has a small p-value.

## Test the factor as a whole

The joint null hypothesis is that **all three population mean weights are
equal at the same heart girth and sex**, under this additive model:

$$
H_0:\quad \mu_{\text{Under 7}}=\mu_{\text{7–9}}=\mu_{\text{10 and over}}.
$$

With **7–9 as the reference**, this is equivalent to
$\beta_{\mathrm{Under\ 7}}=\beta_{\mathrm{10\ and\ over}}=0$:
both age-band coefficients must be zero **together**. The reference group's
coefficient is fixed at zero. The alternative is that at least one age-band
mean differs.

. . .

An **ANOVA F test** asks whether adding both coefficients improves the fit
enough relative to the remaining residual variability. This is a test of
age band after accounting for heart girth and sex.

## Ask whether adding age band improves the fit

Compare two models fitted to the **same donkeys**, differing only by the
whole age-band factor:

```{r age-band-anova}
#| echo: true
#| code-fold: false
reduced <- lm(Bodywt ~ Heartgirth + Sex, data = donkey)
anova(reduced, factor_model)
```

```{r age-band-test-result, include=FALSE}
factor_test <- anova(reduced, factor_model)
```

**F(2, 380) = 5.473, p = 0.00454.** There is evidence
of an age-band difference after adjustment. This does not mean every pair differs.

Use the explicit comparison: `anova(model)` alone gives sequential tests,
adding terms in formula order, which can answer a different adjustment question.

## Changing the reference does not change the factor test

Refit with **Under 7 as the reference**:

```{r age-band-reference}
#| echo: true
#| code-fold: false
donkey_under7 <- donkey |>
  mutate(AgeBand = relevel(AgeBand, ref = "Under 7"))
factor_model_under7 <- update(factor_model, data = donkey_under7)
```

```{r age-band-reference-output, echo=FALSE}
tidy(factor_model_under7) |>
  filter(startsWith(term, "AgeBand")) |>
  transmute(
    Comparison = c("7–9 versus Under 7", "10 and over versus Under 7"),
    `Difference (kg)` = sprintf("%.2f", estimate),
    `p-value` = format.pval(p.value, digits = 3, eps = 0.00001)
  ) |>
  knitr::kable(align = c("l", "r", "r"))

factor_test_under7 <- anova(reduced, factor_model_under7)
stopifnot(
  identical(rownames(model.frame(reduced)), rownames(model.frame(factor_model))),
  isTRUE(all.equal(fitted(factor_model), fitted(factor_model_under7))),
  isTRUE(all.equal(factor_test$F[2], factor_test_under7$F[2])),
  isTRUE(all.equal(factor_test$`Pr(>F)`[2], factor_test_under7$`Pr(>F)`[2]))
)
```

Both age-row p-values are now small. The fitted values and joint F test
are unchanged: **F(2, 380) = 5.473, p = 0.00454**.

**A coefficient test asks about one contrast; the factor test asks about
the contrasts together.**

## For a binary factor, the t and F tests agree

`Sex` has two levels, so it adds just one coefficient. Both tests ask whether
Male and Female population mean weights differ **at the same heart girth and
age band**: $H_0:\beta_{\mathrm{Male}}=0$.

```{r sex-binary-tests}
#| echo: true
#| code-fold: false
#| results: hide
sex_reduced <- update(factor_model, . ~ . - Sex)
sex_coefficient <- summary(factor_model)$coefficients["SexMale", ]
sex_test <- anova(sex_reduced, factor_model)
```

| Test | Statistic | p-value |
|:-----|:----------|--------:|
| SexMale coefficient: two-sided t test | t = 0.9895 | 0.32306 |
| Remove Sex: ANOVA F test | F(1, 380) = 0.9790 | 0.32306 |

With one coefficient being tested, **$F=t^2$**, so the F test and the
two-sided t test give **the same p-value**. With three age levels, the F test
instead tests the two age-band coefficients together.

```{r sex-binary-test-checks, include=FALSE}
stopifnot(
  identical(rownames(model.frame(sex_reduced)), rownames(model.frame(factor_model))),
  isTRUE(all.equal(unname(sex_coefficient["t value"])^2, sex_test$F[2])),
  isTRUE(all.equal(unname(sex_coefficient["Pr(>|t|)"]), sex_test$`Pr(>F)`[2]))
)
```

## Population vs Individual

We have tested population slopes and differences between groups. Now return
to the simple heart-girth model and ask how uncertain its predictions are.

At a heart girth of 120 cm, what is the estimated average body weight? What range might contain the weight of one new donkey at 120 cm?

Both questions use the same fitted value. The second has an additional source of uncertainty because individual weights vary around their mean.

## Ask R for the interval you mean

Use `interval = "confidence"` for uncertainty in the mean weight at 120 cm,
and `interval = "prediction"` for one new donkey's weight at 120 cm.

The prediction interval is wider because it incorporates residual variation as well as uncertainty in the fitted mean. Both intervals depend on the model and its assumptions; neither is a promise about a particular donkey.

```{r prediction-intervals}
#| echo: true
#| code-fold: false
new_donkey <- data.frame(Heartgirth = 120)

predict(mod, newdata = new_donkey, interval = "confidence")
predict(mod, newdata = new_donkey, interval = "prediction")
```

## Which interval answers the question?

Choose the target first: the average outcome at a specified predictor value, or one new individual outcome. Predict which interval should be wider, then reveal both.

Move the predictor value toward the edge of the observed range. What happens to uncertainty in the fitted mean?

```{ojs}
//| echo: false
predictionIntervalLab({ Inputs, Plot })
```

# What least squares assumes {.center}

## Why look at residuals?

For each donkey, the residual is its observed weight minus its fitted weight:

$$
e_i=y_i-\hat y_i,\qquad
\mathrm{RSS}=\sum_{i=1}^{n}e_i^2,\qquad
\mathrm{MSE}=\frac{\mathrm{RSS}}{n}.
$$

Least squares chooses the line with the smallest residual sum of squares (RSS).
For this sample, $n$ is fixed, so the same line also minimises the mean squared
residual (MSE). It will produce a line even if a straight-line mean fits poorly.

The residuals let us look for systematic patterns that the fitted model left
behind. We need separate questions about the shape of the mean, dependence
between observations, the shape of residuals, and changes in spread.

```{r least-squares}
rss <- sum(residuals(mod)^2)
c(RSS = rss, MSE = rss / nobs(mod))
```

## Four checks: L–I–N–E

**Linearity:** Is the conditional mean adequately described by the specified straight-line terms?

**Independence:** Could knowing one observation's error help predict another's? Consider how the observations were collected.

**Normality:** Are residuals sufficiently close to a normal shape for the interval or test being used, especially in the tails?

**Equal variance:** Is the residual spread roughly constant across fitted values?

These checks probe the assumptions; a plot cannot prove that they hold.

## Which assumptions support which claims?

A straight-line mean, appropriate independence and constant error variance
support the familiar least-squares standard errors and their usual
interpretation. Normal errors provide the exact small-sample t and F
distributions used for conventional intervals and tests.

These assumptions also support the joint age-band F test we have just used.

Least squares finds the line with the smallest squared residuals without
assuming normal errors. Normality matters for the exact small-sample inference
we have just used; it is not needed to calculate that best-fitting line.

## L: what pattern has the line missed?

First look at the scatterplot and fitted line. Predict what the residual-versus-fitted plot will show, then reveal it. Increase curvature while keeping the same underlying sample.

Does a high R² rule out a systematic curve in the residuals?

```{ojs}
//| echo: false
linearityLab({ Inputs, Plot })
```

## I: does order contain information?

The same points can look unremarkable in a scatterplot while their residuals form runs when we recover the time order. Change the displayed order without changing any $x$, $y$, fitted value or residual.

Shuffling can hide temporal dependence, but it does not make observations independent. The alternating order is a display arrangement, not a plausible chronology.

A residual plot against fitted values cannot establish independence. We also need the sampling design and, where relevant, the meaningful order of observations.

```{ojs}
//| echo: false
independenceLab({ Inputs, Plot })
```

## N: what does a Q–Q plot compare?

Each point represents one ordered residual against a theoretical normal quantile. Small departures, especially in a small sample, can occur by chance. Look for a sustained pattern, strong asymmetry or unusually heavy tails.

Generate another sample before judging a subtle deviation. The Q–Q plot is evidence about residual shape, not a yes-or-no proof of normality.

```{ojs}
//| echo: false
normalityLab({ Inputs, Plot })
```

## E: does the spread change?

Keep the mean relationship fixed and change how widely observations vary around it. A funnel in residuals versus fitted values suggests non-constant variance.

If the spread grows, uncertainty for an individual outcome also changes across the predictor range. A straight fitted mean does not repair that problem.

```{ojs}
//| echo: false
equalVarianceLab({ Inputs, Plot })
```

## Donkey data: L and I

**Linearity:** Does the residual-versus-fitted plot show a curve that the line
has missed? It can also give an initial indication of changing spread.

**Independence:** This plot cannot tell us whether observations are independent.
We would need to know how the donkeys were sampled and whether they have a
meaningful order or grouping.

```{r donkey-linearity, fig.height=5.5}
plot(mod, which = 1, pch = 19, col = "#00000080")
```

## Donkey data: N

**Normality:** Do the residual quantiles track the reference line, or do the
tails depart systematically? A Q–Q plot can raise a concern, not confirm
normality.

```{r donkey-normality, fig.height=5.5}
plot(mod, which = 2, pch = 19, col = "#00000080")
```

## Donkey data: E

**Equal variance:** Does the residual spread change across fitted weights? A
rising or falling trend in this scale-location plot suggests that it does.

```{r donkey-equal-variance, fig.height=5.5}
plot(mod, which = 3, pch = 19, col = "#00000080")
```

## Donkey data: influence

This plot compares residual size with leverage. The Cook's-distance contours
help flag observations whose removal could change the fitted line. Which
labelled observations would you check in the original data?

```{r donkey-influence-diagnostic, fig.height=5.5}
plot(mod, which = 5, pch = 19, col = "#00000080")
```

## Investigate influence before changing the data

A point can have an unusual predictor value, a large residual, or both. Cook's
distance summarises how much the fitted model would change if that observation
were omitted.

Check the source record and measurement process before deciding what to do. A
correct but unusual donkey is still part of the data; a data-entry error calls
for correction. Report any exclusion and examine whether it changes the
substantive result.

```{r influence}
influence_data <- tibble(
  observation = as.integer(rownames(model.frame(mod))),
  Bodywt = model.response(model.frame(mod)),
  Heartgirth = model.frame(mod)$Heartgirth,
  cook_distance = cooks.distance(mod),
  leverage = hatvalues(mod),
  studentised_residual = rstudent(mod)
)

top_five <- influence_data |>
  slice_max(cook_distance, n = 5, with_ties = FALSE) |>
  arrange(desc(cook_distance))

top_five |>
  select(observation, Heartgirth, Bodywt, cook_distance) |>
  knitr::kable(
    digits = 2,
    col.names = c("Observation", "Heart girth (cm)", "Body weight (kg)", "Cook's D")
  )
```

## Where are those five donkeys?

The orange points are the same five observations listed above. Their positions
show whether their influence comes from unusual heart girth, body weight, or
both. Check the records before considering any exclusion.

```{r influence-on-data, fig.height=5.5}
ggplot(donkey, aes(Heartgirth, Bodywt)) +
  geom_point(colour = "#00000070") +
  geom_abline(
    intercept = coef(mod)[1], slope = coef(mod)[2],
    colour = "#0072B2", linewidth = 0.8
  ) +
  geom_point(data = top_five, colour = "#D55E00", size = 3) +
  geom_text(
    data = top_five, aes(label = observation),
    colour = "#A34100", vjust = -0.8, fontface = "bold"
  ) +
  labs(x = "Heart girth (cm)", y = "Body weight (kg)")
```

## Change the predictor scale

We have fitted a straight line using heart girth in centimetres. A second model
uses the natural log of heart girth as its predictor:

$$
\widehat{W}_{\mathrm{linear}}=a+bH,\qquad
\widehat{W}_{\log H}=c+d\log(H).
$$

The second fitted mean can curve when plotted against heart girth in
centimetres. `exp()` reverses `log()`: $\exp(\log H)=H$. We have **not** logged
body weight, so both models still predict weight in kilograms.

## What does the log-heart-girth coefficient mean?

The original slope $b$ describes a kilogram change for one more centimetre.
In the log-heart-girth model, a fixed centimetre increase has a different
fitted effect at different starting girths.

For a **1% increase** in heart girth, from $H$ to $1.01H$:

$$
\Delta\widehat W=d\{\log(1.01H)-\log(H)\}
=d\log(1.01)\text{ kg}.
$$

For a **1% decrease**, the fitted change is $d\log(0.99)$ kg. These are
percentage changes in **heart girth**; the changes in fitted weight are in
kilograms.

## Could another scale help?

If the residuals show curvature or changing spread, a transformation may be
worth checking. Fit the log-heart-girth model, then compare its fit and
diagnostics with the original.

Both models predict body weight in kilograms, so their in-sample R² and
residual standard deviations refer to the same response scale. A small
improvement in one number alone is not enough to choose a model. Check the
residuals, the observed predictor range and whether the new coefficient remains
useful to interpret.

```{r transformed-model}
mod_log_x <- lm(Bodywt ~ log(Heartgirth), data = donkey)

bind_rows(
  glance(mod) |> mutate(model = "Heart girth"),
  glance(mod_log_x) |> mutate(model = "Log heart girth")
) |>
  select(model, r.squared, sigma) |>
  knitr::kable(
    digits = 3,
    col.names = c("Model", "R²", "Residual SD (kg)")
  )
```

## Compare fitted shapes

Both lines use the same observed donkeys and predict body weight in kilograms.
Does the logged predictor change the shape enough to address the pattern we saw
in the original residuals?

```{r transformed-curves, fig.height=5.5}
girth_grid <- tibble(
  Heartgirth = seq(min(donkey$Heartgirth), max(donkey$Heartgirth), length.out = 200)
)

fitted_curves <- bind_rows(
  girth_grid |>
    mutate(Bodywt = predict(mod, newdata = girth_grid), model = "Heart girth"),
  girth_grid |>
    mutate(Bodywt = predict(mod_log_x, newdata = girth_grid),
           model = "Log heart girth")
)

ggplot(donkey, aes(Heartgirth, Bodywt)) +
  geom_point(colour = "#00000050") +
  geom_line(
    data = fitted_curves, aes(colour = model), linewidth = 1
  ) +
  scale_colour_manual(values = c("Heart girth" = "#0072B2",
                                 "Log heart girth" = "#D55E00")) +
  labs(x = "Heart girth (cm)", y = "Body weight (kg)", colour = "Model")
```

## Check the logged model's residuals

The logged model has lower $R^2$, a larger residual standard deviation and a
similar curved residual pattern. Of these two models, I would keep the original
heart-girth model. The remaining curvature still deserves attention.

```{r transformed-diagnostics, fig.height=5.5}
plot(mod_log_x, which = 1, pch = 19, col = "#0072B280")
```

## Before you trust the output

**R will happily fit the model and give you p-values. That does not mean the model is doing a good job.**

- **Coefficients are estimates.** Standard errors, p-values and confidence intervals tell us how much uncertainty sits around them.
- **Match the test to the question.** A factor coefficient tests one contrast with the reference; a joint F test tests the whole factor.
- **Be clear what you are predicting.** A mean response and one new donkey are not the same target. The prediction interval is wider because individual donkeys vary around the fitted mean.
- **Check what the model has left behind.** Use **L-I-N-E**: look for curvature, think about independence, check the residual distribution and look for changing spread.
- **A diagnostic plot raises a question, it does not settle it.** Independence comes from how the data were collected. An unusual point needs investigating before you delete it.
- **Changing the model should fix an actual problem.** Logging heart girth did not remove the curvature here, so there is no reason to prefer it.

**Do not stop at `summary(model)`. Look at the uncertainty, look at the residuals, and decide whether you believe the model.**
