---
title: "297.101 Week 12: Inference and Assumptions"
subtitle: "Uncertainty, intervals and model diagnostics"
author: Dr. Nick Knowlton
format:
  revealjs:
    code-copy: hover
    embed-resources: true
    css:
      - "knowlton.css"
      - "regression-lab.css"
      - "line-assumptions-lab.css"
    theme: sky
    slide-number: true
    incremental: true
    smaller: true
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-svg.js
    highlight-style: breezedark
    highlight-lines: true
    progress: true
    logo: graphics/660491-ls_black_whitoutunz.png
execute:
  output: true
  eval: true
  code-fold: true
editor:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)
library(ggplot2)
theme_set(theme_bw(base_size = 15))

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  out.width = "100%"
)

donkey <- read.csv(
  "https://www.massey.ac.nz/~jcmarsha/227215/data/donkey.csv"
)

mod <- lm(Bodywt ~ Heartgirth, data = donkey)
```

```{ojs}
//| echo: false
//| output: false
import { predictionIntervalLab } from "./regression-lab.js"
import {
  linearityLab,
  independenceLab,
  normalityLab,
  equalVarianceLab,
} from "./line-assumptions-lab.js"
```

## Learning outcomes

By the end of this lecture, you should be able to:

- read uncertainty and hypothesis tests for a regression coefficient;
- distinguish an interval for a mean response from an interval for one new outcome;
- explain the assumptions behind linear-model inference;
- use residual plots to identify possible departures from those assumptions; and
- investigate influential observations and compare a reasonable alternative model.

# How much can we trust the fitted model? {.center}

Last week we used vehicle models to make adjusted comparisons. A coefficient is still an estimate from a sample, and a fitted value is still a prediction from a model.

Now we need to quantify uncertainty and check whether the model's structure is credible. We will use the donkey model for the worked examples. In the workshop, you will transfer the same reasoning to county-level cancer data.

## A coefficient is an estimate

Fit body weight against heart girth and inspect the coefficient table. The slope is the estimated change in mean body weight, in kilograms, for a one-centimetre difference in heart girth.

The standard error describes sampling variation in that estimated slope under the model. It is not the spread of individual donkey weights around the fitted line. That spread is summarised separately by the residual standard error.

```{r coefficient-output}
summary(mod)$coefficients
sigma(mod)
```

## What does the slope p-value test?

For heart girth, the coefficient test starts from \(H_0:\beta_{\mathrm{girth}}=0\). The t statistic compares the estimated slope with its standard error. The p-value describes how unusual a statistic at least as extreme would be if that null hypothesis and the model assumptions held.

A small p-value is evidence against a zero slope. It does not tell us the probability that the null is true, how accurate individual predictions are, or whether heart girth causes body weight.

## A range for the population slope

Use a 95% confidence interval for the heart-girth coefficient. In repeated comparable samples, the interval procedure would contain the population coefficient about 95% of the time.

For a two-sided 5% test of a zero coefficient, an interval that excludes zero agrees with rejecting that null. Interpret the endpoints in kilograms per centimetre. We do not say that there is a 95% probability that this particular fixed coefficient lies in the calculated interval.

```{r coefficient-interval}
confint(mod, parm = "Heartgirth")
```

## Two questions at the same heart girth

At a heart girth of 120 cm, what is the estimated average body weight? What range might contain the weight of one new donkey at 120 cm?

Both questions use the same fitted value. The second has an additional source of uncertainty because individual weights vary around their mean.

## Ask R for the interval you mean

Set `newdata = data.frame(Heartgirth = 120)`. Use `interval = "confidence"` for uncertainty in the mean weight at 120 cm, and `interval = "prediction"` for one new donkey's weight at 120 cm.

The prediction interval is wider because it incorporates residual variation as well as uncertainty in the fitted mean. Both intervals depend on the model and its assumptions; neither is a promise about a particular donkey.

```{r prediction-intervals}
new_donkey <- data.frame(Heartgirth = 120)

predict(mod, newdata = new_donkey, interval = "confidence")
predict(mod, newdata = new_donkey, interval = "prediction")
```

## Which interval answers the question?

Choose the target first: the average outcome at a specified predictor value, or one new individual outcome. Predict which interval should be wider, then reveal both.

Move the predictor value toward the edge of the observed range. What happens to uncertainty in the fitted mean?

```{ojs}
//| echo: false
predictionIntervalLab({ Inputs, Plot })
```

# What least squares assumes {.center}

## Why look at residuals?

Least squares chooses coefficients to minimise the sum of squared residuals, \(e_i=y_i-\hat y_i\). It will produce a fitted line whether or not a straight-line mean describes the data well.

The residuals let us look for systematic patterns that the fitted model left behind. We need separate questions about the shape of the mean, dependence between observations, the shape of residuals, and changes in spread.

```{r least-squares}
sum(residuals(mod)^2)
sigma(mod)
```

## Four checks: L–I–N–E

**Linearity:** Is the conditional mean adequately described by the specified straight-line terms?

**Independence:** Could knowing one observation's error help predict another's? Consider how the observations were collected.

**Normality:** Are residuals sufficiently close to a normal shape for the interval or test being used, especially in the tails?

**Equal variance:** Is the residual spread roughly constant across fitted values?

These checks probe the assumptions; a plot cannot prove that they hold.

## Which assumptions support which claims?

A straight-line mean, appropriate independence and constant error variance support the familiar least-squares standard errors and their usual interpretation. Normal errors provide the exact small-sample t and F distributions used for conventional intervals and tests.

Normality is not one of the Gauss–Markov conditions for least squares to be best linear unbiased. We still inspect the residual shape because this course also uses the usual finite-sample inference and individual prediction intervals.

## L: what pattern has the line missed?

First look at the scatterplot and fitted line. Predict what the residual-versus-fitted plot will show, then reveal it. Increase curvature while keeping the same underlying sample.

Does a high R² rule out a systematic curve in the residuals?

```{ojs}
//| echo: false
linearityLab({ Inputs, Plot })
```

## I: does order contain information?

The same points can look unremarkable in a scatterplot while their residuals form runs when we recover the time order. Change the displayed order without changing any \(x\), \(y\), fitted value or residual.

Shuffling can hide temporal dependence, but it does not make observations independent. The alternating order is a display arrangement, not a plausible chronology.

A residual plot against fitted values cannot establish independence. We also need the sampling design and, where relevant, the meaningful order of observations.

```{ojs}
//| echo: false
independenceLab({ Inputs, Plot })
```

## N: what does a Q–Q plot compare?

Each point represents one ordered residual against a theoretical normal quantile. Small departures, especially in a small sample, can occur by chance. Look for a sustained pattern, strong asymmetry or unusually heavy tails.

Generate another sample before judging a subtle deviation. The Q–Q plot is evidence about residual shape, not a yes-or-no proof of normality.

```{ojs}
//| echo: false
normalityLab({ Inputs, Plot })
```

## E: does the spread change?

Keep the mean relationship fixed and change how widely observations vary around it. A funnel in residuals versus fitted values suggests non-constant variance.

If the spread grows, uncertainty for an individual outcome also changes across the predictor range. A straight fitted mean does not repair that problem.

```{ojs}
//| echo: false
equalVarianceLab({ Inputs, Plot })
```

## Return to the donkey data

Now inspect the standard diagnostic plots for the fitted donkey model. State the visible pattern first, then say which assumption it bears on and why the departure matters.

Residuals versus fitted values can show curvature or changing spread. The Q–Q plot addresses residual shape. A leverage and Cook's-distance plot flags observations whose removal could materially change the fit. None of these plots tells us how the donkeys were sampled.

```{r donkey-diagnostics}
par(mfrow = c(2, 2))
plot(mod, which = c(1, 2, 3, 5), pch = 19, col = "#00000080")
par(mfrow = c(1, 1))
```

## Investigate influence before changing the data

A point can have an unusual predictor value, a large residual, or both. Cook's distance summarises how much the fitted model would change if that observation were omitted.

Check the source record and measurement process before deciding what to do. A correct but unusual donkey is still part of the data; a data-entry error calls for correction. Report any exclusion and examine whether it changes the substantive result.

```{r influence}
influence_data <- tibble(
  observation = seq_len(nrow(donkey)),
  Bodywt = donkey$Bodywt,
  Heartgirth = donkey$Heartgirth,
  cook_distance = cooks.distance(mod),
  leverage = hatvalues(mod),
  studentised_residual = rstudent(mod)
)

influence_data |>
  slice_max(cook_distance, n = 5) |>
  arrange(desc(cook_distance))
```

## Could another scale help?

If the residuals show curvature or changing spread, a transformation may be worth checking. Fit a model with logged heart girth as the predictor, then compare its fitted pattern and diagnostics with the original.

Both models predict body weight in kilograms, so their in-sample R² and residual standard deviations refer to the same response scale. A small improvement in one number alone is not enough to choose a model. Check the residuals, the observed predictor range and whether the new coefficient remains useful to interpret.

```{r transformed-model}
mod_log_x <- lm(Bodywt ~ log(Heartgirth), data = donkey)

bind_rows(
  glance(mod) |> mutate(model = "Heart girth"),
  glance(mod_log_x) |> mutate(model = "Log heart girth")
) |>
  select(model, r.squared, sigma)
```

```{r transformed-diagnostics}
par(mfrow = c(2, 2))
plot(mod, which = c(1, 2), pch = 19, col = "#00000080")
plot(mod_log_x, which = c(1, 2), pch = 19, col = "#0072B280")
par(mfrow = c(1, 1))
```

## Take the workflow to county data

In the workshop you will model county lung-cancer mortality using incidence and poverty. You will interpret adjusted coefficients, calculate intervals for a mean and a single county, assess LINE and investigate influence.

Nearby counties may share exposures and health services. That concern comes from the data collection and geography, even if the residual plots look tidy. Finish by comparing an alternative model on the same mortality scale and explaining what, if anything, it changes.
