---
title: "Workshop Week 12: Inference and Assumptions"
subtitle: "Student workbook — uncertainty, intervals and diagnostics"
author:
  - name: "Nick Knowlton"
  - name: "297.101 Statistical Data Science"
date-meta: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

Weeks 10 and 11 used fitted lines and adjusted coefficients to study
county-level associations between socioeconomic conditions and lung-cancer
mortality. Those fitted models were useful summaries, but they did not tell us
whether a straight line represents the pattern well, whether the uncertainty
calculations are dependable, or whether a few unusual counties strongly affect
the result.

The data come from a broader project that combined county socioeconomic and
cancer information. Noah Rippner's
[Python data-preparation code](https://github.com/nrippner/can_clin_trails_prep_r/blob/master/extract_zips_10.py)
shows how data from different sources were cleaned and joined by place. Those
joins let us examine the same county's socioeconomic conditions and mortality,
although the resulting associations do not establish causes or describe
individual people.

This workshop asks how uncertain the model estimates are and whether the
model's assumptions are reasonable. We use one cancer model throughout the
main demonstrations so that each new inferential idea answers another part of
the same practical problem.

## Today's question

How well can we trust our model's estimates and predictions of county
lung-cancer mortality?

## Learning objectives

By the end of this workshop you should be able to:

- distinguish estimates from their standard errors;
- interpret confidence intervals and hypothesis tests;
- distinguish confidence intervals for an average from prediction intervals for
  an individual county;
- interpret supplied LINE diagnostic plots and use study design to assess a
  model; and
- evaluate one transformed alternative without relying on R-squared alone.

## 1. Fit and understand one primary cancer model

```{r}
library(readr)
library(dplyr)
library(tibble)
library(broom)
library(performance)
library(see)

cancer <- read_csv("regression/cancer_data.csv", show_col_types = FALSE)

cancer_model_data <- cancer |>
  select(
    area_name, Mortality_Rate_num, all_poverty, popestimate2015,
    Incidence_Rate_num
  ) |>
  filter(
    is.finite(Mortality_Rate_num),
    is.finite(all_poverty),
    is.finite(popestimate2015),
    is.finite(Incidence_Rate_num),
    popestimate2015 > 0
  ) |>
  mutate(
    poverty_percent = 100 * all_poverty / popestimate2015,
    incidence_rate = Incidence_Rate_num
  ) |>
  select(area_name, Mortality_Rate_num, poverty_percent, incidence_rate)

glimpse(cancer_model_data)
```

### Demonstration: fit the primary model

```{r}
cancer_model <- lm(
  Mortality_Rate_num ~ poverty_percent + incidence_rate,
  data = cancer_model_data
)

cancer_model
tidy(cancer_model)
```

The poverty coefficient is the estimated difference in mortality for a
one-percentage-point difference in poverty among counties with the same
incidence rate. The incidence coefficient is interpreted similarly while
holding poverty constant. These are associations, not causal effects.

### Your turn: identify what the coefficients mean

Use the model and its coefficient table to write down:

1. the response variable and its units;
2. the units of each predictor;
3. a context sentence for the poverty slope;
4. a context sentence for the incidence slope; and
5. which slope is estimated more precisely relative to its size.

Explain why a standard error describes uncertainty in a coefficient estimate,
whereas the residual standard deviation describes county-to-county scatter
around the fitted model.

## 2. Confidence intervals and coefficient p-values

```{r}
tidy(cancer_model, conf.int = TRUE)
```

The table contains a two-sided t test for each coefficient. The rows correspond
to these null hypotheses:

| Row | Null hypothesis | Interpretation if the coefficient is not zero |
|:--|:--|:--|
| (Intercept) | $H_0:\beta_0=0$ | fitted mortality is zero when both predictors are zero |
| poverty_percent | $H_0:\beta_{\mathrm{poverty}}=0$ | poverty has no adjusted linear association at fixed incidence |
| incidence_rate | $H_0:\beta_{\mathrm{incidence}}=0$ | incidence has no adjusted linear association at fixed poverty |

A two-sided p-value is the probability, assuming the stated null and model,
of seeing a t statistic at least as far from zero as the observed one. It is
not the probability that the null is true.

### Demonstration: test the poverty slope

```{r}
poverty_row <- tidy(cancer_model, conf.int = TRUE) |>
  filter(term == "poverty_percent")
poverty_row
```

For this test, $H_0:\beta_{\mathrm{poverty}}=0$ and
$H_A:\beta_{\mathrm{poverty}}\ne0$. A small p-value is evidence against a
zero adjusted poverty slope. The confidence interval gives the range of
coefficient values compatible with the model and sample under repeated-sampling
reasoning. Statistical evidence does not by itself establish a causal effect
or practical importance.

### Your turn: test the incidence slope and the intercept

Use the table to answer both sets of questions:

1. For incidence_rate, state $H_0$ and $H_A$, interpret the 95% confidence
   interval, and decide whether the evidence supports a zero slope.
2. For (Intercept), state what $H_0:\beta_0=0$ means and decide whether the
   p-value provides evidence against it.
3. Explain why the intercept conclusion is usually not substantively useful:
   zero poverty and zero incidence are outside the observed predictor ranges.
4. Use “reject” or “fail to reject”; do not say “accept the null.”
5. Explain why a confidence interval excluding zero agrees with rejection of a
   two-sided 5% zero-coefficient test.

Do not describe a large p-value as proof that a coefficient is exactly zero.

## 3. Confidence versus prediction intervals

Use exactly the same hypothetical county for both calculations:

```{r}
new_county <- tibble(
  poverty_percent = 15,
  incidence_rate = 60
)
```

### Demonstration: confidence interval for an average

```{r}
confidence_interval <- as_tibble(
  predict(cancer_model, newdata = new_county, interval = "confidence")
) |>
  mutate(interval = "Confidence: average county")

confidence_interval |>
  select(interval, fit, lwr, upr)
```

A confidence interval describes uncertainty about the average mortality for all
counties with these predictor values.

### Your turn: prediction interval for one county

Calculate the prediction interval for the same new_county:

```{r}
# Your code here
```

Explain why the fitted value is the same as in the confidence interval but the
prediction interval is wider. State whether either interval would be trustworthy
for a county far outside the observed poverty or incidence ranges.

## 4. Interpret the LINE diagnostics

The plots in this section are supplied and executable. Your task is to read the
pattern and explain its consequence; do not recreate the plots from scratch.

### Demonstration: L — what linearity asks

The `performance` linearity plot shows the model residual pattern against fitted
values. A roughly horizontal reference line supports a straight-line mean
structure; a visible curve suggests that the conditional mean is not adequately
represented by straight-line terms.

```{r}
check_model(cancer_model, check = "linearity")
```

### Your turn: interpret the linearity plot

Is the reference line strongly curved or broadly near zero? Explain whether
linearity appears to be a major concern and what a visible curve would imply
for the fitted mean.

### Demonstration: I — what independence asks

No plot in this data file can establish independence. County observations may
share geography, services, environmental exposures or smoking patterns. Row
order is not a spatial ordering.

### Your turn: assess independence from study design

Explain why independence remains an unresolved limitation here. What additional
sampling or spatial information would make an independence assessment more
credible?

### Demonstration: N — what normality asks

`check_normality()` reports a Shapiro–Wilk test and its plot shows the residual
Q-Q diagnostic. With a large sample, the formal test can be highly sensitive to
small departures, so visual inspection of the centre and tails remains the
primary diagnostic.

```{r}
normality_check <- check_normality(cancer_model)
normality_check
plot(normality_check, type = "qq")
```

### Your turn: interpret the normality plot

Describe the centre and both tails. In particular, note whether the extreme
negative residuals and the upper tail follow the reference line. Explain which
parts of small-sample inference or individual prediction are most sensitive to
tail departures.

### Demonstration: E — what equal variance asks

`check_heteroscedasticity()` reports a formal test for non-constant variance and
its plot shows the residual spread across fitted values. Interpret the test
alongside the plot rather than treating either as an automatic pass/fail rule.

```{r}
heteroscedasticity_check <- check_heteroscedasticity(cancer_model)
heteroscedasticity_check
plot(heteroscedasticity_check)
```

### Your turn: interpret the equal-variance plot

Describe how the vertical spread changes as fitted mortality increases. If the
spread grows, explain why conventional standard errors and prediction intervals
need caution.

## 5. Influence and one transformation

### Demonstration: screen for influential observations

Cook's distance is a screening tool, not an automatic deletion rule.

```{r}
influence_check <- check_outliers(
  cancer_model,
  method = "cook",
  threshold = list(cook = 4 / nobs(cancer_model))
)

influence_check
plot(influence_check)

influence_data <- augment(cancer_model, data = cancer_model_data) |>
  mutate(observation = row_number())

influence_data |>
  slice_max(.cooksd, n = 5) |>
  select(observation, area_name, .cooksd, .hat, .std.resid)
```

### Your turn: investigate influence

Identify the largest Cook's-distance observations. List what you would check
in the source records before a sensitivity refit. Explain why exceeding the
informal $4/n$ line does not justify deleting a county automatically.

### Demonstration: compare a log-incidence alternative

```{r}
cancer_log_model <- lm(
  Mortality_Rate_num ~ poverty_percent + log(incidence_rate),
  data = cancer_model_data
)

tibble(
  model = c("Original incidence", "Log incidence"),
  r_squared = c(
    glance(cancer_model)$r.squared,
    glance(cancer_log_model)$r.squared
  ),
  residual_sd = c(
    glance(cancer_model)$sigma,
    glance(cancer_log_model)$sigma
  )
)
```

```{r}
check_model(cancer_model, check = "linearity")
check_model(cancer_log_model, check = "linearity")

plot(check_heteroscedasticity(cancer_model))
plot(check_heteroscedasticity(cancer_log_model))
```

### Your turn: decide about the transformation

Use the fit table and the performance diagnostic plots to decide whether logging
incidence is worth retaining. Discuss fit, residual shape, spread, and
interpretability. If you interpret the logged coefficient, remember that a 1%
increase in incidence changes the fitted response by approximately 0.01 times
that coefficient, in mortality units.

## 6. Independent workflow: income and insurance

Use the same cancer dataset but a fresh pair of predictors. Complete the model
and interpretation steps below. Diagnostic plotting code is supplied; interpret
those plots rather than writing plotting code yourself.

### Your turn: prepare and fit the independent model

Create income_data with complete rows for area_name, Mortality_Rate_num,
med_income, all_without_insur and popestimate2015. Derive:

- income_thousands = med_income / 1000; and
- uninsured_percent = 100 * all_without_insur / popestimate2015.

Then fit and print income_model:

```{r}
# Your code here
```

Use tidy(income_model, conf.int = TRUE) and answer:

1. What is the adjusted mortality difference for a $1,000 income difference?
2. What is the adjusted mortality difference for a one-percentage-point
   uninsured-rate difference?
3. Are the slopes or the intercept plausibly zero according to their tests?
4. Is the intercept at zero income and zero uninsured percentage a meaningful
   county comparison?

### Your turn: predict with uncertainty

For a county with income_thousands = 50 and uninsured_percent = 10, calculate
both a confidence interval for the average and a prediction interval for one
county:

```{r}
# Your code here
```

Explain the difference in target and width.

### Your turn: interpret the independent diagnostics

After fitting income_model, run this supplied code:

```{r}
check_model(income_model, check = "linearity")

income_normality_check <- check_normality(income_model)
income_normality_check
plot(income_normality_check, type = "qq")

income_heteroscedasticity_check <- check_heteroscedasticity(income_model)
income_heteroscedasticity_check
plot(income_heteroscedasticity_check)

income_influence_check <- check_outliers(
  income_model,
  method = "cook",
  threshold = list(cook = 4 / nobs(income_model))
)

income_influence_check
plot(income_influence_check)

income_results <- augment(income_model, data = income_data)

income_results |>
  mutate(observation = row_number()) |>
  slice_max(.cooksd, n = 5) |>
  select(observation, area_name, .cooksd, .hat, .std.resid)
```

Write one paragraph covering L, I, N and E, and identify any influential
observations that require checking. Do not claim that a plot proves independence.

### Your turn: compare a transformed independent model

Fit a second model with log(income_thousands) in place of
income_thousands. Compare R-squared, residual standard deviation and the
performance diagnostic plots. Finish with a recommendation that distinguishes
association from causation and qualifies inference when LINE concerns remain.

```{r}
# Your code here: fit income_log_model and compare fit.
# Use check_model() for linearity and plot(check_heteroscedasticity())
# for both the original and transformed models.
```

## Conclusion

Summarise the primary model's adjusted associations, what each p-value tests,
the confidence-versus-prediction distinction, the most important diagnostic
concern and whether the transformation is worth retaining. Explain what extra
data or study design would make the independence assumption more credible.
