---
title: "Workshop Week 10: Vehicle Relationships"
subtitle: "Exploring relationships and fitting a first linear model"
author:
  - name: "Nick Knowlton"
  - name: "297.101 Statistical Data Science"
date-meta: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    code-block-bg: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

Fuel use affects the running cost and energy demand of a vehicle. A larger
engine, greater power or higher mass might suggest higher fuel use, but vehicles
with similar specifications can still perform differently. Comparing these
features lets us see which patterns are strong enough for a straight line to
summarise and how much variation the line leaves unexplained.

[Waka Kotahi's vehicle-fleet data](https://www.nzta.govt.nz/resources/new-zealand-motor-vehicle-register-statistics/new-zealand-vehicle-fleet-open-data-sets)
provide a snapshot of vehicles recorded on the Motor Vehicle Register, while
[EECA's guide to vehicle energy-economy labels](https://www.eeca.govt.nz/regulations/vehicle-emissions-and-energy-economy-labelling-programme/about-the-vehicle-emissions-and-energy-economy-labelling-programme/)
explains the standard fuel-economy measure in litres per 100 km. This workshop
uses those standard test values to compare vehicle specifications rather than
the fuel use experienced by a particular driver.

::: {.callout-note}
## Today’s question

How much can engine size, power and gross mass tell us about a vehicle's fuel
use?
:::

Scatterplots reveal the direction, form and strength of each relationship;
slopes quantify meaningful average differences in fuel use; and residuals show
how much variation remains after fitting a line. Work through each new idea
with engine size, apply it to engine power, and then use the complete workflow
to investigate gross vehicle mass.

## Learning objectives

By the end of this workshop you should be able to:

- describe the direction, form and strength of a relationship from a plot;
- fit and inspect a simple linear model with `lm()`;
- extract and interpret the fitted coefficients;
- obtain and examine fitted values and residuals; and
- apply the complete workflow to a new explanatory variable.

## 1. Meet and explore the data

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(broom)

vehicles <- read_csv("nz_vehicles_2025.csv", show_col_types = FALSE)
```

Each row is one **unique vehicle specification** found in the New Zealand
vehicle data. The data were prepared from Waka Kotahi’s vehicle fleet data as
at **31 July 2026**. They contain model-year 2025 petrol passenger cars and
vans with complete measurements; repeated specifications have been removed.

The fuel-use values come from a standard test, not from the driving of a
particular person. An association in these data does not show that changing
one vehicle feature would cause another feature to change.

Waka Kotahi publishes the source under the [CC-BY 4.0 International
licence](https://creativecommons.org/licenses/by/4.0/). Source details and
caveats appear on the vehicle-fleet page linked above.

```{r}
#| label: glimpse
glimpse(vehicles)
```

| Column | Meaning | Units |
|:--|:--|:--|
| `make`, `model`, `body_type` | vehicle description | — |
| `engine_l` | engine displacement | litres |
| `power_kw` | engine power rating | kilowatts |
| `gross_mass_kg` | maximum permitted laden mass | kilograms |
| `fuel_use_l_per_100km` | combined-cycle fuel use | litres per 100 km |

### Demonstration: engine size

Before fitting a line, start with a scatterplot and a LOESS smooth. The smooth
helps us see the general shape without requiring it to be a straight line.

```{r}
#| label: engine-loess
#| fig-cap: "Combined fuel use and engine size for 2025 petrol passenger vehicles."
#| fig-alt: "Scatterplot with engine size in litres on the horizontal axis and combined fuel use in litres per 100 kilometres on the vertical axis, showing a generally increasing relationship."
#| out-width: "100%"
ggplot(vehicles, aes(x = engine_l, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Do larger engines tend to use more fuel?"
  ) +
  theme_minimal(base_size = 13)
```

The correlation coefficient measures the direction and strength of a
*linear* relationship. It has no units and ranges from -1 to 1.

```{r}
#| label: engine-correlation
vehicles |>
  summarise(
    engine_correlation = cor(engine_l, fuel_use_l_per_100km)
  )
```

::: {.callout-note title="What to look for"}

Look for the direction, form and strength of the relationship. Correlation
summarises the strength of a *linear* relationship; it has no units and is not
the slope of the fitted line.

:::

Describe the direction, form and strength shown by the plot. Then explain what
the sign and size of the correlation add to that description.

::: {.callout-important title="Discuss"}

Is the relationship positive or negative? Is it approximately linear? How much
scatter remains around the smooth? What does the correlation add?

:::

::: {#wk10-answer-engine-exploration .callout-tip collapse="true" title="Worked interpretation — open after discussion"}

Fuel use increases as engine size increases, and the relationship is strong and
approximately linear. The correlation is about 0.901, which supports that
description and gives a unit-free measure of linear strength. The points still
show vehicle-to-vehicle variation around the smooth, so engine size does not
determine fuel use by itself. Correlation is not a slope: it does not have
L/100 km per litre units and does not state the expected change in fuel use for
one additional litre.

:::

### Your turn: engine power

Repeat the exploration with `power_kw` as the explanatory variable. Create a
scatterplot with a LOESS smooth, use labels with units, and calculate the
correlation between power and fuel use.

```{r}
#| label: power-loess
# Use the engine-size plot as your template.
# Your code here
```

```{r}
#| label: power-correlation
# Use the engine-size correlation as your template.
# Your code here
```

Compare the engine-size and power plots. Which appears more nearly linear?
Which has the larger correlation? Remember that correlation is not a slope.

## 2. Fit and inspect the models

A simple linear model describes the average response as a straight-line
function of one explanatory variable:

$$
\mathsf{Mean}(y) = a + bx.
$$

### Demonstration: engine-size model

Fit the model, save it, and then print the saved model object. Keeping the
model in an object lets us return to it throughout the workshop.

```{r}
#| label: fit-engine-model
engine_model <- lm(
  fuel_use_l_per_100km ~ engine_l,
  data = vehicles
)

engine_model
```

The next plot changes the flexible LOESS smooth to the straight line fitted by
`lm()`.

```{r}
#| label: plot-engine-model
#| fig-cap: "The fitted engine-size linear model."
#| fig-alt: "Scatterplot of engine size and combined fuel use with a fitted straight line."
#| out-width: "100%"
ggplot(vehicles, aes(x = engine_l, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "A straight-line model for engine size and fuel use"
  ) +
  theme_minimal(base_size = 13)
```

::: {.callout-note title="What to look for"}

The straight line estimates average fuel use at each engine size. Compare the
line with the individual points: the vertical gaps are variation the model
does not explain.

:::

::: {.callout-important title="Discuss"}

What does the line capture? Where are observations above and below it? Would
the line predict every vehicle exactly?

:::

::: {#wk10-answer-engine-model .callout-tip collapse="true" title="Worked interpretation — open after discussion"}

The fitted line describes the average fuel use associated with engine size in
these vehicle specifications. It captures the overall increasing trend, while
the spread around it represents differences among vehicles that engine size
alone does not explain. The line is therefore useful for summarising the
association, but it is not an exact rule for an individual vehicle and does not
show that a larger engine causes higher fuel use.

:::

### Your turn: power model

Fit `fuel_use_l_per_100km ~ power_kw`, save it as `power_model`, print the
model object, and make the corresponding scatterplot with a fitted straight
line.

```{r}
#| label: fit-power-model
# Use fit-engine-model as your template.
# Your code here
```

```{r}
#| label: plot-power-model
# Use plot-engine-model as your template.
# Your code here
```

Look at both fitted lines. What does each line capture, and what variation is
still visible around it?

## 3. Extract and interpret the coefficients

### Demonstration: engine-size coefficients

`coef()` extracts the intercept and slope from the saved model.

```{r}
#| label: engine-coefficients
engine_coef <- coef(engine_model)
engine_coef
```

::: {.callout-note title="What to look for"}

Read the intercept and slope together with their units. Check whether zero is
within the observed engine-size range before giving the intercept a practical
meaning.

:::

Use these values to write the fitted equation:

$$
\widehat{\text{fuel use}} = a + b\,\text{engine size}.
$$

Then answer:

1. What does the slope mean in **L/100 km per litre of engine size**?
2. What does the intercept mean mathematically?
3. Is an engine size of zero within the observed data? How does that affect
   the practical interpretation of the intercept?

::: {.callout-important title="Discuss"}

What is the fitted equation? What change in fuel use corresponds to one litre
of engine size? Is zero litres represented by these data?

:::

::: {#wk10-answer-engine-coefficients .callout-tip collapse="true" title="Worked interpretation — open after discussion"}

The fitted equation is approximately

$$
\widehat{\text{fuel use}} = 4.070 + 1.886\,\text{engine size}.
$$

A one-litre difference is associated with about 1.886 L/100 km higher average
fuel use. The intercept is the mathematical line value at zero litres, but zero
is outside the observed range (0.998–6.749 litres), so it has no direct vehicle
interpretation. Both coefficients describe an association among specifications
rather than a causal effect.

:::

### Your turn: power coefficients

Extract the coefficients from `power_model`, save them as `power_coef`, and
write the fitted equation using `power_kw`.

```{r}
#| label: power-coefficients
# Use engine-coefficients as your template.
# Your code here
```

Interpret the slope in **L/100 km per kilowatt**. Also explain the expected
difference associated with 100 kW so that the magnitude is easier to
communicate. Is zero kW within the observed data?

::: {.callout-important}
The coefficients describe associations among these vehicle specifications.
Do not turn either slope into a causal claim.
:::

## 4. Extract and inspect the residuals

For each vehicle specification, the fitted value is the value on the fitted
line. The residual is the observed value minus the fitted value:

$$
e_i = y_i - \hat{y}_i.
$$

### Demonstration: engine-size residuals

`augment()` returns the original variables together with useful model columns,
including `.fitted` and `.resid`.

```{r}
#| label: engine-augment
engine_results <- augment(engine_model)

engine_results |>
  select(engine_l, fuel_use_l_per_100km, .fitted, .resid) |>
  slice_head(n = 8)
```

The orange segments below show a sample of residuals as vertical gaps from the
observed values to the fitted line.

```{r}
#| label: engine-residual-gaps
#| fig-cap: "Observed fuel use, the engine-size fitted line, and a sample of residual gaps."
#| fig-alt: "Scatterplot with a fitted straight line and vertical segments connecting selected observations to the line."
#| out-width: "100%"
set.seed(297101)
engine_gap_data <- engine_results |>
  slice_sample(n = min(35, nrow(engine_results)))

ggplot(vehicles, aes(x = engine_l, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.35) +
  geom_line(
    data = engine_results |> arrange(engine_l),
    aes(y = .fitted),
    colour = "#2C7FB8",
    linewidth = 1
  ) +
  geom_segment(
    data = engine_gap_data,
    aes(xend = engine_l, yend = .fitted),
    alpha = 0.65,
    colour = "#D95F02"
  ) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Residuals are the vertical gaps from observations to the line"
  ) +
  theme_minimal(base_size = 13)
```

::: {.callout-note title="What to look for"}

`.fitted` is the value on the fitted line and `.resid` is observed minus
fitted. A positive residual is above the line; a negative residual is below it.

:::

::: {.callout-important title="Discuss"}

Which segments are positive or negative? What are the units of a residual? Why
are the gaps vertical?

:::

::: {#wk10-answer-engine-residual-gaps .callout-tip collapse="true" title="Worked interpretation — open after discussion"}

Each segment joins an observed fuel-use value to its fitted value at the same
engine size. The vertical distance is the residual

$$
e_i = y_i - \widehat{y}_i.
$$

Residuals are measured in L/100 km. Points above the line have positive
residuals and points below it have negative residuals. Showing the gaps makes
clear that the model predicts a conditional average; individual vehicles can
lie either side of that average.

:::

Plot all residuals against their fitted values, then summarise the mean
residual and compare the original response variance with the residual
variance.

```{r}
#| label: engine-residual-plot
#| fig-cap: "Residuals from the engine-size model versus fitted fuel-use values."
#| fig-alt: "Engine-size model residuals plotted against fitted values with a horizontal zero line."
#| out-width: "100%"
ggplot(engine_results, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.55) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted fuel use (L/100 km)",
    y = "Residual (L/100 km)",
    title = "What the engine-size model leaves unexplained"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: engine-residual-summary
engine_results |>
  summarise(
    mean_residual = mean(.resid),
    response_variance = var(fuel_use_l_per_100km),
    residual_variance = var(.resid)
  )
```

::: {.callout-note title="What to look for"}

Check that the residual mean is close to zero, then compare response variance
with residual variance. A smaller residual variance means the line has removed
some variation; it does not prove every model assumption holds.

:::

::: {.callout-important title="Discuss"}

How much variation remains after fitting the line? Does a near-zero mean prove
the model is adequate? What pattern would concern you in the residual plot?

:::

::: {#wk10-answer-engine-residual-summary .callout-tip collapse="true" title="Worked interpretation — open after discussion"}

The mean residual is essentially zero, as expected for least squares with an
intercept. The response variance is about 7.136 and the engine-model residual
variance is about 1.341, so engine size accounts for substantial variation while
leaving meaningful vehicle-specific variation unexplained. A near-zero mean is
an algebraic property of this fit; it does not establish linearity, constant
variance or accurate individual predictions.

:::

### Your turn: power residuals

Repeat the same core checks with `power_model`:

1. Create `power_results` with `augment()` and inspect `.fitted` and `.resid`.
2. Plot the power-model residuals against fitted values.
3. Calculate the mean residual, response variance and residual variance.
4. Compare the power residual variance with the engine residual variance.

```{r}
#| label: power-augment
# Use engine-augment as your template.
# Your code here
```

```{r}
#| label: power-residual-plot
# Use engine-residual-plot as your template.
# Your code here
```

```{r}
#| label: power-residual-summary
# Use engine-residual-summary as your template.
# Your code here
```

Describe what the power model leaves unexplained. This is an early look at
model leftovers; formal LINE checks belong to Week 12.

## 5. Independent workflow: gross vehicle mass

Now complete the entire workflow independently using `gross_mass_kg` as the
explanatory variable. Reuse only operations already demonstrated above.

### A. Explore

- Plot fuel use against gross mass with a LOESS smooth.
- Calculate and interpret the correlation.

```{r}
#| label: mass-explore
# Your code here
```

### B. Fit and inspect

- Fit `fuel_use_l_per_100km ~ gross_mass_kg` as `mass_model`.
- Print the model and plot its fitted straight line.

```{r}
#| label: mass-fit
# Your code here
```

### C. Extract and interpret coefficients

- Save `coef(mass_model)` as `mass_coef`.
- Write the fitted equation.
- Interpret the slope in **L/100 km per kilogram** and, more usefully, per
  **1000 kg**.

```{r}
#| label: mass-coefficients
# Your code here
```

### D. Inspect residuals

- Create `mass_results` with `augment()`.
- Plot residuals against fitted values.
- Calculate the mean residual, response variance and residual variance.

```{r}
#| label: mass-residuals
# Your code here
```

### E. Compare all three models

Using your plots, correlations and residual variances, compare engine size,
power and gross mass as single predictors of fuel use. Name the strongest
predictor in these data, give evidence, and state what remains unexplained.

::: {.callout-important}
## Scope of today’s conclusion

You can describe an association and how well a line summarises these data.
You cannot yet quantify uncertainty in the slope, test a null hypothesis, or
construct an interval for a future vehicle. Those are Week 11 questions.
:::

## 6. Looking ahead

The evidence here supports conclusions about the direction and strength of
associations and about how well each fitted line summarises these vehicle
specifications. The remaining questions concern uncertainty: how stable is an
estimated slope, and how wide should an interval be when predicting fuel use
for another vehicle?

Next week we will use model output to study point estimates, confidence
intervals, hypothesis tests, prediction intervals and variance explained.
