---
title: "Workshop Week 12: Inference and Assumptions"
subtitle: "Answer guide — uncertainty, intervals and diagnostics"
author:
  - name: "Nick Knowlton"
  - name: "297.101 Statistical Data Science"
date-meta: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

This guide follows Workshop_Week12_Assumptions.qmd. It contains the completed
code and interpretations.

In Weeks 10 and 11 we used vehicle data to build simple and multiple regression
models, interpret coefficients, standard errors, p-values and $R^2$, and make
fitted predictions. This week we apply the same ideas to a new dataset and ask
a harder question: how much can we trust the model's estimates and predictions?

We will use county-level lung-cancer mortality data to introduce confidence
intervals, prediction intervals and regression diagnostics. The change in
dataset is deliberate: the aim is to apply the same regression workflow in a
new context rather than rely on one familiar example.

The data come from a broader project that combined county socioeconomic and
cancer information. Noah Rippner's
[Python data-preparation code](https://github.com/nrippner/can_clin_trails_prep_r/blob/master/extract_zips_10.py)
shows how data from different sources were cleaned and joined by place. Those
joins let us examine the same county's socioeconomic conditions and mortality,
although the resulting associations do not establish causes or describe
individual people.

This workshop asks how uncertain the model estimates are and whether the
model's assumptions are reasonable.

## Today's question

How well can we trust our model's estimates and predictions of county
lung-cancer mortality?

## Learning objectives

By the end of this workshop you should be able to:

- use standard errors to interpret uncertainty in regression coefficients;
- interpret confidence intervals and hypothesis tests;
- distinguish confidence intervals for an average from prediction intervals for
  an individual county;
- interpret supplied LINE diagnostic plots and use study design to assess a
  model; and
- evaluate one transformed alternative without relying on R-squared alone.

## 1. Fit and understand one primary cancer model

**What to do:** Fit the multiple regression model and identify the response and
predictors. Interpret each slope while holding the other predictor constant.
Check that you can distinguish the coefficient estimate from its standard error
and residual variation.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(broom)
library(performance)
library(see)

cancer <- read_csv("regression/cancer_data.csv", show_col_types = FALSE)

cancer_model_data <- cancer |>
  select(
    area_name, Mortality_Rate_num, all_poverty, popestimate2015,
    Incidence_Rate_num
  ) |>
  filter(
    is.finite(Mortality_Rate_num),
    is.finite(all_poverty),
    is.finite(popestimate2015),
    is.finite(Incidence_Rate_num),
    popestimate2015 > 0
  ) |>
  mutate(
    poverty_percent = 100 * all_poverty / popestimate2015,
    incidence_rate = Incidence_Rate_num
  ) |>
  select(area_name, Mortality_Rate_num, poverty_percent, incidence_rate)

glimpse(cancer_model_data)
```

There are 2,562 complete counties after the incidence-rate filter. The
predictor ranges are approximately 2.60–46.49% poverty and 13.5–203.7
incidence cases per 100,000.

### Demonstration: fit the primary model

```{r}
cancer_model <- lm( Mortality_Rate_num ~ poverty_percent + incidence_rate,  data = cancer_model_data)

cancer_model
tidy(cancer_model)
```

The fitted model is approximately

$$
\widehat{\mathrm{mortality}} =
2.527 + 0.265(\mathrm{poverty\ percent}) +
0.665(\mathrm{incidence\ rate}).
$$

The poverty coefficient is the estimated mortality difference for a one
percentage-point poverty difference among counties with the same incidence.
The incidence coefficient is the estimated mortality difference for one
additional incidence case per 100,000 among counties with the same poverty.
Both are adjusted associations.

### Your turn: identify what the coefficients mean

**Your turn:** Identify the response variable and both predictors. Interpret each
slope in the units of the problem, remembering to hold the other predictor
constant. Which coefficient appears to be estimated more precisely relative to
its size?

The response is lung-cancer mortality in deaths per 100,000. Poverty is a
percentage and incidence is cases per 100,000. The incidence slope is estimated
more precisely relative to its size: its standard error is about 0.008 versus
an estimate about 0.665, while the poverty standard error is about 0.023 versus
an estimate about 0.265. A standard error describes sampling uncertainty in a
coefficient; the residual standard deviation (about 6.88 deaths per 100,000)
describes county-to-county scatter around the fitted model.

### A quick return to $R^2$ and adjusted $R^2$

**What to do:** Compare ordinary $R^2$ and adjusted $R^2$ for the fitted
model. Recall that ordinary $R^2$ cannot decrease when predictors are added,
even if the new predictor contributes little. Adjusted $R^2$ applies a
penalty for additional model terms, so it can decrease when extra complexity is
not justified by improved fit.

```{r}
glance(cancer_model) |>
  select(r.squared, adj.r.squared)
```

**Your turn:** Why can ordinary $R^2$ increase when a useless predictor is
added, while adjusted $R^2$ can decrease? What would it suggest if $R^2$
increased but adjusted $R^2$ decreased?

Ordinary $R^2$ rewards any improvement in in-sample fit, whereas adjusted $R^2$
also penalises additional model terms. If $R^2$ rises but adjusted $R^2$ falls,
the new predictor has not improved the fit enough to justify the extra
complexity.

Adjusted $R^2$ is useful when comparing models with different numbers of
predictors, but it is still an in-sample summary. It does not replace checking
residuals or assumptions.

## 2. Confidence intervals and coefficient p-values

**What to do:** Use the coefficient table to interpret the 95% confidence
interval and p-value for each coefficient. State the null hypothesis being
tested, decide whether the data provide evidence against a zero coefficient,
and keep statistical evidence separate from effect size and causality.

Last week we used standard errors and p-values to describe uncertainty and
evidence for adjusted coefficients. We can now extend that idea by attaching a
confidence interval to each coefficient.

```{r}
tidy(cancer_model, conf.int = TRUE)
```

The table contains a two-sided t test for each coefficient. The rows correspond
to these null hypotheses:

| Row | Null hypothesis | Interpretation if the coefficient is not zero |
|:--|:--|:--|
| (Intercept) | $H_0:\beta_0=0$ | fitted mortality is zero when both predictors are zero |
| poverty_percent | $H_0:\beta_{\mathrm{poverty}}=0$ | poverty has no adjusted linear association at fixed incidence |
| incidence_rate | $H_0:\beta_{\mathrm{incidence}}=0$ | incidence has no adjusted linear association at fixed poverty |

A p-value is calculated under the stated null and model. It is not the
probability that the null is true. Use reject or fail to reject; never say that
a large p-value proves a null.

### Demonstration: test the poverty slope

```{r}
poverty_row <- tidy(cancer_model, conf.int = TRUE) |>
  filter(term == "poverty_percent")
poverty_row
```

The poverty estimate is about 0.265 deaths per 100,000 per percentage point,
with a 95% interval of approximately [0.220, 0.311] and p-value below 2e-16.
For $H_0:\beta_{\mathrm{poverty}}=0$ versus
$H_A:\beta_{\mathrm{poverty}}\ne0$, reject the zero-slope null. This is strong
evidence of an adjusted linear association in this sample and model; it is not
evidence of causation or practical importance by itself.

### Your turn: test the incidence slope and the intercept

**Your turn:** Interpret the incidence coefficient, its 95% confidence interval
and p-value. State the null hypothesis and your conclusion. Then do the same
for the intercept. Is the intercept test statistically significant, and is it
substantively useful?

The incidence estimate is about 0.665 deaths per 100,000 per additional
incidence case per 100,000. Its 95% interval is approximately [0.649, 0.681]
and its p-value is below 2e-16, so reject
$H_0:\beta_{\mathrm{incidence}}=0$ conditional on poverty.

The intercept estimate is about 2.527, with 95% interval [1.375, 3.679] and
p-value about 1.77e-5, so its test also rejects $H_0:\beta_0=0$. However,
this asks about mortality when both poverty and incidence are zero. Both values
are outside the observed ranges, so the intercept test has little substantive
meaning. A confidence interval excluding zero agrees with rejection of the
corresponding two-sided 5% zero-coefficient test. A p-value does not measure
effect size and does not make an observational association causal.

## 3. Confidence versus prediction intervals

**What to do:** Make a prediction for the specified county characteristics
using both a confidence interval and a prediction interval. Compare their
widths and explain why the prediction interval for one county is wider than the
confidence interval for the average response.

```{r}
new_county <- tibble(
  poverty_percent = 15,
  incidence_rate = 60
)
```

### Demonstration: confidence interval for an average

```{r}
confidence_interval <- as_tibble(
  predict(cancer_model, newdata = new_county, interval = "confidence")
) |>
  mutate(interval = "Confidence: average county")

confidence_interval |>
  select(interval, fit, lwr, upr)
```

The fitted average mortality is approximately 46.39 deaths per 100,000. The
confidence interval targets the average mortality for all counties at these
predictor values; it is relatively narrow because it only reflects uncertainty
in the fitted mean.

### Your turn: prediction interval for one county

**Your turn:** Calculate a prediction interval for one county with these same
predictor values. Compare it with the confidence interval above. Which is
wider, and why?

```{r}
prediction_interval <- as_tibble(
  predict(cancer_model, newdata = new_county, interval = "prediction")
) |>
  mutate(interval = "Prediction: individual county")

bind_rows(confidence_interval, prediction_interval) |>
  select(interval, fit, lwr, upr)
```

The fitted value is identical because both intervals use the same new county and
same fitted model. The prediction interval is wider because it includes
county-to-county residual variation as well as uncertainty in the average.
Neither interval should be trusted as an extrapolation far outside the observed
poverty or incidence ranges.

## 4. Interpret the LINE diagnostics

**What to do:** Work through the LINE assumptions one at a time. For each
assumption, state what it means, identify what evidence the supplied diagnostic
provides, and decide whether there is a meaningful concern. Do not treat any
diagnostic plot as an automatic pass/fail test.

The plots are supplied and executable. Students interpret them; they do not
recreate them.

### Demonstration: L — what linearity asks

The `performance` linearity plot shows the model residual pattern against fitted
values. A roughly horizontal reference line supports a straight-line mean
structure; a visible curve suggests that the conditional mean is not adequately
represented by straight-line terms.

```{r}
check_model(cancer_model, check = "linearity")
```

### Your turn: interpret the linearity plot

**Your turn:** Does the reference line show clear curvature, or is a
straight-line mean structure reasonably supported?

The reference line is broadly near zero with only a modest bend. Linearity is
not the dominant concern for this primary model, although no diagnostic is an
automatic pass/fail test. A persistent curve would mean that straight-line
terms miss part of the conditional mean.

### Demonstration: I — what independence asks

There is no valid row-order plot for independence. Counties may share
geography, services, environmental exposures or regional smoking patterns.

### Your turn: assess independence from study design

**Your turn:** Can independence be assessed from these diagnostic plots alone?
What features of the study design could make counties non-independent?

Independence remains unresolved because the CSV has no spatial adjacency or
sampling-design information. Additional spatial data, clustered sampling
information or a design that samples independent units would be needed to assess
or model dependence. A plot cannot prove independence.

### Demonstration: N — what normality asks

`check_normality()` reports a Shapiro–Wilk test and its plot shows the residual
Q-Q diagnostic. With a large sample, the formal test can be highly sensitive to
small departures, so visual inspection of the centre and tails remains the
primary diagnostic.

```{r}
normality_check <- check_normality(cancer_model)
normality_check
plot(normality_check, type = "qq")
```

### Your turn: interpret the normality plot

**Your turn:** How closely do the residuals follow the reference line? Pay
particular attention to the centre and tails.

The centre is closer to the reference line than the tails. Both tails depart,
with particularly extreme negative residuals and a heavy upper tail. Tail
departures matter most for small-sample model-based inference and individual
prediction; they are less damaging to a rough description of the centre in this
large sample. The Shapiro–Wilk result should not be treated as a mechanical
pass/fail decision in a sample this large.

### Demonstration: E — what equal variance asks

`check_heteroscedasticity()` reports a formal test for non-constant variance and
its plot shows the residual spread across fitted values. Interpret the test
alongside the plot rather than treating either as an automatic pass/fail rule.

```{r}
heteroscedasticity_check <- check_heteroscedasticity(cancer_model)
heteroscedasticity_check
plot(heteroscedasticity_check)
```

### Your turn: interpret the equal-variance plot

**Your turn:** Does the residual spread stay approximately constant as the
fitted values increase, or does it change systematically?

The vertical spread and smooth increase toward larger fitted mortality. This is
evidence of non-constant variance. Conventional standard errors and prediction
intervals therefore need caution, even though the sample is large.

## 5. Influence and one transformation

**What to do:** Identify observations that may have unusual influence on the
fitted model, then compare the original model with one transformed alternative.
Do not delete observations automatically or choose a transformation using
$R^2$ alone. Consider the residual pattern, influence, interpretability and
purpose of the model.

### Demonstration: screen for influential observations

```{r}
influence_check <- check_outliers(
  cancer_model,
  method = "cook",
  threshold = list(cook = 4 / nobs(cancer_model))
)

influence_check
plot(influence_check)

influence_data <- augment(cancer_model, data = cancer_model_data) |>
  mutate(observation = row_number())

influence_data |>
  slice_max(.cooksd, n = 5) |>
  select(observation, area_name, .cooksd, .hat, .std.resid)
```

### Your turn: investigate influence

**Your turn:** Identify the counties with the largest Cook's distances. Which
observations would you investigate further, and what would you check before
deciding whether any observation should be excluded?

The largest Cook's distances are approximately:

| County | Cook's distance | Standardized residual |
|:--|--:|--:|
| Covington city, Virginia | 0.252 | -7.90 |
| Union County, Florida | 0.152 | -4.11 |
| Williamsburg city, Virginia | 0.143 | -9.19 |
| Greene County, Alabama | 0.065 | -5.44 |

Check the source records, units, missing-value handling, predictor leverage and
data-entry plausibility. Then refit as a sensitivity analysis. Crossing the
informal 4/n line is a prompt for inspection, not a reason to delete a county
automatically.

### Demonstration: compare a log-incidence alternative

```{r}
cancer_log_model <- lm(
  Mortality_Rate_num ~ poverty_percent + log(incidence_rate),
  data = cancer_model_data
)

tibble(
  model = c("Original incidence", "Log incidence"),
  r_squared = c(
    glance(cancer_model)$r.squared,
    glance(cancer_log_model)$r.squared
  ),
  residual_sd = c(
    glance(cancer_model)$sigma,
    glance(cancer_log_model)$sigma
  )
)
```

```{r}
check_model(cancer_model, check = "linearity")
check_model(cancer_log_model, check = "linearity")

plot(check_heteroscedasticity(cancer_model))
plot(check_heteroscedasticity(cancer_log_model))
```

The original model has R-squared about 0.767 and residual SD about 6.88. The
log-incidence model has slightly lower R-squared and slightly larger residual
SD. Unless its residual pattern materially improves, the original scale is the
more direct interpretation. A transformation should be retained only when its
overall diagnostic and prediction benefits outweigh the loss of direct
interpretability.

### Your turn: decide about the transformation

**Your turn:** Compare the original and log-incidence models. Look at $R^2$,
residual SD, residual curvature and spread. Does the transformation improve the
model enough to justify making the coefficient harder to interpret?

The fit table and both performance diagnostic plots should be compared,
including variance, curvature, tails and interpretability rather than choosing by
R-squared alone. A 1% increase in incidence changes the fitted response by
approximately 0.01 times the logged-incidence coefficient, in mortality units.

## 6. Independent workflow: income and insurance

**Your turn: complete the full workflow using income and insurance.**

1. Prepare the analysis variables and fit the multiple regression model.
2. Interpret both adjusted coefficients in their original units.
3. Examine the standard errors, 95% confidence intervals and p-values.
4. Interpret $R^2$ and adjusted $R^2$.
5. Predict mortality for the specified predictor values using both confidence and prediction intervals.
6. Interpret the LINE diagnostics.
7. Identify any influential observations and explain what you would investigate.
8. Fit the transformed alternative and decide whether it materially improves the model.
9. Justify your decision using diagnostics, interpretability and model fit rather than $R^2$ alone.

### Your turn: prepare and fit the independent model

Fit the model and interpret both adjusted slopes, confidence intervals and
p-values.

```{r}
income_data <- cancer |>
  select(
    area_name, Mortality_Rate_num, med_income, all_without_insur,
    popestimate2015
  ) |>
  filter(
    is.finite(Mortality_Rate_num),
    is.finite(med_income),
    is.finite(all_without_insur),
    is.finite(popestimate2015),
    popestimate2015 > 0
  ) |>
  mutate(
    income_thousands = med_income / 1000,
    uninsured_percent = 100 * all_without_insur / popestimate2015
  ) |>
  select(area_name, Mortality_Rate_num, income_thousands, uninsured_percent)

income_model <- lm(
  Mortality_Rate_num ~ income_thousands + uninsured_percent,
  data = income_data
)

glimpse(income_data)
income_model
tidy(income_model, conf.int = TRUE)
```

There are 2,809 complete counties. The income slope is approximately -0.532
deaths per 100,000 per $1,000 income difference, holding uninsured percentage
constant. Its 95% interval is approximately [-0.574, -0.490] and its p-value is
below 2e-16. The uninsured slope is approximately -0.150 deaths per 100,000
per percentage point, holding income constant. Its 95% interval is
approximately [-0.258, -0.041] and its p-value is about 0.0068. Both slopes
provide evidence against a zero adjusted coefficient at the 5% level.

The intercept is approximately 79.99 and its p-value is below 2e-16, but it
describes zero income and zero uninsured percentage, outside the data and not
a meaningful county comparison. The negative uninsured coefficient is an
observational conditional association; it is not evidence that lack of
insurance reduces mortality.

### Your turn: predict with uncertainty

Calculate both intervals and explain why they differ in width.

```{r}
new_income_county <- tibble(
  income_thousands = 50,
  uninsured_percent = 10
)

income_confidence <- as_tibble(
  predict(income_model, newdata = new_income_county, interval = "confidence")
) |>
  mutate(interval = "Confidence: average county")

income_prediction <- as_tibble(
  predict(income_model, newdata = new_income_county, interval = "prediction")
) |>
  mutate(interval = "Prediction: individual county")

bind_rows(income_confidence, income_prediction) |>
  select(interval, fit, lwr, upr)
```

The fitted mortality is approximately 51.91. The confidence interval is about
[51.37, 52.44], while the prediction interval is about [27.30, 76.52]. The
prediction interval is wider because it includes residual variation for one
county.

### Your turn: interpret the independent diagnostics

Work through LINE and influence just as you did for the primary model. Which
assumptions cause the most concern?

```{r}
check_model(income_model, check = "linearity")

income_normality_check <- check_normality(income_model)
plot(income_normality_check, type = "qq")

income_heteroscedasticity_check <- check_heteroscedasticity(income_model)
plot(income_heteroscedasticity_check)

income_influence_check <- check_outliers(
  income_model,
  method = "cook",
  threshold = list(cook = 4 / nobs(income_model))
)

income_influence_check
plot(income_influence_check)

income_results <- augment(income_model, data = income_data)

income_results |>
  mutate(observation = row_number()) |>
  slice_max(.cooksd, n = 5) |>
  select(observation, area_name, .cooksd, .hat, .std.resid)
```

The independent model usually shows a broad residual cloud, increasing spread
toward larger fitted values, non-normal tails and a small number of influential
counties. Students should report what their plots show and note that
independence still requires study-design evidence.

### Your turn: compare a transformed independent model

Compare the original and log-income models. Decide whether the transformation
produces a worthwhile improvement in fit or diagnostics.

```{r}
income_log_model <- lm(
  Mortality_Rate_num ~ log(income_thousands) + uninsured_percent,
  data = income_data
)

tibble(
  model = c("Original income", "Log income"),
  r_squared = c(
    glance(income_model)$r.squared,
    glance(income_log_model)$r.squared
  ),
  residual_sd = c(
    glance(income_model)$sigma,
    glance(income_log_model)$sigma
  )
)

check_model(income_model, check = "linearity")
check_model(income_log_model, check = "linearity")

plot(check_heteroscedasticity(income_model))
plot(check_heteroscedasticity(income_log_model))
```

The original income model has R-squared about 0.202 and residual SD about 12.55.
The log-income model has R-squared about 0.221 and residual SD about 12.39, a
modest fit improvement. Students should still check whether curvature and
spread improve enough to justify the less direct log-income interpretation.
Neither model establishes causal effects.

## Conclusion

The primary model provides adjusted associations, but confidence and prediction
intervals answer different questions. The coefficient p-values test whether
slopes or the intercept are zero under the model; they do not prove hypotheses
or causation. The diagnostic plots show broadly reasonable linearity, increasing
variance, tail departures and unresolved independence. Influential counties and
any transformation should be checked through sensitivity and model-purpose
reasoning.
