---
title: "Workshop Week 10: Simple Regression"
subtitle: "Student workbook: relationships, fitted lines and prediction"
author:
  - name: "Nick Knowlton"
  - name: "297.101 Statistical Data Science"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

For the first part of this course we focused on working with data in R and RStudio,
including using `ggplot2` to explore and visualise relationships. More recently we
have looked at probability distributions, sampling variation and the central limit
theorem, which give us a way to think about uncertainty in estimates from data.

In this workshop we start bringing those ideas together. We will use plots to explore
relationships between vehicle characteristics and fuel use, then fit our first
regression models to describe those relationships numerically. Along the way we will
interpret the fitted line, coefficients, standard errors, p-values, $R^2$, predictions
and residuals.

[Waka Kotahi's vehicle-fleet data](https://www.nzta.govt.nz/resources/new-zealand-motor-vehicle-register-statistics/new-zealand-vehicle-fleet-open-data-sets)
provide a snapshot of vehicles recorded on the Motor Vehicle Register, while
[EECA's guide to vehicle energy-economy labels](https://www.eeca.govt.nz/regulations/vehicle-emissions-and-energy-economy-labelling-programme/about-the-vehicle-emissions-and-energy-economy-labelling-programme/)
explains the standard fuel-economy measure in litres per 100 km. The supplied
extract contains 467 unique vehicle specifications, prepared from Waka Kotahi's
vehicle-fleet data as at **31 July 2026** and containing model-year 2025 petrol
passenger cars and vans with complete measurements.

## Today's question

How much can one vehicle specification tell us about fuel use?

## Learning objectives

By the end of this workshop you should be able to:

- recognise an approximately linear association from a scatterplot;
- fit and interpret a simple linear regression model;
- interpret the slope, intercept, standard error, p-value and $R^2$;
- calculate and interpret a fitted value; and
- define and interpret a residual.

## 1. Explore engine size and power

**What to do:** Start by looking at the data before fitting a model. For each
scatterplot, describe the direction, strength and shape of the relationship and
look for anything unusual. Then calculate the correlation and ask whether a
straight line looks like a reasonable summary.

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(broom)

vehicles <- read_csv("nz_vehicles_2025.csv", show_col_types = FALSE)
glimpse(vehicles)
```

Each row is one unique 2025 petrol vehicle specification. Fuel use is the
combined-cycle test value in litres per 100 km, so the plots describe
associations among specifications rather than causal effects or real-world
consumption.

### Demonstration: engine size

Let's start with the flexible smoother we learned earlier in the course to get a feel
for the relationship between engine size and fuel use. A reminder that this
lets us look at the *shape* of the relationship first and decide whether a linear
model is a reasonable summary.

```{r}
ggplot(vehicles, aes(engine_l, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Do larger engines tend to use more fuel?"
  ) +
  theme_minimal(base_size = 13)

vehicles |>
  summarise(engine_correlation = cor(engine_l, fuel_use_l_per_100km))
```

What does the correlation add to the plot? Remember that it describes linear
strength, has no units, and is not a slope.

### Your turn: engine power

Repeat the same process for engine power. Before looking at the correlation, use
the plot to decide whether the relationship is positive or negative, whether it
looks approximately linear, and whether it appears stronger or weaker than the
engine-size relationship.

```{r}
# Your code here
```

Which relationship appears more nearly linear? Which has the larger
correlation? Do not compare the correlations as if they were slopes.

## 2. Fit simple linear models

**What to do:** Fit a simple linear regression model with fuel use as the
response and engine size as the explanatory variable. Identify the intercept and
slope, then add the fitted regression line to the scatterplot. Repeat the same
process using engine power.

The relationship looks approximately linear, so we can now summarise it using
a straight-line model.

A simple linear model describes the average response as a straight-line
function of one explanatory variable:

$$
\widehat{y} = \widehat{\beta}_0 + \widehat{\beta}_1 x.
$$

```{r}
engine_model <- lm(
  fuel_use_l_per_100km ~ engine_l,
  data = vehicles
)

engine_model
```

```{r}
ggplot(vehicles, aes(engine_l, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "A straight-line model for engine size and fuel use"
  ) +
  theme_minimal(base_size = 13)
```

There are many possible straight lines we could draw through these data.
`lm()` uses least squares: it chooses the intercept and slope that minimise the
sum of the squared vertical distances between the observed values and the fitted
line. This is the same idea as minimising the mean squared error we looked at
previously.

What is `lm()` changing when it searches for the best-fitting line?

The line captures the increasing average trend, but individual vehicles remain
above and below it. It is a summary of the conditional mean, not an exact rule.

### Your turn: power model

Fit the power model yourself before looking at the answer. Compare its fitted
line with the LOESS smooth from Section 1. Does a straight line still seem like
a reasonable summary?

```{r}
# Your code here
```

```{r}
# Your code here
```

## 3. Coefficients and R²

**What to do:** Use `tidy()` to examine the fitted coefficients. For each
coefficient, identify the estimate, standard error and p-value. Interpret the
slope in the units of the problem rather than just reporting the number. Then
use $R^2$ to describe how much of the observed variation in fuel use is
accounted for by the model.

```{r}
engine_coefficients <- tidy(engine_model)
engine_coefficients
```

Write the fitted equation. Which number is the intercept, and which number is
the slope? Interpret the slope in L/100 km per litre of engine size. The
intercept is the mathematical line value at zero litres; check whether zero is
represented by these data before giving it a practical meaning.

Use the `estimate`, `std.error`, `statistic`, and `p.value` columns to explain
what each part of the coefficient table tells us. A small p-value is evidence
against a zero population coefficient under the model; it does not by itself
show that an association is large, important or causal.

We already calculated the correlation between engine size and fuel use. Square
it and compare it with the model $R^2$. What do you notice? Write one sentence
about how much variation the engine-size model explains.

```{r}
# Your code here: square the correlation and compare it with model R^2
```

### Power coefficients and fit

Now interpret the power model yourself. What does the slope mean in L/100 km
per kW? Is that a useful scale, or would reporting the change per 100 kW be
easier to understand? Compare its standard error, p-value and $R^2$ with the
engine-size model.

```{r}
# Your code here
```

## 4. Make a prediction

**What to do:** Use the fitted equation to calculate the fitted fuel use for a
vehicle with a 2.0 L engine by hand. Then use `predict()` to obtain the same
value in R. The two answers should agree.

```{r}
# Your code here
```

What does 7.84 L/100 km represent? Does the model say that every 2.0 L vehicle
will have this fuel use?

## 5. What is a residual?

For one vehicle, the residual is observed fuel use minus fitted fuel use:

$$
e_i = y_i-\hat{y}_i
$$

A positive residual means the observed fuel use was above the fitted value; a
negative residual means it was below.

**What to do:** Use `augment()` to compare the observed fuel use, fitted value
and residual for several vehicles. Check that

$$
\text{residual}=\text{observed}-\text{fitted}.
$$

Find one positive residual and one negative residual and explain what each means
in the context of fuel use.

```{r}
# Your code here
```

On the plot, the residual is the vertical distance from each observed value to
the fitted line. Before looking at the residual column, use the graph to decide
whether a point should have a positive or negative residual.

```{r}
# Your code here
```

What are the units of a residual? A near-zero average residual is expected from
least squares, but it does not by itself establish that the model is adequate.
Formal residual diagnostics belong to Week 12.

## 6. Independent workflow: gross vehicle mass

**Your turn: complete the full workflow using gross vehicle mass.**

1. Plot gross mass against fuel use and use a smooth to examine the shape of the relationship.
2. Describe the direction, strength and approximate form of the relationship.
3. Calculate the correlation.
4. Fit a simple linear regression model and save it as `mass_model`.
5. Interpret the slope. Report it per 1000 kg rather than per kilogram if that makes the effect easier to understand.
6. Examine the standard error and p-value for the slope.
7. Report and interpret $R^2$.
8. Predict the fitted mean fuel use for a 2000 kg vehicle.
9. Compare the gross-mass model with the engine-size and power models. Which has the largest in-sample $R^2$?

```{r}
# Your code here
```

## Looking ahead

Next week we return to these models and add body type as another predictor. This
introduces multiple regression, reference categories and adjusted coefficients.
The coefficient estimates, standard errors and p-values are interpreted in the
same general way as today, but the meaning of each coefficient changes because
it is now conditional on the other variables in the model.

[View the worked solutions online](https://knowlton.co.nz/297101/labs/Workshop_Week10_Relationships_guide.html).
