---
title: "Workshop Week 11: Regression Testing"
subtitle: "Answer guide — testing associations and making predictions"
author:
  - name: "Nick Knowlton"
  - name: "297.101 Statistical Data Science"
date-meta: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    code-block-bg: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
resources:
  - regression/cancer_data.csv.gz
---

Cancer mortality is not evenly distributed across places. Comparing US
counties lets us ask whether counties with greater poverty or lower median
income also tend to have higher lung-cancer mortality. We might find a clear
trend, substantial differences between counties with similar socioeconomic
conditions, or a relationship that a straight line summarises poorly.

The broader project assembled open data on socioeconomic conditions, cancer
incidence, mortality and clinical trials by county. Noah Rippner's
[original data-preparation code](https://github.com/nrippner/can_clin_trails_prep_r/blob/master/extract_zips_10.py)
shows how records from different sources were cleaned and joined through county
identifiers. Those joins make this investigation possible: mortality and
socioeconomic measurements must refer to the same place before their
relationship can be studied.

::: {.callout-note}
## Today's question

What can county data tell us about the association between poverty, median
income and lung-cancer mortality?

:::

This analysis develops two related ideas. A fitted line estimates an average
association across counties; a prediction for another county must also allow
for the large differences that remain between counties. The results therefore
move from visible patterns to effect sizes, uncertainty, variation explained
and prediction. Because the data are observational and county-level, they
describe associations between places rather than causes or outcomes for
individual people.

## 1. What patterns appear across counties?

::: {.callout-important title="Question"}
What does each row represent, and are the variables on scales that allow
meaningful comparisons between counties?
:::

::: {.callout-note title="What to look for"}
Check how many rows are retained and keep track of the units. A raw poverty
count still contains information about county population; a rate supports a
fairer comparison between counties of different sizes.
:::

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(purrr)
library(broom)
library(GGally)

cancer <- read_csv("regression/cancer_data.csv.gz", show_col_types = FALSE)
cancer_clean <- cancer |>
  select(Mortality_Rate_num, all_poverty, med_income, popestimate2015) |>
  filter(if_all(everything(), is.finite), popestimate2015 > 0) |>
  mutate(
    pov_norm = all_poverty / popestimate2015 * 100000,
    poverty_percent = pov_norm / 1000
  )
tibble(rows_in_source = nrow(cancer), rows_retained = nrow(cancer_clean))
```

| Variable | Meaning | Units |
|:--|:--|:--|
| `Mortality_Rate_num` | Lung-cancer mortality rate | deaths per 100,000 people |
| `all_poverty` | Residents below the poverty threshold | people |
| `pov_norm` | Population-adjusted poverty count | people per 100,000 |
| `poverty_percent` | Population-adjusted poverty rate | percent |
| `med_income` | Median household income | US dollars |

#### Result

All 2,809 rows are retained. Each row represents a US county with complete
values for these variables. The supplied [data file](regression/cancer_data.csv.gz)
combines lung-cancer information with US Census Bureau socioeconomic measures.

### Why use a poverty rate?

::: {.callout-important title="Question"}
If two counties each have 10,000 residents in poverty, do they have the same
level of poverty?
:::

::: {.callout-note title="What to look for"}
Compare the percentages after dividing by population. The predictor and
response do not need matching units; the units determine how the slope will be
interpreted.
:::

```{r}
#| label: count-rate-example
tibble(
  county = c("County A", "County B"),
  all_poverty = c(10000, 10000),
  population = c(50000, 1000000)
) |>
  mutate(
    poverty_per_100000 = all_poverty / population * 100000,
    poverty_percent = poverty_per_100000 / 1000
  )
```

#### Result

The same count represents 20% poverty in County A and 1% in County B. Dividing
by population changes the predictor from a mixture of county size and poverty
into a comparable rate.

### Poverty, income and mortality together

::: {.callout-important title="Question"}
Which pairs of variables appear related, and how much spread remains around
each trend?
:::

::: {.callout-note title="What to look for"}
Read distributions on the diagonal, scatterplots below and correlations above.
In each scatterplot, look for direction, form, strength and unusual points;
the correlation only summarises linear direction and strength.
:::

```{r}
#| label: draftsman-plot
#| fig-cap: "Distributions and pairwise relationships among poverty rate, median income and lung-cancer mortality."
#| fig-alt: "A three by three matrix with distributions on the diagonal, scatterplots below and correlations above."
#| fig-width: 9
#| fig-height: 9
#| out-width: "100%"
draftsman_data <- cancer_clean |>
  transmute(
    `Poverty (%)` = poverty_percent,
    `Median income ($)` = med_income,
    `Mortality (deaths per 100,000)` = Mortality_Rate_num
  )
draftsman_plot <- ggpairs(
  draftsman_data,
  progress = FALSE,
  diag = list(continuous = wrap("densityDiag", alpha = 0.5)),
  lower = list(continuous = wrap("points", alpha = 0.25, size = 0.55)),
  upper = list(continuous = wrap("cor", size = 4, stars = FALSE))
) + theme_minimal(base_size = 11)
print(draftsman_plot)
```

#### Result

Poverty and mortality have a moderate positive association
(\(r=0.398\)). The cloud is broad: counties at a similar poverty rate can have
quite different mortality rates, so poverty alone will not predict county
mortality closely.

### A closer look at median income

::: {.callout-important title="Question"}
Do counties with higher median income tend to have lower lung-cancer mortality?
:::

::: {.callout-note title="What to look for"}
Use the full-sized plot to judge the downward trend, spread and possible
curvature. Then compare that visual evidence with the correlation.
:::

```{r}
#| label: income-exploration
#| fig-cap: "Lung-cancer mortality and median household income."
#| fig-alt: "Scatterplot of income and mortality with a generally decreasing smooth curve."
#| out-width: "100%"
ggplot(cancer_clean, aes(x = med_income, y = Mortality_Rate_num)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Median household income (US dollars)",
    y = "Lung-cancer mortality (deaths per 100,000)",
    title = "Do counties with higher median income have lower mortality?"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: income-correlation
cancer_clean |>
  summarise(income_mortality_correlation =
              cor(med_income, Mortality_Rate_num))
```

#### Result

The association is moderately negative (\(r=-0.447\)): counties with higher
median income tend to have lower mortality. The considerable vertical spread
shows that income alone does not determine a county's mortality rate.

## 2. How large are the associations?

### Poverty and mortality

::: {.callout-important title="Question"}
What difference in fitted mortality corresponds to a 10-percentage-point
difference in poverty rate?
:::

The `tidy()` output reports an `Estimate` and `Std. Error` for each term. The
estimate is the fitted coefficient in the units of the model; its standard
error describes sampling uncertainty. These columns supply the quantities
used for the t statistic, confidence interval and p-value.

::: {.callout-note title="What to look for"}
The model uses people in poverty per 100,000, so 10 percentage points is a
change of 10,000 in `pov_norm`. Check whether the intercept describes a
realistic county.
:::

```{r}
#| label: fit-poverty-model
poverty_model <- lm(Mortality_Rate_num ~ pov_norm, data = cancer_clean)
tidy(poverty_model)
```

```{r}
#| label: interpret-poverty-slope
poverty_slope <- coef(poverty_model)[["pov_norm"]]
tibble(
  change_in_poverty_percentage_points = 10,
  fitted_change_in_mortality = poverty_slope * 10000
)
```

#### Result

A 10-percentage-point higher poverty rate is associated with about 9.13 more
deaths per 100,000. Zero poverty is outside the observed range, so the
intercept is mainly mathematical. This slope describes an association between
counties; it is not the effect of changing a county's poverty rate.

### Income and mortality

::: {.callout-important title="Question"}
What difference in fitted mortality corresponds to a $10,000 difference in
median household income?
:::

::: {.callout-note title="What to look for"}
Rescale the small per-dollar coefficient before interpreting it. Raw poverty
and income slopes cannot be compared as effect sizes because their predictor
units differ.
:::

```{r}
#| label: fit-income-model
income_model <- lm(Mortality_Rate_num ~ med_income, data = cancer_clean)
tidy(income_model)
```

```{r}
#| label: interpret-income-slope
income_slope <- coef(income_model)[["med_income"]]
tibble(
  change_in_income = 10000,
  fitted_change_in_mortality = income_slope * 10000
)
```

#### Result

A $10,000 higher median income is associated with about 5.05 fewer deaths per
100,000. This is an average difference between counties with different median
incomes, not evidence that raising income would cause mortality to fall.

## 3. How uncertain are those estimates?

### Uncertainty in the poverty association

::: {.callout-important title="Question"}
What range of population slopes is compatible with the poverty estimate, and
how strong is the evidence against no linear association?
:::

::: {.callout-note title="What to look for"}
Compare the manual t interval with `confint()`, then check its units and whether
it contains zero. The p-value is calculated under a zero-slope null and the
model assumptions.
:::

```{r}
#| label: poverty-manual-confidence-interval
poverty_slope_row <- tidy(poverty_model) |> filter(term == "pov_norm")
critical_t <- qt(0.975, df = df.residual(poverty_model))
poverty_ci_manual <- poverty_slope_row |>
  transmute(
    lower = estimate - critical_t * std.error,
    estimate,
    upper = estimate + critical_t * std.error
  )
poverty_ci_manual
confint(poverty_model, "pov_norm", level = 0.95)
```

For a fitted slope, the 95% confidence interval is

$$
\widehat{\beta}_1 \;\pm\; t^* \times
\operatorname{SE}\!\left(\widehat{\beta}_1\right).
$$

Here \(\widehat{\beta}_1\) is the estimated slope and
\(\operatorname{SE}(\widehat{\beta}_1)\) is its standard error. The critical
value \(t^*\) leaves 2.5% in each tail of a t distribution with \(n-2\)
degrees of freedom. The model has \(n-2\) residual degrees of freedom because
it estimates two coefficients: the intercept and slope.

```{r}
#| label: poverty-manual-p-value
poverty_t <- poverty_slope_row$estimate / poverty_slope_row$std.error
poverty_p_manual <- 2 * pt(-abs(poverty_t), df.residual(poverty_model))
tibble(
  t_statistic = poverty_t,
  manual_p_value = poverty_p_manual,
  tidy_p_value = poverty_slope_row$p.value
)
```

#### Result

For 10 percentage points, the estimated mortality difference is 9.13 deaths
per 100,000, with a 95% confidence interval from 8.35 to 9.91. The interval
excludes zero and the p-value gives very strong evidence against a zero
population slope under the model assumptions. In repeated sampling, about 95%
of intervals constructed this way would contain the fixed population slope.
Neither result establishes causation or practical importance.

### Uncertainty in the income association

::: {.callout-important title="Question"}
Does the interval for the income slope include zero, and does the manual
calculation agree with the model output?
:::

::: {.callout-note title="What to look for"}
Check the signs of both interval endpoints and interpret the final interval per
$10,000. Relate the p-value to a zero population slope rather than to the size
or importance of the association.
:::

```{r}
#| label: income-manual-confidence-interval
income_slope_row <- tidy(income_model) |> filter(term == "med_income")
income_critical_t <- qt(0.975, df = df.residual(income_model))
income_ci_manual <- income_slope_row |>
  transmute(
    lower = estimate - income_critical_t * std.error,
    estimate,
    upper = estimate + income_critical_t * std.error
  )
income_ci_manual
confint(income_model, "med_income", level = 0.95)
```

```{r}
#| label: income-manual-p-value
income_t <- income_slope_row$estimate / income_slope_row$std.error
income_p_manual <- 2 * pt(-abs(income_t), df.residual(income_model))
tibble(
  t_statistic = income_t,
  manual_p_value = income_p_manual,
  tidy_p_value = income_slope_row$p.value
)
```

#### Result

For $10,000, the estimated difference is -5.05 deaths per 100,000, with a 95%
confidence interval from -5.43 to -4.68. The manual and model calculations
agree, the interval excludes zero, and the small p-value is strong evidence
against a zero slope, subject to the model assumptions.

## 4. How much variation remains?

### Variation after fitting poverty

::: {.callout-important title="Question"}
How much of the observed variation in mortality does the poverty model account
for, and how much remains between counties?
:::

::: {.callout-note title="What to look for"}
Connect response variance, residual variance and the resulting proportion
explained. A model can have a clearly nonzero slope while still leaving most
county-to-county variation unexplained.
:::

```{r}
#| label: poverty-r-squared
poverty_results <- augment(poverty_model)
poverty_results |>
  summarise(
    response_variance = var(Mortality_Rate_num),
    residual_variance = var(.resid),
    variance_explained = 1 - residual_variance / response_variance
  ) |>
  mutate(model_r_squared = glance(poverty_model)$r.squared)
```

#### Result

The residual-variance calculation agrees with the reported \(R^2\). Poverty
accounts for 15.8% of sample mortality variation, so 84.2% remains associated
with other differences, unmeasured factors and random variation. This is a
descriptive decomposition, not a causal claim.

### Comparing poverty and income

::: {.callout-important title="Question"}
Which of the two single-predictor models accounts for more variation in these
counties?
:::

::: {.callout-note title="What to look for"}
Both models use the same response and the same counties, so their \(R^2\)
values are directly comparable. Also consider whether either value suggests
close prediction for an individual county.
:::

```{r}
#| label: income-r-squared
income_results <- augment(income_model)
income_fit_summary <- income_results |>
  summarise(
    response_variance = var(Mortality_Rate_num),
    residual_variance = var(.resid),
    variance_explained = 1 - residual_variance / response_variance
  ) |>
  mutate(model_r_squared = glance(income_model)$r.squared)
income_fit_summary
```

#### Result

Income accounts for 20.0% of sample mortality variation, compared with 15.8%
for poverty. Income is the stronger of these two single predictors by this
measure, but about four-fifths of the variation remains unexplained. The
comparison does not identify a cause.

## 5. What can we predict for another county?

### Staying within the observed data

::: {.callout-important title="Question"}
Which poverty rates and median incomes are supported by the observed counties?
:::

::: {.callout-note title="What to look for"}
Check the predictor ranges before calculating a prediction. Values inside the
range are interpolations; values outside it require the fitted line to continue
where there are no supporting observations.
:::

```{r}
#| label: observed-ranges
cancer_clean |>
  summarise(
    minimum_poverty_percent = min(poverty_percent),
    maximum_poverty_percent = max(poverty_percent),
    minimum_income = min(med_income),
    maximum_income = max(med_income)
  )
```

#### Result

Observed poverty ranges from 2.60% to 46.49%. The predictions at 20% poverty
and $50,000 income below are within the corresponding observed ranges.

### Mortality at 20% poverty

::: {.callout-important title="Question"}
What mortality rate does the poverty model predict at 20%, and why are there
two different kinds of interval?
:::

::: {.callout-note title="What to look for"}
Convert 20% to 20,000 people per 100,000 and compare manual substitution with
`predict()`. A confidence interval describes mean mortality at this poverty
rate; a prediction interval allows for another county's deviation from that
mean. Both widen as the confidence level increases.
:::

```{r}
#| label: poverty-point-prediction
poverty_20 <- tibble(pov_norm = 20000)
poverty_coef <- coef(poverty_model)
manual_prediction <- poverty_coef[["(Intercept)"]] +
  poverty_coef[["pov_norm"]] * poverty_20$pov_norm
tibble(manual_prediction)
predict(poverty_model, poverty_20)
```

```{r}
#| label: poverty-intervals
predict(poverty_model, poverty_20, interval = "confidence", level = 0.95)
predict(poverty_model, poverty_20, interval = "prediction", level = 0.95)
```

```{r}
#| label: poverty-confidence-levels
map_dfr(c(0.90, 0.95, 0.99), \(confidence_level) {
  as_tibble(predict(
    poverty_model, poverty_20,
    interval = "prediction", level = confidence_level
  )) |> mutate(level = confidence_level, .before = 1)
})
```

#### Result

Manual substitution and `predict()` agree at 56.61 deaths per 100,000. The
confidence interval for mean county mortality is narrower than the prediction
interval for one additional county because the latter also includes
county-to-county variation. Both intervals widen from 90% to 95% to 99%. This
is a prediction of a county mortality rate, not an individual person's outcome.

### Mortality at $50,000 median income

::: {.callout-important title="Question"}
How precisely can the model estimate average mortality at $50,000, and how
precisely can it predict mortality for one county with that median income?
:::

::: {.callout-note title="What to look for"}
Compare the point prediction with the two 95% intervals. The difference in
their widths reveals the distinction between estimating a mean relationship
and predicting one county.
:::

```{r}
#| label: income-point-prediction
income_50000 <- tibble(med_income = 50000)
income_coef <- coef(income_model)
income_manual_prediction <- income_coef[["(Intercept)"]] +
  income_coef[["med_income"]] * income_50000$med_income
tibble(income_manual_prediction)
predict(income_model, income_50000)
```

```{r}
#| label: income-intervals
predict(income_model, income_50000, interval = "confidence", level = 0.95)
predict(income_model, income_50000, interval = "prediction", level = 0.95)
```

```{r}
#| label: income-confidence-levels
map_dfr(c(0.90, 0.95, 0.99), \(confidence_level) {
  as_tibble(predict(
    income_model, income_50000,
    interval = "prediction", level = confidence_level
  )) |> mutate(level = confidence_level, .before = 1)
})
```

#### Result

At $50,000, fitted mortality is 51.58 deaths per 100,000. The 95% confidence
interval for mean mortality is 51.10–52.06, while the 95% prediction interval
for one county is 26.94–76.21. The average trend is estimated quite precisely,
but income alone gives a much less precise prediction for another county.

### What happens beyond the observed range?

::: {.callout-important title="Question"}
Can predictions at 0% and 75% poverty be trusted simply because R can calculate
them?
:::

::: {.callout-note title="What to look for"}
Compare both requested values with the observed range of 2.60% to 46.49%.
Numerical output does not supply data where none were observed.
:::

```{r}
#| label: poverty-extrapolation
predict(
  poverty_model,
  newdata = tibble(pov_norm = c(0, 75000)),
  interval = "prediction"
)
```

#### Result

Both values are extrapolations. R evaluates the fitted equation, but the data
cannot show whether the line or its uncertainty behaves similarly beyond the
observed poverty range.

## Putting the results together

Counties with $10,000 higher median income have an estimated 5.05 fewer
lung-cancer deaths per 100,000, with a 95% confidence interval from -5.43 to
-4.68 and strong evidence against a zero slope under the model. Yet income
explains only 20.0% of sample mortality variation. At $50,000, the fitted mean
is 51.58 and is estimated precisely, while the interval for another county is
far wider at 26.94–76.21. A clear average association can therefore coexist
with substantial uncertainty about an individual county.

These patterns could reflect access to healthcare, smoking, age distributions,
environmental exposure or other differences between counties. The models do
not separate those explanations, and the observational data cannot show that
changing poverty or income would itself change mortality.

::: {.callout-tip title="Optional question"}
If poverty and income are related, would including both necessarily improve
predictions for new counties?

Not necessarily. The two predictors may contain overlapping information, so a
model fitted to the current data can become more complicated without gaining
much new predictive information. Improvement should be judged using predictions
for data that were not used to fit the model, while also checking whether the
model's assumptions are reasonable.
:::

Next week we examine regression assumptions and use diagnostic plots to assess
how well these models support inference and prediction.
