---
title: "Workshop Week 10: Vehicle Relationships"
subtitle: "Answer guide — exploring relationships and fitting a first linear model"
author:
  - name: "Nick Knowlton"
  - name: "297.101 Statistical Data Science"
date-meta: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    code-block-bg: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

Fuel use affects the running cost and energy demand of a vehicle. A larger
engine, greater power or higher mass might suggest higher fuel use, but vehicles
with similar specifications can still perform differently. Comparing these
features lets us see which patterns are strong enough for a straight line to
summarise and how much variation the line leaves unexplained.

[Waka Kotahi's vehicle-fleet data](https://www.nzta.govt.nz/resources/new-zealand-motor-vehicle-register-statistics/new-zealand-vehicle-fleet-open-data-sets)
provide a snapshot of vehicles recorded on the Motor Vehicle Register, while
[EECA's guide to vehicle energy-economy labels](https://www.eeca.govt.nz/regulations/vehicle-emissions-and-energy-economy-labelling-programme/about-the-vehicle-emissions-and-energy-economy-labelling-programme/)
explains the standard fuel-economy measure in litres per 100 km. This analysis
uses those standard test values to compare vehicle specifications rather than
the fuel use experienced by a particular driver.

::: {.callout-note}
## Today’s question

How much can engine size, power and gross mass tell us about a vehicle's fuel
use?
:::

This guide revisits the analysis in
[`Workshop_Week10_Relationships.qmd`](Workshop_Week10_Relationships.qmd).
Scatterplots reveal the direction, form and strength of each relationship;
slopes quantify meaningful average differences in fuel use; and residuals show
how much variation remains after fitting a line. The analysis begins with
engine size, repeats the same ideas using engine power, and then applies the
full workflow to gross vehicle mass.

## 1. Meet and explore the data

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(broom)

vehicles <- read_csv("nz_vehicles_2025.csv", show_col_types = FALSE)
```

```{r}
#| label: glimpse
glimpse(vehicles)
```

There are **467** unique vehicle specifications. The data were prepared from
Waka Kotahi’s vehicle fleet data as at **31 July 2026** and contain model-year
2025 petrol passenger cars and vans with complete measurements. Standard test
values are not the observed fuel use of individual drivers.

Waka Kotahi publishes the source under the [CC-BY 4.0 International
licence](https://creativecommons.org/licenses/by/4.0/). Source details and
caveats appear on the vehicle-fleet page linked above.

| Column | Meaning | Units |
|:--|:--|:--|
| `make`, `model`, `body_type` | vehicle description | — |
| `engine_l` | engine displacement | litres |
| `power_kw` | engine power rating | kilowatts |
| `gross_mass_kg` | maximum permitted laden mass | kilograms |
| `fuel_use_l_per_100km` | combined-cycle fuel use | litres per 100 km |

### Engine size

::: {.callout-important title="Question"}
Is the relationship positive? Is it approximately linear? How much scatter remains?
:::

::: {.callout-note title="What to look for"}
Look for direction, form and strength. Correlation is unit-free linear strength,
not a slope.
:::

```{r}
#| label: engine-loess
#| fig-cap: "Combined fuel use and engine size for 2025 petrol passenger vehicles."
#| fig-alt: "Scatterplot with engine size in litres on the horizontal axis and combined fuel use in litres per 100 kilometres on the vertical axis, showing a generally increasing relationship."
#| out-width: "100%"
ggplot(vehicles, aes(x = engine_l, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Do larger engines tend to use more fuel?"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: engine-correlation
vehicles |>
  summarise(
    engine_correlation = cor(engine_l, fuel_use_l_per_100km)
  )
```

::: {#wk10-answer-engine-exploration .callout-tip title="Worked interpretation"}
Fuel use increases with engine size in a strong, approximately linear association.
The correlation is about 0.901. The remaining scatter shows that engine size
does not determine each vehicle’s fuel use, and correlation is not a slope.
:::

### Engine power

```{r}
#| label: power-loess
#| fig-cap: "Combined fuel use and engine power for 2025 petrol passenger vehicles."
#| fig-alt: "Scatterplot with power in kilowatts on the horizontal axis and combined fuel use in litres per 100 kilometres on the vertical axis, showing a generally increasing relationship."
#| out-width: "100%"
ggplot(vehicles, aes(x = power_kw, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Engine power (kW)",
    y = "Combined fuel use (L/100 km)",
    title = "Do more powerful engines tend to use more fuel?"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: power-correlation
vehicles |>
  summarise(
    power_correlation = cor(power_kw, fuel_use_l_per_100km)
  )
```

Power also has a strong positive relationship with fuel use. Its correlation
is approximately **0.884**, slightly smaller than the engine-size
correlation. Both plots are broadly linear, with some departures and scatter.

## 2. Fit and inspect the models

### Engine-size model

::: {.callout-important title="Question"}
What does the line capture? Where are observations above and below it? Would
the line predict every vehicle exactly?
:::

::: {.callout-note title="What to look for"}
The straight line estimates average fuel use at each engine size. Compare it
with the individual points and their vertical gaps.
:::

```{r}
#| label: fit-engine-model
engine_model <- lm(
  fuel_use_l_per_100km ~ engine_l,
  data = vehicles
)

engine_model
```

```{r}
#| label: plot-engine-model
#| fig-cap: "The fitted engine-size linear model."
#| fig-alt: "Scatterplot of engine size and combined fuel use with a fitted straight line."
#| out-width: "100%"
ggplot(vehicles, aes(x = engine_l, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "A straight-line model for engine size and fuel use"
  ) +
  theme_minimal(base_size = 13)
```

::: {#wk10-answer-engine-model .callout-tip title="Worked interpretation"}
The fitted line describes the average fuel use associated with engine size in
these vehicle specifications. It captures the overall increasing trend, while
the spread around it represents differences among vehicles that engine size
alone does not explain. It summarises an association, rather than an exact
rule for an individual vehicle or a causal mechanism.
:::

### Power model

```{r}
#| label: fit-power-model
power_model <- lm(
  fuel_use_l_per_100km ~ power_kw,
  data = vehicles
)

power_model
```

```{r}
#| label: plot-power-model
#| fig-cap: "The fitted power linear model."
#| fig-alt: "Scatterplot of engine power and combined fuel use with a fitted straight line."
#| out-width: "100%"
ggplot(vehicles, aes(x = power_kw, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Engine power (kW)",
    y = "Combined fuel use (L/100 km)",
    title = "A straight-line model for engine power and fuel use"
  ) +
  theme_minimal(base_size = 13)
```

The power model also captures a clear increasing trend. Its points have
slightly more variation around the line than the engine-size model.

## 3. Extract and interpret the coefficients

### Engine-size coefficients

::: {.callout-important title="Question"}
What is the fitted equation? What does the slope mean? Is the intercept practical?
:::

::: {.callout-note title="What to look for"}
Interpret slope and intercept with units, and check whether zero engine size is observed.
:::

```{r}
#| label: engine-coefficients
engine_coef <- coef(engine_model)
engine_coef
```

::: {#wk10-answer-engine-coefficients .callout-tip title="Worked interpretation"}
The fitted equation is approximately

$$
\widehat{\text{fuel use}} = 4.070 + 1.886\,\text{engine size}.
$$

A one-litre difference in engine size is associated with about **1.886
L/100 km** higher average fuel use. The intercept is the line’s fitted value
at zero litres. Zero is outside the observed engine-size range, so the
intercept has no useful direct vehicle interpretation.
:::

### Power coefficients

```{r}
#| label: power-coefficients
power_coef <- coef(power_model)
power_coef
```

The fitted equation is approximately

$$
\widehat{\text{fuel use}} = 4.705 + 0.0186\,\text{power}.
$$

A one-kilowatt difference in power is associated with about **0.0186 L/100
km** higher average fuel use. A 100 kW difference is therefore associated
with about **1.86 L/100 km** higher average fuel use. Zero kW is outside the
observed data, so the intercept should not be given a direct practical
interpretation.

::: {.callout-important}
The coefficients describe associations among these vehicle specifications.
Neither slope is a causal effect.
:::

## 4. Extract and inspect the residuals

### Engine-size residuals

::: {.callout-important title="Question"}
Which gaps are positive or negative, and why are they vertical?
:::

::: {.callout-note title="What to look for"}
`.fitted` is the line value and `.resid` is observed minus fitted, in L/100 km.
:::

```{r}
#| label: engine-augment
engine_results <- augment(engine_model)

engine_results |>
  select(engine_l, fuel_use_l_per_100km, .fitted, .resid) |>
  slice_head(n = 8)
```

`.fitted` is the point on the fitted line for an engine size. `.resid` is the
observed value minus the fitted value; positive residuals lie above the line
and negative residuals lie below it.

```{r}
#| label: engine-residual-gaps
#| fig-cap: "Observed fuel use, the engine-size fitted line, and a sample of residual gaps."
#| fig-alt: "Scatterplot with a fitted straight line and vertical segments connecting selected observations to the line."
#| out-width: "100%"
set.seed(297101)
engine_gap_data <- engine_results |>
  slice_sample(n = min(35, nrow(engine_results)))

ggplot(vehicles, aes(x = engine_l, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.35) +
  geom_line(
    data = engine_results |> arrange(engine_l),
    aes(y = .fitted),
    colour = "#2C7FB8",
    linewidth = 1
  ) +
  geom_segment(
    data = engine_gap_data,
    aes(xend = engine_l, yend = .fitted),
    alpha = 0.65,
    colour = "#D95F02"
  ) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Residuals are the vertical gaps from observations to the line"
  ) +
  theme_minimal(base_size = 13)
```

::: {#wk10-answer-engine-residual-gaps .callout-tip title="Worked interpretation"}
The vertical gaps are

$$
e_i = y_i - \widehat{y}_i.
$$

Positive gaps are observations above the line and negative gaps are below it;
all residuals use L/100 km units.
:::

::: {.callout-important title="Question"}
How much variation remains? What would a residual pattern tell you beyond the mean?
:::

::: {.callout-note title="What to look for"}
Compare response and residual variance. A near-zero mean is expected but is not a complete adequacy check.
:::

```{r}
#| label: engine-residual-plot
#| fig-cap: "Residuals from the engine-size model versus fitted fuel-use values."
#| fig-alt: "Engine-size model residuals plotted against fitted values with a horizontal zero line."
#| out-width: "100%"
ggplot(engine_results, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.55) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted fuel use (L/100 km)",
    y = "Residual (L/100 km)",
    title = "What the engine-size model leaves unexplained"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: engine-residual-summary
engine_results |>
  summarise(
    mean_residual = mean(.resid),
    response_variance = var(fuel_use_l_per_100km),
    residual_variance = var(.resid)
  )
```

::: {#wk10-answer-engine-residual-summary .callout-tip title="Worked interpretation"}
The mean residual is essentially zero. Response variance is about 7.136 and
engine residual variance about 1.341, so the model removes substantial variation
but leaves vehicle-specific differences. The near-zero mean does not establish
linearity or constant variance.
:::

### Power residuals

```{r}
#| label: power-augment
power_results <- augment(power_model)

power_results |>
  select(power_kw, fuel_use_l_per_100km, .fitted, .resid) |>
  slice_head(n = 8)
```

```{r}
#| label: power-residual-plot
#| fig-cap: "Residuals from the power model versus fitted fuel-use values."
#| fig-alt: "Power-model residuals plotted against fitted values with a horizontal zero line."
#| out-width: "100%"
ggplot(power_results, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.55) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted fuel use (L/100 km)",
    y = "Residual (L/100 km)",
    title = "What the power model leaves unexplained"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: power-residual-summary
power_results |>
  summarise(
    mean_residual = mean(.resid),
    response_variance = var(fuel_use_l_per_100km),
    residual_variance = var(.resid)
  )
```

The power-model mean residual is essentially zero and its residual variance
is approximately **1.557**. This is slightly larger than the engine model’s
1.341, agreeing with the slightly smaller correlation and greater visible
scatter. Formal LINE checks still belong to Week 12.

## 5. Independent workflow: gross vehicle mass

### A. Explore

```{r}
#| label: mass-explore
#| fig-cap: "Combined fuel use and gross vehicle mass."
#| fig-alt: "Scatterplot with gross vehicle mass in kilograms on the horizontal axis and combined fuel use in litres per 100 kilometres on the vertical axis, showing a moderately increasing relationship."
#| out-width: "100%"
ggplot(vehicles, aes(x = gross_mass_kg, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Gross vehicle mass (kg)",
    y = "Combined fuel use (L/100 km)",
    title = "Do heavier vehicles tend to use more fuel?"
  ) +
  theme_minimal(base_size = 13)

vehicles |>
  summarise(
    mass_correlation = cor(gross_mass_kg, fuel_use_l_per_100km)
  )
```

The mass correlation is approximately **0.576**, indicating a moderate
positive linear association that is weaker than the engine-size and power
relationships.

### B. Fit and inspect

```{r}
#| label: mass-fit
mass_model <- lm(
  fuel_use_l_per_100km ~ gross_mass_kg,
  data = vehicles
)

mass_model

ggplot(vehicles, aes(x = gross_mass_kg, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Gross vehicle mass (kg)",
    y = "Combined fuel use (L/100 km)",
    title = "A straight-line model for mass and fuel use"
  ) +
  theme_minimal(base_size = 13)
```

### C. Extract and interpret coefficients

```{r}
#| label: mass-coefficients
mass_coef <- coef(mass_model)
mass_coef
```

The fitted equation is approximately

$$
\widehat{\text{fuel use}} = 1.666 + 0.00329\,\text{gross mass}.
$$

A one-kilogram difference in gross mass is associated with about **0.00329
L/100 km** higher average fuel use. More usefully, a 1000 kg difference is
associated with about **3.29 L/100 km** higher average fuel use. Zero mass is
outside the data and gives the intercept no direct practical interpretation.

### D. Inspect residuals

```{r}
#| label: mass-residuals
mass_results <- augment(mass_model)

ggplot(mass_results, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.55) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted fuel use (L/100 km)",
    y = "Residual (L/100 km)",
    title = "What the mass model leaves unexplained"
  ) +
  theme_minimal(base_size = 13)

mass_results |>
  summarise(
    mean_residual = mean(.resid),
    response_variance = var(fuel_use_l_per_100km),
    residual_variance = var(.resid)
  )
```

The mass-model mean residual is essentially zero and its residual variance is
approximately **4.772**, substantially larger than those of the engine-size
and power models.

### E. Compare all three models

| Predictor | Correlation | Residual variance |
|:--|--:|--:|
| `engine_l` | 0.901 | 1.341 |
| `power_kw` | 0.884 | 1.557 |
| `gross_mass_kg` | 0.576 | 4.772 |

Engine size is the strongest of these three single predictors: it has the
largest correlation and smallest residual variance. Power performs almost as
well. Gross mass has a moderate relationship but leaves substantially more
variation unexplained. This comparison does not establish causation or test
whether observed differences would persist in other vehicle data.

::: {.callout-important}
## Scope of today’s conclusion

These models describe associations and how well each line summarises these
data. They do not establish causal effects or show that the same relationships
will hold for other vehicle years, fuel types or driving conditions.
:::

## Conclusion

Larger engines tend to have higher combined fuel use among the unique 2025
petrol passenger-vehicle specifications in these data. Engine size has the
strongest relationship of the three predictors considered, although power
performs almost as well and none of the fitted lines predicts every vehicle
exactly.

The analysis uses standard test values rather than real-world driving and is
restricted to one vehicle year and fuel type. Other vehicle features remain
unmeasured, the relationships are associations, and predictions beyond the
observed ranges would be extrapolations.

Next week we use point estimates, confidence intervals, hypothesis tests,
prediction intervals and variance explained to quantify the uncertainty in a
fitted relationship and in a prediction for a new observation.
