---
title: "Workshop Week 11: Multiple Regression and Factors"
subtitle: "Student workbook: adjusted coefficients and body type"
author:
  - name: "Nick Knowlton"
  - name: "297.101 Statistical Data Science"
date-meta: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

Last week we used one numerical vehicle specification at a time. Today we return
to fuel use and add a categorical predictor, `body_type`, represented in R as a
factor. This lets us compare body types while accounting for a numerical vehicle
characteristic at the same time. The key idea is that a comparison between body
types after accounting for another variable is not necessarily the same as
comparing their raw averages.

## Today's question

What changes when we add body type to a model of vehicle characteristics and fuel use?

## Learning objectives

By the end of this workshop you should be able to:

- fit a multiple linear regression (MLR) model with numerical and categorical predictors;
- interpret an adjusted coefficient using “holding the other variables constant”;
- identify the reference category and interpret coefficients for categorical predictors;
- interpret a factor with more than two levels;
- compare ordinary and adjusted $R^2$ across models fitted to the same response; and
- make predictions for specified combinations of predictors.

## How to use this workbook

Work through the sections in order. Run each supplied demonstration and inspect
its output before moving on. In each **Your turn** section, write your own code
in the blank chunk, then answer the question in words. Hints appear when a
function or idea is first introduced; later tasks expect you to reuse it. Open
this file from the `labs` folder so the supplied data file can be found.

## 1. Revisit a simple model

**What to do:** Start by refitting the simple engine-size and power models from
last week. Check that you can still interpret the slope, intercept and $R^2$.
The coefficient table also shows the standard error and p-value introduced in
this week's lecture. For now, use them to recognise that a fitted coefficient
is an estimate from a sample. We will do the formal confidence-interval and
hypothesis-testing interpretation next week.

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(broom)

vehicles <- read_csv("nz_vehicles_2025.csv", show_col_types = FALSE) |>
  mutate(
    # SALOON is the raw CSV value; display it as Sedan for us.
    station_wagon = factor(
      if_else(body_type == "STATION WAGON", "Station wagon", "Other"),
      levels = c("Station wagon", "Other")
    ),
    body_type = factor(
      body_type,
      levels = c("STATION WAGON", "HATCHBACK", "SALOON", "SPORTS CAR", "CONVERTIBLE"),
      labels = c("Station wagon", "Hatchback", "Sedan", "Sports car", "Convertible")
    )
  )

glimpse(vehicles)
count(vehicles, body_type)
```

### Demonstration: engine-size baseline

```{r}
engine_model <- vehicles |>
  lm(fuel_use_l_per_100km ~ engine_l, data = _)

# Or as above
engine_model <- lm(fuel_use_l_per_100km ~ engine_l, data = vehicles)

engine_model
tidy(engine_model)
```

Read the coefficient table and identify the engine-size slope. Then identify
its standard error and p-value. Which number is the fitted estimate, and which
number describes sampling uncertainty in that estimate? Use the plot to check
how the fitted line summarises the overall pattern. Is the intercept a realistic
vehicle comparison?

```{r}
ggplot(vehicles, aes(engine_l, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.45) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Simple engine-size model"
  ) +
  theme_minimal(base_size = 13)
```

The engine-size slope is the average change in L/100 km for a one-litre engine
difference. The intercept is the fitted value at zero litres. Since the data
contain positive engine sizes, the intercept is a mathematical *anchor* rather
than a realistic vehicle comparison.

### Your turn: power baseline

Refit the power model yourself and interpret the slope in useful units. Before
moving on, make sure you can explain what the coefficient tells us about the
overall association between power and fuel use.

Reuse the Week 10 workflow: fit the model, inspect its coefficients and plot the
fitted line. Report the power slope per 10 kW.

```{r}
# Your code here
```

```{r}
# Your code here
```

## 2. Add a two-level factor

**What to do:** Add the two-level `station_wagon` factor to the engine-size
model. Identify the reference category, then interpret the intercept, engine-size
coefficient and `station_wagonOther` coefficient.

Ask yourself: which coefficient is shared by both groups, and which coefficient
represents the difference between the groups?

`station_wagon` has two levels. Use the factor levels, coefficient names and
design matrix to determine which level R is using as the reference and what the
other factor coefficient compares.

### Demonstration: engine size plus Station wagon/Other

```{r}
binary_model <- lm(fuel_use_l_per_100km ~ engine_l + station_wagon, data = vehicles)

tidy(binary_model)

# Pull out the values used in fitting the model to see how the
# station_wagon variable is treated.
model.matrix(binary_model) |> head()
```

The intercept corresponds to Station wagons, while the engine-size slope is
shared by both groups. The `station_wagonOther` coefficient is the fitted
difference between Other vehicles and Station wagons when engine size is held
constant. Allowing the engine-size slope to differ between groups would require
an interaction term, which is outside the scope of this course.

### Your turn: power plus Station wagon/Other

Fit the same model using engine power instead of engine size. Compare the power
coefficient with the simple power model from Section 1. Has the coefficient
changed? What question does the adjusted coefficient answer that the simple
coefficient did not?

Also interpret the `station_wagonOther` coefficient at a common value of engine
power. Identify its p-value and state which zero population coefficient that
p-value is testing. Do not use the p-value by itself to decide whether the model
is useful. Show the first rows of the design matrix and explain its columns.

**Hint:** `model.matrix()` shows the numerical columns R creates for the
factor. The first factor level is the reference group.

```{r}
# Your code here
```

## 3. Use all five body types

**What to do:** Replace the two-level factor with the full five-level
`body_type` factor. Identify the reference category and count how many body-type
coefficients are fitted. Interpret each body-type coefficient as a comparison
with the reference category while holding engine size constant.

Pay particular attention to the direction and magnitude of the coefficients.
Do the adjusted comparisons make sense given what you know about the vehicles?

The two-level factor hides distinctions among Hatchbacks, Sedans, Sports cars
and Convertibles. Fit the five-level model, then use the coefficient names to
identify the reference category before interpreting the remaining levels.

### Demonstration: engine size plus body type

```{r}
body_model <- lm(
  fuel_use_l_per_100km ~ engine_l + body_type,
  data = vehicles
)

tidy(body_model)
```

There is one engine-size slope and four body-type differences relative to
Station wagons. For example, `body_typeSedan` is the fitted Sedan-minus-Station
wagon difference at the same engine size. It is not the raw difference between
the two group means. The additive model assumes that body-type lines have a
common engine-size slope; interactions are not part of this workshop. Looking
at the fitted coefficients, Sports cars are estimated to use slightly more fuel
than Station wagons at the same engine size, while Hatchbacks are estimated to
use less. After fitting a model, always sense-check the direction and magnitude
of the estimates against what you know about the problem. If something looks
surprising, work out why rather than simply assuming the model is wrong.

### Your turn: power plus body type

Repeat the model using engine power. Interpret the power coefficient per 10 kW
while holding body type constant. Interpret the Sedan coefficient relative to
Station wagon at the same power, and interpret one other body-type coefficient.
Compare these results with the engine-size model.

Do the conclusions about body type change when you hold engine power constant
instead of engine size? If they do, think about why these models are answering
different questions. Do the negative body-type estimates make sense at the same
engine power?

```{r}
# Your code here
```

## 4. Make comparable predictions

**What to do:** Use the fitted model to predict fuel use for every body type
while holding engine size at 2.0 L. Because engine size is fixed, differences
between these fitted values come directly from the body-type coefficients.

Identify which body type has the highest and lowest fitted mean fuel use.
Remember that these are model-based average predictions for specified
combinations of predictors, not guarantees for individual vehicles.

### Demonstration: every body type at 2.0 L

```{r}
new_body_vehicles <- tibble(
  engine_l = 2,
  body_type = factor(
    levels(vehicles$body_type),
    levels = levels(vehicles$body_type)
  )
)

new_body_vehicles |>
  mutate(predicted_fuel_use = predict(body_model, newdata = new_body_vehicles))
```

Holding engine size at 2.0 L makes each fitted value a comparable body-type
prediction.

What if we only wanted to predict a Sports car?

```{r sports_car}
sports_car_new <- tibble(
  engine_l = 2,
  body_type = "Sports car" # Match the factor level exactly.
)

sports_car_new |>
  mutate(predicted_fuel_use = predict(body_model, newdata = sports_car_new))
```

This illustrates how you can predict for a specified body type and engine-size
combination. As always, predictions far outside the combinations observed in
the data are extrapolations.

### Your turn: every body type at 150 kW

Repeat the prediction exercise with engine power fixed at 150 kW. Compare the
ordering of the predicted values with the 2.0 L predictions. Which body type has
the highest and lowest fitted mean fuel use in each comparison? Explain why the
two rankings can differ when a different numerical characteristic is held
constant.

```{r}
# Your code here
```

## 5. Independent workflow: gross vehicle mass and body type

**Your turn: complete the full workflow using gross vehicle mass.**

1. Fit a simple model of fuel use and `gross_mass_kg`, then add
   `station_wagon`. Replace that factor with the five-level `body_type` and fit
   the third model. Station wagon is the reference level.
2. For each model, inspect the coefficients. Interpret the adjusted slopes and
   body-type differences, including the Sedan-minus-Station-wagon comparison.
   Remember that the body-type lines share a slope.
3. Compare the simple and adjusted mass slopes per 1000 kg. Explain what each
   comparison holds constant and whether the slopes suggest mass and body type
   carry overlapping information.
4. Predict fuel use for each body type at 2000 kg. Identify the largest fitted
   mean and explain what the adjustment adds.
5. Compare $R^2$ and adjusted $R^2$ across all three models. State what happens
   as terms are added and which model's adjusted $R^2$ is largest.

### Fit and interpret the mass models

Now model both the binary and five-level body-type factors with
`gross_mass_kg`. Interpret the mass slope per 1000 kg and discuss whether the
simple-model intercept is meaningful. After adjusting for gross mass, compare
the fitted Sedan and Station wagon values. Do these results reveal any new
relationships?

```{r}
# Your code here: fit and print mass_model,
# mass_binary_model, and mass_body_model; inspect their coefficients.
```

### Compare the slopes and predictions

Compare the slopes from the simple linear model and the factor model. Report
the effect for 1000 kg to make its size easier to interpret. What do you see?
Explain why the adjusted slope answers a different question rather than
necessarily being more accurate, and whether the change suggests that body type
and mass carry overlapping information. Then predict fuel use for every body
type at 2000 kg, identify the largest fitted value, and explain what the
adjustment adds.

```{r}
# Your code here: compare the simple and adjusted mass slopes.
```

```{r}
# Your code here: predict for every body type at gross_mass_kg = 2000.
```

### Compare model fit

Thinking back to last week, what are the $R^2$ and adjusted $R^2$ values for
these models, and which explains the most variation? Use `glance()` to inspect
both measures. What happens to ordinary $R^2$ when terms are added, and why
does adjusted $R^2$ provide a useful penalty?

```{r comparing_models}
# Your code here: compare glance(mass_model),
# glance(mass_binary_model), and glance(mass_body_model).
```

## Optional extra: fitting a model to the Week 9 coffee ratings

In Homework 9, you examined ratings for five coffee brands from 1 to 9 and
adjusted for the nuisance effects of day and time. This example uses a
[balanced 50-rating subset](https://knowlton.co.nz/297101/labs/coffee_tasting_balanced.csv).
Each brand has two ratings on every day and at every time. The Week 9
hand-adjusted brand means have been calculated for you, so use them to compare
with the linear model. Your personalised homework data may give different
results. See the
[original Week 9 homework instructions](https://r-resources.massey.ac.nz/297101/CoffeeTasting.html)
for details.

| Coffee | Week 9 hand-adjusted mean |
|:--|--:|
| African Cream | 6.10 |
| Brew Maestro | 5.70 |
| Cafe Special | 7.30 |
| Delicio | 5.40 |
| Ethiopian Highlands | 8.30 |

Fit an additive model of rating using Time, Day and Coffee. The coffee
coefficients compare each brand with the reference brand after accounting for
day and time.

```{r coffee-example-model}
Coffee <- read.csv("coffee_tasting_balanced.csv")

Coffee <- Coffee |>
  mutate(
    Coffee = factor(Coffee),
    Time = factor(Time),
    Day = factor(Day)
  )

count(Coffee, Coffee)

coffee_model <- lm(Rating ~ Time + Day + Coffee, data = Coffee)
tidy(coffee_model)
```

Average the fitted ratings within each brand to get its model-adjusted mean.
Because the subset is balanced across day and time for every brand, these
means can be compared directly with the supplied Week 9 hand-adjusted means.

```{r coffee-example-adjusted-means}
Coffee |>
  mutate(predicted_rating = predict(coffee_model)) |>
  group_by(Coffee) |>
  summarise(
    model_adjusted_mean = mean(predicted_rating),
    .groups = "drop"
  ) |>
  arrange(desc(model_adjusted_mean)) |>
  mutate(model_adjusted_mean = round(model_adjusted_mean, 2))
```

Do the model-adjusted means match the supplied hand-adjusted means to two
decimal places? Which brand has the highest adjusted mean rating? Interpret
one coffee coefficient and its standard error relative to the reference brand.

## Conclusion

Summarise how adding body type changed the interpretation of a numerical slope,
how the Station wagon reference category works, and what you learned from the
gross-mass workflow. Explain why model complexity can change fitted slopes and
why ordinary $R^2$ and adjusted $R^2$ respond differently to added terms.
Explain why the “best” body type can depend on the numerical specification used
for the comparison. Distinguish these observational associations and conditional
predictions from causal effects.

## Looking ahead

Next week we will study uncertainty in estimates and predictions and interpret
diagnostic plots for the assumptions behind linear-model inference.

[View the worked solutions online](https://knowlton.co.nz/297101/labs/Workshop_Week11_Testing_guide.html).
