---
title: "Workshop Week 10: Vehicle Fuel Efficiency"
subtitle: "Exploring and fitting a first linear model"
author: "297.101 Statistical Data Science"
date: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    code-block-bg: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

::: {.callout-note}
## Today’s question

How does engine size relate to the fuel use of a petrol passenger vehicle?

For each new topic, first work through the **Demonstration** using engine
size. Then complete **Your turn** using engine power. At the end, you will
apply the whole workflow independently using vehicle mass.
:::

## Learning objectives

By the end of this workshop you should be able to:

- describe the direction, form and strength of a relationship from a plot;
- fit and inspect a simple linear model with `lm()`;
- extract and interpret the fitted coefficients;
- obtain and examine fitted values and residuals; and
- apply the complete workflow to a new explanatory variable.

## 1. Meet and explore the data (15 minutes)

```{r}
#| label: setup
library(tidyverse)
library(broom)

vehicles <- read_csv("nz_vehicles_2025.csv", show_col_types = FALSE)
```

Each row is one **unique vehicle specification** found in the New Zealand
vehicle data. The data were prepared from Waka Kotahi’s vehicle fleet data as
at **31 July 2026**. They contain model-year 2025 petrol passenger cars and
vans with complete measurements; repeated specifications have been removed.

The fuel-use values come from a standard test, not from the driving of a
particular person. An association in these data does not show that changing
one vehicle feature would cause another feature to change.

Waka Kotahi publishes the source under the [CC-BY 4.0 International
licence](https://creativecommons.org/licenses/by/4.0/). Read more in the
[vehicle-fleet documentation](https://nzta.govt.nz/resources/new-zealand-motor-vehicle-register-statistics/new-zealand-vehicle-fleet-open-data-sets).

```{r}
#| label: glimpse
glimpse(vehicles)
```

| Column | Meaning | Units |
|:--|:--|:--|
| `make`, `model`, `body_type` | vehicle description | — |
| `engine_l` | engine displacement | litres |
| `power_kw` | engine power rating | kilowatts |
| `gross_mass_kg` | maximum permitted laden mass | kilograms |
| `fuel_use_l_per_100km` | combined-cycle fuel use | litres per 100 km |

### Demonstration: engine size

Before fitting a line, start with a scatterplot and a LOESS smooth. The smooth
helps us see the general shape without requiring it to be a straight line.

```{r}
#| label: engine-loess
#| fig-cap: "Combined fuel use and engine size for 2025 petrol passenger vehicles."
#| fig-alt: "Scatterplot with engine size in litres on the horizontal axis and combined fuel use in litres per 100 kilometres on the vertical axis, showing a generally increasing relationship."
#| out-width: "100%"
ggplot(vehicles, aes(x = engine_l, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Do larger engines tend to use more fuel?"
  ) +
  theme_minimal(base_size = 13)
```

The correlation coefficient measures the direction and strength of a
*linear* relationship. It has no units and ranges from -1 to 1.

```{r}
#| label: engine-correlation
vehicles |>
  summarise(
    engine_correlation = cor(engine_l, fuel_use_l_per_100km)
  )
```

Describe the direction, form and strength shown by the plot. Then explain what
the sign and size of the correlation add to that description.

### Your turn: engine power

Repeat the exploration with `power_kw` as the explanatory variable. Create a
scatterplot with a LOESS smooth, use labels with units, and calculate the
correlation between power and fuel use.

```{r}
#| label: power-loess
# Use the engine-size plot as your template.
# Your code here
```

```{r}
#| label: power-correlation
# Use the engine-size correlation as your template.
# Your code here
```

Compare the engine-size and power plots. Which appears more nearly linear?
Which has the larger correlation? Remember that correlation is not a slope.

## 2. Fit and inspect the models (15 minutes)

A simple linear model describes the average response as a straight-line
function of one explanatory variable:

$$
\mathsf{Mean}(y) = a + bx.
$$

### Demonstration: engine-size model

Fit the model, save it, and then print the saved model object. Keeping the
model in an object lets us return to it throughout the workshop.

```{r}
#| label: fit-engine-model
engine_model <- lm(
  fuel_use_l_per_100km ~ engine_l,
  data = vehicles
)

engine_model
```

The next plot changes the flexible LOESS smooth to the straight line fitted by
`lm()`.

```{r}
#| label: plot-engine-model
#| fig-cap: "The fitted engine-size linear model."
#| fig-alt: "Scatterplot of engine size and combined fuel use with a fitted straight line."
#| out-width: "100%"
ggplot(vehicles, aes(x = engine_l, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "A straight-line model for engine size and fuel use"
  ) +
  theme_minimal(base_size = 13)
```

### Your turn: power model

Fit `fuel_use_l_per_100km ~ power_kw`, save it as `power_model`, print the
model object, and make the corresponding scatterplot with a fitted straight
line.

```{r}
#| label: fit-power-model
# Use fit-engine-model as your template.
# Your code here
```

```{r}
#| label: plot-power-model
# Use plot-engine-model as your template.
# Your code here
```

Look at both fitted lines. What does each line capture, and what variation is
still visible around it?

## 3. Extract and interpret the coefficients (15 minutes)

### Demonstration: engine-size coefficients

`coef()` extracts the intercept and slope from the saved model.

```{r}
#| label: engine-coefficients
engine_coef <- coef(engine_model)
engine_coef
```

Use these values to write the fitted equation:

$$
\widehat{\text{fuel use}} = a + b\,\text{engine size}.
$$

Then answer:

1. What does the slope mean in **L/100 km per litre of engine size**?
2. What does the intercept mean mathematically?
3. Is an engine size of zero within the observed data? How does that affect
   the practical interpretation of the intercept?

### Your turn: power coefficients

Extract the coefficients from `power_model`, save them as `power_coef`, and
write the fitted equation using `power_kw`.

```{r}
#| label: power-coefficients
# Use engine-coefficients as your template.
# Your code here
```

Interpret the slope in **L/100 km per kilowatt**. Also explain the expected
difference associated with 100 kW so that the magnitude is easier to
communicate. Is zero kW within the observed data?

::: {.callout-important}
The coefficients describe associations among these vehicle specifications.
Do not turn either slope into a causal claim.
:::

## 4. Extract and inspect the residuals (25 minutes)

For each vehicle specification, the fitted value is the value on the fitted
line. The residual is the observed value minus the fitted value:

$$
e_i = y_i - \hat{y}_i.
$$

### Demonstration: engine-size residuals

`augment()` returns the original variables together with useful model columns,
including `.fitted` and `.resid`.

```{r}
#| label: engine-augment
engine_results <- augment(engine_model)

engine_results |>
  select(engine_l, fuel_use_l_per_100km, .fitted, .resid) |>
  slice_head(n = 8)
```

The orange segments below show a sample of residuals as vertical gaps from the
observed values to the fitted line.

```{r}
#| label: engine-residual-gaps
#| fig-cap: "Observed fuel use, the engine-size fitted line, and a sample of residual gaps."
#| fig-alt: "Scatterplot with a fitted straight line and vertical segments connecting selected observations to the line."
#| out-width: "100%"
set.seed(297101)
engine_gap_data <- engine_results |>
  slice_sample(n = min(35, nrow(engine_results)))

ggplot(vehicles, aes(x = engine_l, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.35) +
  geom_line(
    data = engine_results |> arrange(engine_l),
    aes(y = .fitted),
    colour = "#2C7FB8",
    linewidth = 1
  ) +
  geom_segment(
    data = engine_gap_data,
    aes(xend = engine_l, yend = .fitted),
    alpha = 0.65,
    colour = "#D95F02"
  ) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Residuals are the vertical gaps from observations to the line"
  ) +
  theme_minimal(base_size = 13)
```

Plot all residuals against their fitted values, then summarise the mean
residual and compare the original response variance with the residual
variance.

```{r}
#| label: engine-residual-plot
#| fig-cap: "Residuals from the engine-size model versus fitted fuel-use values."
#| fig-alt: "Engine-size model residuals plotted against fitted values with a horizontal zero line."
#| out-width: "100%"
ggplot(engine_results, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.55) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted fuel use (L/100 km)",
    y = "Residual (L/100 km)",
    title = "What the engine-size model leaves unexplained"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: engine-residual-summary
engine_results |>
  summarise(
    mean_residual = mean(.resid),
    response_variance = var(fuel_use_l_per_100km),
    residual_variance = var(.resid)
  )
```

### Your turn: power residuals

Repeat the same core checks with `power_model`:

1. Create `power_results` with `augment()` and inspect `.fitted` and `.resid`.
2. Plot the power-model residuals against fitted values.
3. Calculate the mean residual, response variance and residual variance.
4. Compare the power residual variance with the engine residual variance.

```{r}
#| label: power-augment
# Use engine-augment as your template.
# Your code here
```

```{r}
#| label: power-residual-plot
# Use engine-residual-plot as your template.
# Your code here
```

```{r}
#| label: power-residual-summary
# Use engine-residual-summary as your template.
# Your code here
```

Describe what the power model leaves unexplained. This is an early look at
model leftovers; formal LINE checks belong to Week 12.

## 5. Independent workflow: gross vehicle mass (25 minutes)

Now complete the entire workflow independently using `gross_mass_kg` as the
explanatory variable. Reuse only operations already demonstrated above.

### A. Explore

- Plot fuel use against gross mass with a LOESS smooth.
- Calculate and interpret the correlation.

```{r}
#| label: mass-explore
# Your code here
```

### B. Fit and inspect

- Fit `fuel_use_l_per_100km ~ gross_mass_kg` as `mass_model`.
- Print the model and plot its fitted straight line.

```{r}
#| label: mass-fit
# Your code here
```

### C. Extract and interpret coefficients

- Save `coef(mass_model)` as `mass_coef`.
- Write the fitted equation.
- Interpret the slope in **L/100 km per kilogram** and, more usefully, per
  **1000 kg**.

```{r}
#| label: mass-coefficients
# Your code here
```

### D. Inspect residuals

- Create `mass_results` with `augment()`.
- Plot residuals against fitted values.
- Calculate the mean residual, response variance and residual variance.

```{r}
#| label: mass-residuals
# Your code here
```

### E. Compare all three models

Using your plots, correlations and residual variances, compare engine size,
power and gross mass as single predictors of fuel use. Name the strongest
predictor in these data, give evidence, and state what remains unexplained.

::: {.callout-important}
## Scope of today’s conclusion

You can describe an association and how well a line summarises these data.
You cannot yet quantify uncertainty in the slope, test a null hypothesis, or
construct an interval for a future vehicle. Those are Week 11 questions.
:::

## 6. Exit ticket and bridge to Week 11 (10 minutes)

Write three short responses:

1. **Supported conclusion:** What does the engine-size model suggest about
   fuel use for these vehicles?
2. **Limitation:** Name one feature of the data or model that limits the
   conclusion.
3. **Next question:** What would you want to know about the fitted slope or a
   prediction that this workshop did not answer?

Next week we will use model output to study point estimates, confidence
intervals, hypothesis tests, prediction intervals and variance explained.
