---
title: "Workshop Week 11: Multiple Regression and Factors"
subtitle: "Answer guide: adjusted coefficients and body type"
author:
  - name: "Nick Knowlton"
  - name: "297.101 Statistical Data Science"
date-meta: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

This guide follows Workshop_Week11_Testing.qmd.

Last week we used one numerical vehicle specification at a time. Today we return
to fuel use and add a categorical predictor, `body_type`, represented in R as a
factor. This lets us compare body types while accounting for a numerical vehicle
characteristic at the same time. The key idea is that a comparison between body
types after accounting for another variable is not necessarily the same as
comparing their raw averages.

## Today's question

What changes when we add body type to a model of vehicle characteristics and fuel use?

## Learning objectives

By the end of this workshop you should be able to:

- fit a multiple linear regression (MLR) model with numerical and categorical predictors;
- interpret an adjusted coefficient using “holding the other variables constant”;
- identify the reference category and interpret coefficients for categorical predictors;
- interpret a factor with more than two levels; and
- make predictions for specified combinations of predictors.

## 1. Revisit a simple model

**What to do:** Start by refitting the simple engine-size and power models from
last week. Check that you can still interpret the slope, intercept, standard
error, p-value and $R^2$. These give us the baseline models that we will compare
with once body type is added.

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(broom)

vehicles <- read_csv("nz_vehicles_2025.csv", show_col_types = FALSE) |>
  mutate(
    # SALOON is the raw CSV value; display it as Sedan for us.
    station_wagon = factor(
      if_else(body_type == "STATION WAGON", "Station wagon", "Other"),
      levels = c("Station wagon", "Other")
    ),
    body_type = factor(
      body_type,
      levels = c("STATION WAGON", "HATCHBACK", "SALOON", "SPORTS CAR", "CONVERTIBLE"),
      labels = c("Station wagon", "Hatchback", "Sedan", "Sports car", "Convertible")
    )
  )

glimpse(vehicles)
count(vehicles, body_type)
```

### Demonstration: engine-size baseline

```{r}
engine_model <- vehicles |>
  lm(fuel_use_l_per_100km ~ engine_l, data = _)

# or
engine_model <- lm(fuel_use_l_per_100km ~ engine_l, data = vehicles)


engine_model
tidy(engine_model)
```

Read the coefficient table and identify the engine-size slope, its standard
error and p-value. Then use the plot to check how the fitted line summarises the
overall pattern. Is the intercept a realistic vehicle comparison?

```{r}
ggplot(vehicles, aes(engine_l, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.45) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Simple engine-size model"
  ) +
  theme_minimal(base_size = 13)
```

The engine-size slope is the average change in L/100 km for a one-litre engine
difference. The intercept is the fitted value at zero litres. Since the data
contain positive engine sizes, the intercept is a *mathematical anchor* rather
than a realistic vehicle comparison.

### Your turn: power baseline

Refit the power model yourself and interpret the slope in useful units. Before
moving on, make sure you can explain what the coefficient tells us about the
overall association between power and fuel use.

Reuse the Week 10 workflow: fit the model, inspect its coefficients and plot the
fitted line. Report the power slope per 10 kW.

```{r}
power_model <- lm(
  fuel_use_l_per_100km ~ power_kw,
  data = vehicles
)

power_model
tidy(power_model)
```

```{r}
ggplot(vehicles, aes(power_kw, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.45) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Engine power (kW)",
    y = "Combined fuel use (L/100 km)",
    title = "Simple power model"
  ) +
  theme_minimal(base_size = 13)
```

The power slope is interpreted per kW; multiplying it by 10 gives the fitted
change for a 10 kW difference. Zero kW is outside the observed vehicle range,
so the intercept is not a practical zero-power prediction.

The fitted equation is approximately

$$
\widehat{\text{fuel use}} = 4.705 + 0.0186(\text{power}) .
$$

Thus a 10 kW increase is associated with about 0.186 L/100 km higher average
fuel use. The intercept is the mathematical line value at zero power, not a
realistic vehicle comparison because zero power is outside the observed range.

## 2. Add a two-level factor

**What to do:** Add the two-level `station_wagon` factor to the engine-size
model. Identify the reference category, then interpret the intercept, engine-size
coefficient and `station_wagonOther` coefficient.

Ask yourself: which coefficient is shared by both groups, and which coefficient
represents the difference between the groups?

### Demonstration: engine size plus Station wagon/Other

```{r}
binary_model <- lm(
  fuel_use_l_per_100km ~ engine_l + station_wagon,
  data = vehicles
)

tidy(binary_model)

# Pull out the values used in fitting the model to see how the
# station_wagon variable is treated.
model.matrix(binary_model) |> head()
```

The first factor level, Station wagon, is the reference. The intercept
corresponds to Station wagons, while the engine-size slope is shared by both
groups. `station_wagonOther` is 0 for Station wagons and 1 for Other vehicles,
so its coefficient is the fitted `Other-minus-Station wagon` difference at the
same engine size. Allowing the engine-size slope to differ between groups would
require an interaction term, which is outside the scope of this course.

The columns in `tidy()` have the same interpretation as last week: `estimate`
gives the fitted coefficient, `std.error` describes its uncertainty, and the
p-value tests a null hypothesis that the corresponding population coefficient
is zero. The important difference is that these are now adjusted coefficients:
each one is interpreted conditional on the other variables in the model.

### Your turn: power plus Station wagon/Other

Fit the same model using engine power instead of engine size. Compare the power
coefficient with the simple power model from Section 1. Has the coefficient
changed? What question does the adjusted coefficient answer that the simple
coefficient did not?

Also interpret the `station_wagonOther` coefficient and its p-value at a common
value of engine power. Show the first rows of the design matrix and explain its
columns.

**Hint:** `model.matrix()` shows the numerical columns R creates for the
factor. The first factor level is the reference group.

```{r}
power_binary_model <- lm(
  fuel_use_l_per_100km ~ power_kw + station_wagon,
  data = vehicles
)

tidy(power_binary_model)

model.matrix(power_binary_model) |> head()
```

The simple power slope is about 0.0186 L/100 km per kW, while the adjusted
power slope is about 0.0192 L/100 km per kW, or 0.192 L/100 km per 10 kW. The
coefficient changes because the adjusted model compares vehicles at the same
Station wagon/Other group rather than averaging across all vehicles. The
`station_wagonOther` estimate is about -0.577 L/100 km: at the same power, the
model predicts Other vehicles to use about 0.577 L/100 km less than Station
wagons. Its p-value is about 1.76e-6, providing strong evidence against a zero
adjusted difference under the model. In the design matrix, the intercept column
is 1, `power_kw` contains the numerical predictor, and
`station_wagonOther` is 0 for Station wagons and 1 for Other vehicles.

Without body type, the power slope summarises the overall association across
all vehicles. Once body type is included, the power coefficient compares
vehicles at the same body type, under the assumption of a common slope.

## 3. Use all five body types

**What to do:** Replace the two-level factor with the full five-level
`body_type` factor. Identify the reference category and count how many body-type
coefficients are fitted. Interpret each body-type coefficient as a comparison
with the reference category while holding engine size constant.

Pay particular attention to the direction and magnitude of the coefficients.
Do the adjusted comparisons make sense given what you know about the vehicles?

### Demonstration: engine size plus body type

```{r}
body_model <- lm(
  fuel_use_l_per_100km ~ engine_l + body_type,
  data = vehicles
)

tidy(body_model)
```

There is one engine-size slope and four body-type coefficients. Each coefficient
is a comparison with Station wagon at the same engine size. In particular,
`body_typeSedan` is the fitted `Sedan-minus-Station wagon` difference, not the
difference between their unadjusted mean fuel uses. Looking at the fitted
coefficients, Sports cars are estimated to use slightly more fuel than Station
wagons at the same engine size, while Hatchbacks are estimated to use less.
For `body_typeHatchback`, the coefficient is the estimated difference from
Station wagons at the same engine size. Its p-value tests whether that adjusted
population difference is zero. Does this make sense? After fitting a model,
always sense-check the direction and magnitude of the estimates against what
you know about the problem. If something looks surprising, work out why rather
than simply assuming the model is wrong.

### Your turn: power plus body type

Repeat the model using engine power. Interpret the power coefficient per 10 kW
while holding body type constant. Interpret the Sedan coefficient relative to
Station wagon at the same power, and interpret one other body-type coefficient.
Compare these results with the engine-size model.

Do the conclusions about body type change when you hold engine power constant
instead of engine size? If they do, think about why these models are answering
different questions. Do the negative body-type estimates make sense at the same
engine power?

```{r}
power_body_model <- lm(
  fuel_use_l_per_100km ~ power_kw + body_type,
  data = vehicles
)

tidy(power_body_model)
```

The power slope is about 0.0183 L/100 km per kW, or about 0.183 L/100 km per
10 kW, among vehicles of the same body type. `body_typeSedan` is about -0.705
L/100 km: at equal power, the model predicts Sedans to use about 0.705 L/100 km
less than Station wagons. For another example, `body_typeHatchback` is about
-0.907 L/100 km at equal power. All four fitted body-type coefficients are
negative, meaning that at the same engine power the model predicts lower fuel
use than for Station wagons. This is a conditional comparison, so it need not
match the raw group means.

## 4. Make comparable predictions

**What to do:** Use the fitted model to predict fuel use for every body type
while holding engine size at 2.0 L. Because engine size is fixed, differences
between these fitted values come directly from the body-type coefficients.

Identify which body type has the highest and lowest fitted mean fuel use.
Remember that these are model-based average predictions for specified
combinations of predictors, not guarantees for individual vehicles.

### Demonstration: every body type at 2.0 L

```{r}
new_body_vehicles <- tibble(
  engine_l = 2,
  body_type = factor(
    levels(vehicles$body_type),
    levels = levels(vehicles$body_type)
  )
)

new_body_vehicles |>
  mutate(predicted_fuel_use = predict(body_model, newdata = new_body_vehicles))
```

Holding engine size at 2.0 L makes each fitted value a comparable body-type
prediction. The largest value, Sports car, is the body type with the largest
adjusted prediction in the output. These are average predictions for the
specified combinations, not causal effects or guarantees for individual cars.

What if we only wanted to predict a Sports car?
```{r sports_car}

sports_car_new <- tibble(
  engine_l = 2,
  body_type = "Sports car" # Match the factor level exactly.
)

sports_car_new |>
  mutate(predicted_fuel_use = predict(body_model, newdata = sports_car_new))

```
This illustrates how you can predict for a specified body type and engine-size
combination. As always, predictions far outside the combinations observed in
the data are extrapolations.

### Your turn: every body type at 150 kW

Repeat the prediction exercise with engine power fixed at 150 kW. Compare the
ordering of the predicted values with the 2.0 L predictions. Which body type has
the highest and lowest fitted mean fuel use in each comparison? Explain why the
two rankings can differ when a different numerical characteristic is held
constant.

```{r}
new_power_vehicles <- tibble(
  power_kw = 150,
  body_type = factor(
    levels(vehicles$body_type),
    levels = levels(vehicles$body_type)
  )
)

new_power_vehicles |>
  mutate(predicted_fuel_use = predict(power_body_model, newdata = new_power_vehicles))
```

As we saw with the coefficients, the `Station wagon` has the largest fitted
fuel use at 150 kW in this model (about 7.73 L/100 km, compared with about
6.82 for Hatchbacks, 7.03 for Sedans, 7.65 for Sports cars and 7.38 for
Convertibles). This does not make the model incorrect. The engine-size model
and power model hold different numerical specifications constant, so they are
answering different conditional questions about fuel use.

The factor-level order matters when fitting the model because it determines the
reference category. The order of rows in `newdata` does not change the
predictions.

## 5. Independent workflow: gross vehicle mass and body type

**Your turn: complete the full workflow using gross vehicle mass.**

1. Fit a simple model of fuel use and `gross_mass_kg`, then add
   `station_wagon`. Replace that factor with the five-level `body_type` and fit
   the third model. Station wagon is the reference level.
2. For each model, inspect the coefficients. Interpret the adjusted slopes and
   body-type differences, including the Sedan-minus-Station-wagon comparison.
   Remember that the body-type lines share a slope.
3. Compare the simple and adjusted mass slopes per 1000 kg. Explain what each
   comparison holds constant and whether the slopes suggest mass and body type
   carry overlapping information.
4. Predict fuel use for each body type at 2000 kg. Identify the largest fitted
   mean and explain what the adjustment adds.
5. Compare $R^2$ and adjusted $R^2$ across all three models. State what happens
   as terms are added and which model's adjusted $R^2$ is largest.

### Fit and interpret the mass models

Now model both the binary and five-level body-type factors with
`gross_mass_kg`. Interpret the mass slope per 1000 kg and discuss whether the
simple-model intercept is meaningful. After adjusting for gross mass, compare
the fitted Sedan and Station wagon values. Do these results reveal any new
relationships?

```{r}
mass_model <- lm(
  fuel_use_l_per_100km ~ gross_mass_kg,
  data = vehicles
)

mass_binary_model <- lm(
  fuel_use_l_per_100km ~ gross_mass_kg + station_wagon,
  data = vehicles
)

mass_body_model <- lm(
  fuel_use_l_per_100km ~ gross_mass_kg + body_type,
  data = vehicles
)

tidy(mass_model)
tidy(mass_binary_model)
tidy(mass_body_model)
```

After adjusting for gross mass, the estimated Sedan-Station wagon difference is
small relative to its standard error, and the data provide little evidence
against a zero adjusted difference. In the full body-type model, the Sedan
coefficient is about 0.248 L/100 km, so at 2000 kg the fitted Sedan value is
about 7.43 L/100 km versus about 7.18 L/100 km for a Station wagon. The simple
mass-model intercept is about 1.67 L/100 km at zero kilograms; because zero is
outside the observed vehicle range, it is a mathematical anchor rather than a
practical prediction.

### Compare the slopes and predictions

Compare the slopes from the simple linear model and the factor model. Report
the effect for 1000 kg to make its size easier to interpret. What do you see?
Explain why the adjusted slope answers a different question rather than
necessarily being more accurate, and whether the change suggests that body type
and mass carry overlapping information. Then predict fuel use for every body
type at 2000 kg, identify the largest fitted value, and explain what the
adjustment adds.
```{r}
mass_slope_comparison <- tibble(
  model = c("Simple mass", "Mass adjusted for body type"),
  mass_slope_per_kg = c(
    coef(mass_model)[["gross_mass_kg"]],
    coef(mass_body_model)[["gross_mass_kg"]]
  ),
  mass_slope_per_1000kg = 1000 * mass_slope_per_kg
)
mass_slope_comparison
```

The mass slope changes from 3.29 to 4.17 L/100 km per 1000 kg after
accounting for body type. These coefficients answer different questions: the
simple model describes the overall association between mass and fuel use, while
the adjusted model describes the association between mass and fuel use among
vehicles of the same body type. The difference tells us that body type and mass
are carrying some overlapping information.

```{r}
new_mass_vehicles <- tibble(
  gross_mass_kg = 2000,
  body_type = factor(
    levels(vehicles$body_type),
    levels = levels(vehicles$body_type)
  )
)

mass_predictions <- new_mass_vehicles |>
  mutate(predicted_fuel_use = predict(mass_body_model, newdata = new_mass_vehicles))
mass_predictions
```

The mass slope in the simple model is the average association without body-type
adjustment. In `mass_body_model`, it is the expected difference for 1000 kg
among vehicles with the same body type. `body_typeSedan` is the fitted Sedan
difference from Station wagon at 2000 kg (and at every other mass under this
additive model). The largest prediction is for Sports cars, at about 11.5
L/100 km. These are observational associations and conditional model
predictions; they do not show that changing mass or body type would cause fuel
use to change.


### Compare model fit

Thinking back to last week, what are the $R^2$ and adjusted $R^2$ values for
these models, and which explains the most variation? Use `glance()` to inspect
both measures. What happens to ordinary $R^2$ when terms are added, and why
does adjusted $R^2$ provide a useful penalty?

```{r comparing_models}

bind_rows(
  Mass = glance(mass_model),
  `Mass + Station wagon` = glance(mass_binary_model),
  `Mass + Body type` = glance(mass_body_model),
  .id = "model"
) |>
  select(model, r.squared, adj.r.squared)
```
The output is approximately: simple mass, R-squared 0.331 and adjusted
R-squared 0.330; mass plus Station wagon/Other, 0.476 and 0.474; and mass plus
all body types, 0.691 and 0.688. Ordinary R-squared increases when terms are
added because the fitted model has more freedom. Adjusted R-squared applies a
penalty for extra terms, so it increases only when the added terms improve the
fit enough to justify their complexity. The full body-type model explains the
most in-sample variation here.


## Conclusion

Summarise how adding body type changed the interpretation of a numerical slope,
how the Station wagon reference category works, and what you learned from the
gross-mass workflow. Explain why model complexity can change fitted slopes and
why ordinary $R^2$ and adjusted $R^2$ respond differently to added terms.
Explain why the "best" body type can depend on the numerical specification used
for the comparison. Distinguish these observational associations and
conditional predictions from causal effects.

The factor coefficients are always comparisons with the *chosen reference
category* at a common numerical value. Adding body type changes a numerical
coefficient because the model no longer attributes all between-group variation
to the numerical predictor. Ordinary $R^2$ cannot decrease when terms are added
to the same in-sample OLS fit, although adjusted $R^2$ can decrease and
out-of-sample performance can get worse. We also saw that the "best" body type
depends on the question asked of the data.

## Looking ahead

Next week we will study diagnostic plots and the assumptions behind linear-model
inference, building on the coefficient uncertainty introduced this week.
