---
title: "Workshop Week 10: Vehicle Fuel Efficiency"
subtitle: "Answer guide — exploring and fitting a first linear model"
author: "297.101 Statistical Data Science"
date: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    code-block-bg: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

This guide follows the student workshop in
[`Workshop_Week10_Vehicle_Regression.qmd`](Workshop_Week10_Vehicle_Regression.qmd).
It mirrors each Demonstration and Your turn exercise and then completes the
independent gross-mass workflow.

::: {.callout-note}
## Today’s question

How does engine size relate to the fuel use of a petrol passenger vehicle?

Each new operation is demonstrated with engine size, repeated with engine
power, and finally applied independently using gross vehicle mass.
:::

## Learning objectives

Students should be able to describe a quantitative relationship, fit and
inspect a simple linear model, extract and interpret its coefficients, obtain
and examine fitted values and residuals, and repeat the workflow with a new
explanatory variable.

## 1. Meet and explore the data (15 minutes)

```{r}
#| label: setup
library(tidyverse)
library(broom)

vehicles <- read_csv("nz_vehicles_2025.csv", show_col_types = FALSE)
```

```{r}
#| label: glimpse
glimpse(vehicles)
```

There are **467** unique vehicle specifications. The data were prepared from
Waka Kotahi’s vehicle fleet data as at **31 July 2026** and contain model-year
2025 petrol passenger cars and vans with complete measurements. Standard test
values are not the observed fuel use of individual drivers.

Waka Kotahi publishes the source under the [CC-BY 4.0 International
licence](https://creativecommons.org/licenses/by/4.0/). Source details and
caveats are in the [vehicle-fleet documentation](https://nzta.govt.nz/resources/new-zealand-motor-vehicle-register-statistics/new-zealand-vehicle-fleet-open-data-sets).

| Column | Meaning | Units |
|:--|:--|:--|
| `make`, `model`, `body_type` | vehicle description | — |
| `engine_l` | engine displacement | litres |
| `power_kw` | engine power rating | kilowatts |
| `gross_mass_kg` | maximum permitted laden mass | kilograms |
| `fuel_use_l_per_100km` | combined-cycle fuel use | litres per 100 km |

### Demonstration: engine size

```{r}
#| label: engine-loess
#| fig-cap: "Combined fuel use and engine size for 2025 petrol passenger vehicles."
#| fig-alt: "Scatterplot with engine size in litres on the horizontal axis and combined fuel use in litres per 100 kilometres on the vertical axis, showing a generally increasing relationship."
#| out-width: "100%"
ggplot(vehicles, aes(x = engine_l, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Do larger engines tend to use more fuel?"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: engine-correlation
vehicles |>
  summarise(
    engine_correlation = cor(engine_l, fuel_use_l_per_100km)
  )
```

The plot shows a positive, approximately straight and fairly strong
relationship with visible scatter. The engine-size correlation is
approximately **0.901**, supporting the description of a strong positive
linear association.

### Your turn: engine power

```{r}
#| label: power-loess
#| fig-cap: "Combined fuel use and engine power for 2025 petrol passenger vehicles."
#| fig-alt: "Scatterplot with power in kilowatts on the horizontal axis and combined fuel use in litres per 100 kilometres on the vertical axis, showing a generally increasing relationship."
#| out-width: "100%"
ggplot(vehicles, aes(x = power_kw, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Engine power (kW)",
    y = "Combined fuel use (L/100 km)",
    title = "Do more powerful engines tend to use more fuel?"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: power-correlation
vehicles |>
  summarise(
    power_correlation = cor(power_kw, fuel_use_l_per_100km)
  )
```

Power also has a strong positive relationship with fuel use. Its correlation
is approximately **0.884**, slightly smaller than the engine-size
correlation. Both plots are broadly linear, with some departures and scatter.

## 2. Fit and inspect the models (15 minutes)

### Demonstration: engine-size model

```{r}
#| label: fit-engine-model
engine_model <- lm(
  fuel_use_l_per_100km ~ engine_l,
  data = vehicles
)

engine_model
```

```{r}
#| label: plot-engine-model
#| fig-cap: "The fitted engine-size linear model."
#| fig-alt: "Scatterplot of engine size and combined fuel use with a fitted straight line."
#| out-width: "100%"
ggplot(vehicles, aes(x = engine_l, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "A straight-line model for engine size and fuel use"
  ) +
  theme_minimal(base_size = 13)
```

The printed object identifies the model formula and gives the two fitted
coefficients. The straight line captures the main increasing trend but not
all differences among vehicles.

### Your turn: power model

```{r}
#| label: fit-power-model
power_model <- lm(
  fuel_use_l_per_100km ~ power_kw,
  data = vehicles
)

power_model
```

```{r}
#| label: plot-power-model
#| fig-cap: "The fitted power linear model."
#| fig-alt: "Scatterplot of engine power and combined fuel use with a fitted straight line."
#| out-width: "100%"
ggplot(vehicles, aes(x = power_kw, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Engine power (kW)",
    y = "Combined fuel use (L/100 km)",
    title = "A straight-line model for engine power and fuel use"
  ) +
  theme_minimal(base_size = 13)
```

The power model also captures a clear increasing trend. Its points have
slightly more variation around the line than the engine-size model.

## 3. Extract and interpret the coefficients (15 minutes)

### Demonstration: engine-size coefficients

```{r}
#| label: engine-coefficients
engine_coef <- coef(engine_model)
engine_coef
```

The fitted equation is approximately

$$
\widehat{\text{fuel use}} = 4.070 + 1.886\,\text{engine size}.
$$

A one-litre difference in engine size is associated with about **1.886
L/100 km** higher average fuel use. The intercept is the line’s fitted value
at zero litres. Zero is outside the observed engine-size range, so the
intercept has no useful direct vehicle interpretation.

### Your turn: power coefficients

```{r}
#| label: power-coefficients
power_coef <- coef(power_model)
power_coef
```

The fitted equation is approximately

$$
\widehat{\text{fuel use}} = 4.705 + 0.0186\,\text{power}.
$$

A one-kilowatt difference in power is associated with about **0.0186 L/100
km** higher average fuel use. A 100 kW difference is therefore associated
with about **1.86 L/100 km** higher average fuel use. Zero kW is outside the
observed data, so the intercept should not be given a direct practical
interpretation.

::: {.callout-important}
The coefficients describe associations among these vehicle specifications.
Neither slope is a causal effect.
:::

## 4. Extract and inspect the residuals (25 minutes)

### Demonstration: engine-size residuals

```{r}
#| label: engine-augment
engine_results <- augment(engine_model)

engine_results |>
  select(engine_l, fuel_use_l_per_100km, .fitted, .resid) |>
  slice_head(n = 8)
```

`.fitted` is the point on the fitted line for an engine size. `.resid` is the
observed value minus the fitted value; positive residuals lie above the line
and negative residuals lie below it.

```{r}
#| label: engine-residual-gaps
#| fig-cap: "Observed fuel use, the engine-size fitted line, and a sample of residual gaps."
#| fig-alt: "Scatterplot with a fitted straight line and vertical segments connecting selected observations to the line."
#| out-width: "100%"
set.seed(297101)
engine_gap_data <- engine_results |>
  slice_sample(n = min(35, nrow(engine_results)))

ggplot(vehicles, aes(x = engine_l, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.35) +
  geom_line(
    data = engine_results |> arrange(engine_l),
    aes(y = .fitted),
    colour = "#2C7FB8",
    linewidth = 1
  ) +
  geom_segment(
    data = engine_gap_data,
    aes(xend = engine_l, yend = .fitted),
    alpha = 0.65,
    colour = "#D95F02"
  ) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Residuals are the vertical gaps from observations to the line"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: engine-residual-plot
#| fig-cap: "Residuals from the engine-size model versus fitted fuel-use values."
#| fig-alt: "Engine-size model residuals plotted against fitted values with a horizontal zero line."
#| out-width: "100%"
ggplot(engine_results, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.55) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted fuel use (L/100 km)",
    y = "Residual (L/100 km)",
    title = "What the engine-size model leaves unexplained"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: engine-residual-summary
engine_results |>
  summarise(
    mean_residual = mean(.resid),
    response_variance = var(fuel_use_l_per_100km),
    residual_variance = var(.resid)
  )
```

The mean residual is essentially zero. The response variance is approximately
**7.136**, while the residual variance is approximately **1.341**. Engine
size accounts for substantial variation, although other differences remain.

### Your turn: power residuals

```{r}
#| label: power-augment
power_results <- augment(power_model)

power_results |>
  select(power_kw, fuel_use_l_per_100km, .fitted, .resid) |>
  slice_head(n = 8)
```

```{r}
#| label: power-residual-plot
#| fig-cap: "Residuals from the power model versus fitted fuel-use values."
#| fig-alt: "Power-model residuals plotted against fitted values with a horizontal zero line."
#| out-width: "100%"
ggplot(power_results, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.55) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted fuel use (L/100 km)",
    y = "Residual (L/100 km)",
    title = "What the power model leaves unexplained"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: power-residual-summary
power_results |>
  summarise(
    mean_residual = mean(.resid),
    response_variance = var(fuel_use_l_per_100km),
    residual_variance = var(.resid)
  )
```

The power-model mean residual is essentially zero and its residual variance
is approximately **1.557**. This is slightly larger than the engine model’s
1.341, agreeing with the slightly smaller correlation and greater visible
scatter. Formal LINE checks still belong to Week 12.

## 5. Independent workflow: gross vehicle mass (25 minutes)

### A. Explore

```{r}
#| label: mass-explore
#| fig-cap: "Combined fuel use and gross vehicle mass."
#| fig-alt: "Scatterplot with gross vehicle mass in kilograms on the horizontal axis and combined fuel use in litres per 100 kilometres on the vertical axis, showing a moderately increasing relationship."
#| out-width: "100%"
ggplot(vehicles, aes(x = gross_mass_kg, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Gross vehicle mass (kg)",
    y = "Combined fuel use (L/100 km)",
    title = "Do heavier vehicles tend to use more fuel?"
  ) +
  theme_minimal(base_size = 13)

vehicles |>
  summarise(
    mass_correlation = cor(gross_mass_kg, fuel_use_l_per_100km)
  )
```

The mass correlation is approximately **0.576**, indicating a moderate
positive linear association that is weaker than the engine-size and power
relationships.

### B. Fit and inspect

```{r}
#| label: mass-fit
mass_model <- lm(
  fuel_use_l_per_100km ~ gross_mass_kg,
  data = vehicles
)

mass_model

ggplot(vehicles, aes(x = gross_mass_kg, y = fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Gross vehicle mass (kg)",
    y = "Combined fuel use (L/100 km)",
    title = "A straight-line model for mass and fuel use"
  ) +
  theme_minimal(base_size = 13)
```

### C. Extract and interpret coefficients

```{r}
#| label: mass-coefficients
mass_coef <- coef(mass_model)
mass_coef
```

The fitted equation is approximately

$$
\widehat{\text{fuel use}} = 1.666 + 0.00329\,\text{gross mass}.
$$

A one-kilogram difference in gross mass is associated with about **0.00329
L/100 km** higher average fuel use. More usefully, a 1000 kg difference is
associated with about **3.29 L/100 km** higher average fuel use. Zero mass is
outside the data and gives the intercept no direct practical interpretation.

### D. Inspect residuals

```{r}
#| label: mass-residuals
mass_results <- augment(mass_model)

ggplot(mass_results, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.55) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted fuel use (L/100 km)",
    y = "Residual (L/100 km)",
    title = "What the mass model leaves unexplained"
  ) +
  theme_minimal(base_size = 13)

mass_results |>
  summarise(
    mean_residual = mean(.resid),
    response_variance = var(fuel_use_l_per_100km),
    residual_variance = var(.resid)
  )
```

The mass-model mean residual is essentially zero and its residual variance is
approximately **4.772**, substantially larger than those of the engine-size
and power models.

### E. Compare all three models

| Predictor | Correlation | Residual variance |
|:--|--:|--:|
| `engine_l` | 0.901 | 1.341 |
| `power_kw` | 0.884 | 1.557 |
| `gross_mass_kg` | 0.576 | 4.772 |

Engine size is the strongest of these three single predictors: it has the
largest correlation and smallest residual variance. Power performs almost as
well. Gross mass has a moderate relationship but leaves substantially more
variation unexplained. This comparison does not establish causation or test
whether observed differences would persist in other vehicle data.

::: {.callout-important}
## Scope of today’s conclusion

Students can describe associations and how well each line summarises these
data. Inference, tests, intervals and formal LINE diagnosis belong to Weeks 11
and 12.
:::

## 6. Exit ticket and bridge to Week 11 (10 minutes)

An acceptable supported conclusion is that larger engines tend to have higher
combined fuel use among the unique 2025 petrol passenger-vehicle
specifications in these data.

Acceptable limitations include standard test values rather than real-world
driving, restriction to one vehicle year and fuel type, omitted vehicle
features, association rather than causation, and the danger of extrapolating
beyond the observed values.

An unanswered uncertainty question should point toward Week 11—for example,
how uncertain the slope is, how much variation the model explains, or how wide
a prediction interval for a new vehicle should be.

Next week covers point estimates, confidence intervals, hypothesis tests,
prediction intervals and variance explained.
