---
title: "Workshop Week 12"
subtitle: "Regression assumptions and diagnostic plots"
author: "297.101 Statistical Data Science"
date: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    code-block-bg: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
resources:
  - regression/cancer_data.csv.gz
---

Week 11 found county-level associations between poverty, median income and
lung-cancer mortality. Those fitted lines were useful summaries, but they did
not tell us whether a straight line represents the pattern well, whether its
uncertainty calculations are dependable, or whether a few unusual counties
strongly affect the result.

The data come from a broader project that combined county socioeconomic and
cancer information. Noah Rippner's
[original data-preparation code](https://github.com/nrippner/can_clin_trails_prep_r/blob/master/extract_zips_10.py)
shows how data from different sources were cleaned and joined by place. Those
joins let us examine the same county's socioeconomic conditions and mortality,
although the resulting associations do not establish causes or describe
individual people.

::: {.callout-note}
## Today's question

How well can we trust our models' estimates and predictions of county
lung-cancer mortality?
:::

We will examine prediction bands, residual patterns and influential counties,
then compare straight-line and log-predictor models. Each diagnostic reveals a
different limitation, so the aim is to connect visible evidence to its
consequences for inference and prediction.

## 1. What are we comparing?

::: {.callout-important title="Question"}
Are the poverty and income models fitted to the same counties and expressed in
units that allow a fair comparison?
:::

::: {.callout-note title="What to look for"}
Check that the three row counts agree. Both models must use the same counties
and response for their fit and diagnostics to be comparable.
:::

We rebuild both Week 11 models and check their observation counts.

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(broom)

cancer <- read_csv("regression/cancer_data.csv.gz", show_col_types = FALSE)
cancer_clean <- cancer |>
  select(Mortality_Rate_num, all_poverty, med_income, popestimate2015) |>
  filter(if_all(everything(), is.finite), popestimate2015 > 0) |>
  mutate(
    pov_norm = all_poverty / popestimate2015 * 100000,
    poverty_percent = pov_norm / 1000
  )

poverty_model <- lm(Mortality_Rate_num ~ pov_norm, data = cancer_clean)
income_model <- lm(Mortality_Rate_num ~ med_income, data = cancer_clean)
poverty_results <- augment(poverty_model)
income_results <- augment(income_model)

tibble(
  rows_retained = nrow(cancer_clean),
  poverty_model_rows = nobs(poverty_model),
  income_model_rows = nobs(income_model)
)
```

| Model | Response | Predictor |
|:--|:--|:--|
| `poverty_model` | mortality per 100,000 | poverty per 100,000 |
| `income_model` | mortality per 100,000 | median income in US dollars |

## 2. How precise are predictions for another county?

::: {.callout-important title="Question"}
How closely can either model predict mortality for an individual additional
county, and what does observed coverage tell us about future counties?
:::

A prediction band shows intervals for individual additional counties across a
range of predictor values. Its width includes the county-to-county variation
left around the fitted mean.

### Poverty-rate prediction band

::: {.callout-note title="What to look for"}
The line estimates mean mortality at each poverty value. Look at the vertical
scatter around it, then compare the band's width near the middle of the poverty
range with its width at sparse extremes.
:::

```{r}
#| label: poverty-prediction-band-data
poverty_grid <- tibble(
  pov_norm = seq(
    min(cancer_clean$pov_norm),
    max(cancer_clean$pov_norm),
    length.out = 200
  )
)

poverty_band <- bind_cols(
  poverty_grid,
  as_tibble(predict(
    poverty_model,
    newdata = poverty_grid,
    interval = "prediction",
    level = 0.95
  ))
)

stopifnot(all(poverty_band$lwr <= poverty_band$fit))
stopifnot(all(poverty_band$fit <= poverty_band$upr))
```

```{r}
#| label: poverty-prediction-band
#| fig-cap: "Fitted mortality and a 95% prediction band across the observed poverty range."
#| fig-alt: "Scatterplot of poverty rate and mortality with a fitted line and shaded prediction band."
#| out-width: "100%"
ggplot(cancer_clean, aes(x = pov_norm, y = Mortality_Rate_num)) +
  geom_ribbon(
    data = poverty_band,
    aes(x = pov_norm, ymin = lwr, ymax = upr),
    inherit.aes = FALSE,
    alpha = 0.2
  ) +
  geom_line(data = poverty_band, aes(y = fit), linewidth = 1) +
  geom_point(alpha = 0.25, size = 0.7) +
  labs(
    x = "Poverty (people per 100,000)",
    y = "Mortality (deaths per 100,000)",
    title = "Individual county outcomes vary widely around the fitted line"
  ) +
  theme_minimal(base_size = 13)
```

::: {#wk12-answer-poverty-band .callout-tip title="Worked interpretation"}
The line describes average mortality associated with poverty. The prediction
band includes county-to-county variation, so points can be far from the line.
It is generally narrowest near the centre of the observed poverty range and
wider at sparse extremes. The band displays uncertainty; it does not make
poverty a precise standalone predictor.
:::

::: {.callout-important title="Question"}
What percentage of the fitting observations falls inside the nominal 95%
prediction intervals, and why is that not validation on future counties?
:::

::: {.callout-note title="What to look for"}
The coverage calculation counts observations that were used to fit the model.
Compare the observed percentage with 95%, while remembering that this is an
in-sample description rather than a test on new counties.
:::

```{r}
#| label: poverty-observed-coverage
poverty_observed_intervals <- bind_cols(
  cancer_clean,
  as_tibble(predict(poverty_model, interval = "prediction"))
)

poverty_observed_intervals |>
  summarise(
    observations_inside = sum(between(Mortality_Rate_num, lwr, upr)),
    observations = n(),
    observed_percentage = mean(between(Mortality_Rate_num, lwr, upr)) * 100
  )
```

::: {#wk12-answer-poverty-coverage .callout-tip title="Worked interpretation"}
About 94.7% of observations used to fit the model fall inside its 95%
prediction intervals. This is an in-sample description because the same data
estimated the line and intervals. It is not validation of future coverage;
independent test data would be needed for that.
:::

### Income prediction band

::: {.callout-important title="Question"}
Where is the income prediction band narrowest, and does its width suggest
precise prediction for an individual county?
:::

::: {.callout-note title="What to look for"}
Compare the width of the income band near typical incomes with its width at
sparse extremes. Keep the fitted line's mean response distinct from an
individual county prediction.
:::

Create a 200-value income grid, calculate its 95% prediction intervals, check
that each fitted value lies between its lower and upper limits, and draw the
scatterplot, ribbon and fitted line. Then calculate observed in-sample coverage.

```{r}
#| label: income-prediction-band-data
# Use poverty-prediction-band-data as your template.
# Your code here
```

```{r}
#| label: income-prediction-band
#| fig-cap: "Fitted mortality and a 95% prediction band across the observed income range."
#| fig-alt: "Scatterplot of income and mortality with a fitted line and shaded prediction band."
#| out-width: "100%"
# Your code here
```

```{r}
#| label: income-observed-coverage
# Calculate in-sample coverage and label it accurately.
# Your code here
```

## 3. What patterns does the straight line miss?

::: {.callout-important title="Question"}
Do the residuals support a straight-line mean with independent, approximately
normal errors and reasonably constant variance?
:::

Use this reference table while discussing each diagnostic. The plots provide
evidence to weigh; none is a mechanical pass/fail certificate.

| Assumption | What it means | Evidence to inspect | Why it matters |
|:--|:--|:--|:--|
| Linearity | The conditional mean is represented adequately by a straight line. | Residuals versus fitted and the smooth. | Curvature can bias fitted means and predictions. |
| Independence | Errors across observations are independent; consider design and spatial dependence. | County sampling information and geographic residual patterns. | Dependence can make standard errors and intervals too optimistic. |
| Normality | Conditional errors are approximately normal; residuals are an imperfect diagnostic. | Centre and both tails of the Q-Q plot. | Tail departures can affect small-sample inference and prediction intervals. |
| Equal variance | Conditional error variance is reasonably constant. | Scale-location spread and smooth. | Changing spread makes interval precision vary across predictors. |

Use the same three-step routine for every plot:

1. state what pattern would support the assumption;
2. describe what you actually see; and
3. judge whether the departure matters for inference or prediction.

### L — Linearity

::: {.callout-important title="Question"}
Does the fitted line leave a curved pattern in the residuals, and where would
that cause the model to overpredict or underpredict mean mortality?
:::

Residuals versus fitted values should form a shapeless cloud around zero. A
curved smooth suggests that a straight-line mean misses structure.

#### Poverty-rate model

::: {.callout-note title="What to look for"}
Look for a cloud centred around zero and a roughly horizontal smooth. A bend or
region where residuals share a sign indicates that the straight line misses
part of the conditional mean.
:::

```{r}
#| label: poverty-linearity
#| fig-cap: "Residuals versus fitted values for the poverty-rate model."
#| fig-alt: "Residual plot with a horizontal zero line and smooth curve."
#| out-width: "100%"
ggplot(poverty_results, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.3, size = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Fitted mortality", y = "Residual",
       title = "Does the poverty model leave a curved pattern?") +
  theme_minimal(base_size = 13)
```

::: {#wk12-answer-poverty-linearity .callout-tip title="Worked interpretation"}
The poverty smooth is slightly negative at low fitted mortality, rises around
the middle and falls below zero at high values. The broad cloud still surrounds
zero, but the curved smooth suggests modest non-linearity. At the sparse high
end the line tends to overpredict some counties, so a straight-line mean is an
approximation whose consequences should be considered.
:::

#### Income model

::: {.callout-important title="Question"}
Is the income model's curvature similar to the poverty model's, and what would
it imply about using a straight-line mean?
:::

::: {.callout-note title="What to look for"}
Recreate the same residual plot, then compare the smooth's direction and its
position relative to zero with the poverty model.
:::

```{r}
#| label: income-linearity
# Recreate the plot with income_results.
# Your code here
```

### E — Equal variance

::: {.callout-important title="Question"}
Does residual variation stay reasonably constant across fitted mortality, or
does prediction precision change across the range?
:::

The scale-location plot uses

$$
\sqrt{|\text{standardized residual}|}.
$$

A roughly horizontal band and smooth support constant variance; a trend or
funnel suggests that prediction precision changes across fitted values.

#### Poverty-rate model

::: {.callout-note title="What to look for"}
The vertical axis shows residual magnitude without its sign. Look for a level
band and flat smooth; the smooth need not be at zero.
:::

```{r}
#| label: poverty-equal-variance
poverty_results |>
  mutate(root_abs_standardized_residual = sqrt(abs(.std.resid))) |>
  ggplot(aes(x = .fitted, y = root_abs_standardized_residual)) +
  geom_point(alpha = 0.3, size = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Fitted mortality",
    y = "Sqrt(|standardized residual|)",
    title = "Is residual spread stable for the poverty model?"
  ) +
  theme_minimal(base_size = 13)
```

::: {#wk12-answer-poverty-equal-variance .callout-tip title="Worked interpretation"}
The scale-location smooth rises as fitted mortality increases, suggesting larger
residual magnitudes in high-fitted-mortality counties. Prediction precision is
therefore less stable across the range, and a constant-variance description is
an approximation. This plot shows the direction and size of the departure rather
than providing a formal pass/fail decision.
:::

#### Income model

::: {.callout-important title="Question"}
Does income residual magnitude rise, fall or form a U-shape, and where would
prediction intervals be least precise?
:::

::: {.callout-note title="What to look for"}
Use residual magnitude without its sign. A non-horizontal smooth indicates
that prediction precision changes across fitted mortality.
:::

```{r}
#| label: income-equal-variance
# Recreate the scale-location plot with income_results.
# Your code here
```

### N — Normality

::: {.callout-important title="Question"}
How closely do the standardised residuals follow a normal distribution,
especially in the tails?
:::

In a normal Q-Q plot, standardized residuals should follow the reference line.
Systematic bends or strong tail departures indicate non-normal residuals.

#### Poverty-rate model

::: {.callout-note title="What to look for"}
Compare theoretical normal quantiles with the ordered standardised residuals.
Inspect the centre and the lower and upper tails separately.
:::

```{r}
#| label: poverty-normality
ggplot(poverty_results, aes(sample = .std.resid)) +
  geom_qq(alpha = 0.35, size = 0.8) +
  geom_qq_line() +
  labs(
    x = "Theoretical normal quantiles",
    y = "Standardized residual quantiles",
    title = "How closely do poverty-model residuals follow a normal distribution?"
  ) +
  theme_minimal(base_size = 13)
```

::: {#wk12-answer-poverty-normality .callout-tip title="Worked interpretation"}
The centre follows the reference line fairly closely, while both tails depart,
especially the upper tail. This indicates residual tails unlike a normal model.
The plot diagnoses errors, not the distribution of predictors or mortality
values. With many counties, coefficient estimates may be stable, but tail
departures can still affect individual prediction intervals.
:::

#### Income model

::: {.callout-important title="Question"}
How close is the income model's centre to the reference line, and how does each
tail depart?
:::

::: {.callout-note title="What to look for"}
Describe the centre and both tails separately, then connect any departures to
coefficient inference and individual-county prediction.
:::

```{r}
#| label: income-normality
# Recreate the Q-Q plot with income_results.
# Your code here
```

### I — Independence

::: {.callout-important title="Question"}
Could nearby counties have related errors, and can these data investigate that
possibility?
:::

Independence comes mainly from study design and data collection. One row per
county prevents duplicated counties, but nearby counties may share health
systems, environment and demographics. Row order is not meaningful geographic
distance, so a residual-versus-row-number plot cannot resolve this concern.

#### Poverty-rate model

::: {.callout-note title="What to look for"}
Independence is about errors across counties, not the order of rows in the
file. Look for information about the sampling design, county identifiers and
geographic locations; neighbouring counties may share unmeasured conditions.
:::

::: {#wk12-answer-independence .callout-tip title="Worked interpretation"}
Neighbouring counties may share air quality, smoking patterns, health services,
policy and population composition, producing related errors. CSV row order is
arbitrary rather than geographic, so a row-number plot cannot establish
independence. Coordinates, adjacency information or a sampling design that
accounts for spatial clusters would help investigate spatial dependence.
:::

#### Income model

::: {.callout-important title="Question"}
Would changing the predictor from poverty to income remove possible spatial
dependence? Explain your answer.
:::

## 4. Do a few counties strongly affect the fitted line?

::: {.callout-important title="Question"}
Do a small number of counties have enough influence to change the fitted
association appreciably?
:::

Influence is separate from independence. Cook's distance measures how strongly
one observation affects the fitted coefficients through its predictor value
and residual.

### Poverty-rate model

::: {.callout-note title="What to look for"}
Cook's distance combines an unusual predictor position with a large residual.
The screening guide used in the plot is

$$
\frac{4}{n}.
$$

Values above it prompt investigation rather than automatic deletion.
:::

```{r}
#| label: poverty-influence
poverty_influence <- poverty_results |>
  mutate(observation = row_number())

ggplot(poverty_influence, aes(x = observation, y = .cooksd)) +
  geom_col() +
  geom_hline(
    yintercept = 4 / nrow(poverty_influence),
    linetype = "dashed"
  ) +
  labs(
    x = "Observation",
    y = "Cook's distance",
    title = "Which counties most influence the poverty-rate model?",
    subtitle = "The dashed 4/n line is a screening guide, not a pass/fail test"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: poverty-most-influential
poverty_influence |>
  arrange(desc(.cooksd)) |>
  select(observation, pov_norm, Mortality_Rate_num, .fitted, .resid, .cooksd) |>
  slice_head(n = 5)
```

::: {#wk12-answer-poverty-influence .callout-tip title="Worked interpretation"}
The largest poverty-model Cook's distances are around 0.043, with several
observations worth checking against the screening line. These values combine
unusual poverty positions and residual sizes. Investigate data quality, county
context and sensitivity of the fitted line before considering any removal.
Influence concerns coefficient stability and is distinct from independence.
:::

### Income model

::: {.callout-important title="Question"}
Which income observations have the largest Cook's distances, and do their
values reflect unusual income, large residuals or both?
:::

Create the Cook's-distance plot and table of five largest values for the
income model. Compare the influential observations with the poverty model.

::: {.callout-note title="What to look for"}
Compare the largest income values with the counties highlighted by the poverty
model, then consider a sensitivity refit before drawing a conclusion.
:::

```{r}
#| label: income-influence
# Your code here
```

```{r}
#| label: income-most-influential
# Your code here
```

## 5. Does logging the predictor improve the model?

::: {.callout-important title="Question"}
Does logging poverty or income reduce the diagnostic problems enough to make a
more useful model?
:::

Logging a predictor can straighten a curved relationship. The response remains
on its original mortality scale. In the model

$$
y=a+b\log(x),
$$

a 1% increase in the predictor is associated approximately with

$$
0.01b
$$

units of the response.

### Log-poverty model

::: {.callout-note title="What to look for"}
Compare the linear and log-predictor models on the same counties. Inspect
curvature, residual magnitude, tail behaviour and influential observations
before choosing on the basis of R-squared alone.
:::

```{r}
#| label: fit-log-poverty
poverty_comparison_data <- cancer_clean |> filter(pov_norm > 0)

poverty_linear_same_rows <- lm(
  Mortality_Rate_num ~ pov_norm,
  data = poverty_comparison_data
)
poverty_log_model <- lm(
  Mortality_Rate_num ~ log(pov_norm),
  data = poverty_comparison_data
)

stopifnot(nobs(poverty_linear_same_rows) == nobs(poverty_log_model))

tibble(
  model = c("Linear predictor", "Log predictor"),
  observations = c(nobs(poverty_linear_same_rows), nobs(poverty_log_model)),
  r_squared = c(
    glance(poverty_linear_same_rows)$r.squared,
    glance(poverty_log_model)$r.squared
  ),
  residual_sd = c(
    glance(poverty_linear_same_rows)$sigma,
    glance(poverty_log_model)$sigma
  )
)
```

```{r}
#| label: log-poverty-interpretation
log_poverty_slope <- coef(poverty_log_model)[["log(pov_norm)"]]
tibble(
  approximate_change_for_one_percent_more_poverty =
    0.01 * log_poverty_slope
)
```

```{r}
#| label: log-poverty-diagnostics
augment(poverty_log_model) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.3, size = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Fitted mortality",
    y = "Residual (deaths per 100,000)",
    title = "Residual pattern after logging poverty"
  ) +
  theme_minimal(base_size = 13)
```

The comparison uses exactly the same positive-poverty counties for both
models. `bind_rows()` adds a model label, and the ordered factor keeps the
linear predictor beside the log predictor in each figure.

```{r}
#| label: log-poverty-comparison-prep
poverty_comparison_results <- bind_rows(
  augment(poverty_linear_same_rows) |> mutate(model = "Linear predictor"),
  augment(poverty_log_model) |> mutate(model = "Log predictor")
) |>
  mutate(
    model = factor(model, levels = c("Linear predictor", "Log predictor")),
    observation = rep(seq_len(nrow(poverty_comparison_data)), 2)
  )
```

```{r}
#| label: log-poverty-diagnostics-comparison
#| fig-cap: "Residuals versus fitted values for the linear and log-poverty models."
#| fig-alt: "Two faceted residual plots comparing linear and log poverty predictors."
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "100%"
ggplot(poverty_comparison_results, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.2, size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ model, nrow = 1) +
  labs(x = "Fitted mortality", y = "Residual (deaths per 100,000)") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: log-poverty-scale-location
#| fig-cap: "Scale-location diagnostics for the two poverty models."
#| fig-alt: "Faceted scale-location plots showing residual magnitude against fitted mortality."
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "100%"
ggplot(poverty_comparison_results,
       aes(x = .fitted, y = sqrt(abs(.std.resid)))) +
  geom_point(alpha = 0.2, size = 0.5) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ model, nrow = 1) +
  labs(x = "Fitted mortality", y = "Square root of absolute standardized residual") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: log-poverty-qq
#| fig-cap: "Normal Q-Q diagnostics for the two poverty models."
#| fig-alt: "Faceted Q-Q plots comparing theoretical normal quantiles with standardized residuals."
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "100%"
ggplot(poverty_comparison_results, aes(sample = .std.resid)) +
  stat_qq(alpha = 0.2, size = 0.6) +
  stat_qq_line() +
  facet_wrap(~ model, nrow = 1) +
  labs(x = "Theoretical normal quantiles", y = "Ordered standardized residuals") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: log-poverty-influence
#| fig-cap: "Cook's distances for the two poverty models."
#| fig-alt: "Faceted Cook's-distance plots with a 4 over n screening line."
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "100%"
ggplot(poverty_comparison_results, aes(x = observation, y = .cooksd)) +
  geom_col(width = 1, alpha = 0.55) +
  geom_hline(yintercept = 4 / nrow(poverty_comparison_data),
             linetype = "dashed") +
  facet_wrap(~ model, nrow = 1) +
  labs(x = "Observation index (same counties in both panels)",
       y = "Cook's distance") +
  theme_minimal(base_size = 12)
```

::: {#wk12-answer-log-poverty .callout-tip title="Worked interpretation"}
Logging poverty raises R-squared modestly, from about 0.158 to 0.171, and can
reduce the visible curvature in the residual smooth. The remaining spread and
tail departures mean the transformed model is not automatically adequate, and
possible county-level spatial dependence remains. In

$$
y=a+b\log(x),
$$

a 1% relative increase in poverty changes fitted mortality by approximately

$$
0.01b
$$

deaths per 100,000; this is not a percentage change in mortality.

| Evidence | Linear predictor | Log predictor | Interpretation |
|:--|:--|:--|:--|
| Residual curvature | Curved smooth | Closer to zero centrally | Logging modestly reduces mean-pattern curvature. |
| Changing residual magnitude | Increasing at high fitted values | Spread remains | Intervals are not equally precise throughout. |
| Q-Q tail departures | Both tails depart | Both tails depart | Tail behaviour remains a limitation. |
| Influential observations | Spikes worth checking | Spikes worth checking | Investigate rather than delete automatically. |
| R-squared and residual SD | 0.158; larger SD | 0.171; smaller SD | Fit improves modestly on identical rows. |
:::

### Log-income model

::: {.callout-important title="Question"}
Does the log-income model improve the residual pattern and fit on the same
counties, and how would we interpret a 1% relative income increase?
:::

Fit `income_linear_same_rows` and `income_log_model` on identical positive
income observations. Compare R-squared, residual standard deviation, the
residual plot and the fitted mortality difference.

::: {.callout-note title="What to look for"}
Check that both models use identical observations. Compare curvature, changing
spread, Q-Q tails, Cook's-distance spikes and overall fit rather than relying
on one diagnostic.
:::

```{r}
#| label: fit-log-income
# Your code here
```

```{r}
#| label: log-income-interpretation
# Your code here
```

```{r}
#| label: log-income-diagnostics
# Your code here
```

Repeat the four comparison figures for income. Keep the rows identical within
the comparison, use `income_comparison_results`, and retain the four labels
`log-income-diagnostics-comparison`, `log-income-scale-location`,
`log-income-qq` and `log-income-influence`.

```{r}
#| label: log-income-diagnostics-comparison
# Your code here
```

```{r}
#| label: log-income-scale-location
# Your code here
```

```{r}
#| label: log-income-qq
# Your code here
```

```{r}
#| label: log-income-influence
# Your code here
```

Complete the same comparison table for income:

| Evidence | Linear predictor | Log predictor | Interpretation |
|:--|:--|:--|:--|
| Residual curvature |  |  |  |
| Changing residual magnitude |  |  |  |
| Q-Q tail departures |  |  |  |
| Influential observations |  |  |  |
| R-squared and residual SD |  |  |  |

## 6. Which model would we use, and with what limitations?

::: {.callout-important title="Question"}
Which income model would we use, and what limitations must accompany that
recommendation?
:::

::: {.callout-note title="What to look for"}
Weigh linearity, equal variance, normality, influence and relative fit on the
same counties. Include the unresolved independence concern and the lack of
out-of-sample validation.
:::

Recommend either the linear-income or log-income model. Support the choice
with specific evidence about:

1. linearity;
2. equal variance;
3. normality;
4. influential observations; and
5. relative fit on the same counties.

Also state the unresolved independence concern and one limitation of judging a
model with the data used to fit it. A transformation is useful only if the
diagnostics and interpretation support it.

## Conclusion

The diagnostics show why an association and a fitted prediction interval are
only the beginning of evaluating a model. Curvature, changing residual spread,
tail departures, influential counties and possible spatial dependence all
qualify how confidently these models can support inference and prediction.
Comparing transformations on identical counties can improve the model's shape,
but it does not provide evidence about performance on new counties.

::: {.callout-tip title="Optional question"}
If a transformation improves the diagnostics on these counties, does that
guarantee better predictions for new counties?
:::
