---
title: "Workshop Week 11: Regression Testing"
subtitle: "Testing associations and making predictions"
author:
  - name: "Nick Knowlton"
  - name: "297.101 Statistical Data Science"
date-meta: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    code-block-bg: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
resources:
  - regression/cancer_data.csv.gz
---

::: {.callout-note}
## Today's question

What can county data tell us about the association between poverty, median
income and lung-cancer mortality?

For each new method, first work through the **Demonstration** using poverty
rate. Then complete **Your turn** using median income. The instructor leads
the Demonstration; run that code and discuss it before starting Your turn.

Discuss each demonstration before opening its worked interpretation. Then
complete Your turn using the demonstration as your template.
:::

## Learning objectives

By the end of this workshop you should be able to:

- interpret coefficients, confidence intervals and p-values;
- connect residual variance with \(R^2\);
- make predictions with confidence and prediction intervals; and
- identify extrapolation and limits of an observational model.

## 1. Prepare and explore the data (20 minutes)

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(purrr)
library(broom)
library(GGally)

cancer <- read_csv("regression/cancer_data.csv.gz", show_col_types = FALSE)

cancer_clean <- cancer |>
  select(Mortality_Rate_num, all_poverty, med_income, popestimate2015) |>
  filter(if_all(everything(), is.finite), popestimate2015 > 0) |>
  mutate(
    pov_norm = all_poverty / popestimate2015 * 100000,
    poverty_percent = pov_norm / 1000
  )

tibble(rows_in_source = nrow(cancer), rows_retained = nrow(cancer_clean))
```

Each row represents one US county with complete values for today's variables.
The source combines lung-cancer incidence and mortality information from the
National Cancer Institute's SEER programme with socioeconomic measures from
the US Census Bureau's American Community Survey. The data were preprocessed
following [Nick Rippner's cancer-analysis tutorial](https://data.world/nrippner/cancer-analysis-hackathon-challenge).
You can [download the supplied data file](regression/cancer_data.csv.gz).
These are observational county-level data, so a fitted association does not
show that changing poverty or income would cause mortality to change.

| Variable | Meaning | Units |
|:--|:--|:--|
| `Mortality_Rate_num` | Lung-cancer mortality rate | deaths per 100,000 people |
| `all_poverty` | Residents below the poverty threshold | people |
| `pov_norm` | Population-adjusted poverty count | people per 100,000 |
| `poverty_percent` | Population-adjusted poverty rate | percent |
| `med_income` | Median household income | US dollars |

### Demonstration: why adjust the poverty count?

A count of 10,000 people in poverty means something different in a county of
50,000 people than in a county of one million people:

```{r}
#| label: count-rate-example
tibble(
  county = c("County A", "County B"),
  all_poverty = c(10000, 10000),
  population = c(50000, 1000000)
) |>
  mutate(
    poverty_per_100000 = all_poverty / population * 100000,
    poverty_percent = poverty_per_100000 / 1000
  )
```

::: {.callout-important title="Discuss"}
The two counties have the same poverty count. How different are their poverty
percentages? What does population adjustment change about the predictor?
:::

::: {#wk11-answer-poverty-counts .callout-tip collapse="true" title="Worked interpretation — open after discussion"}
The same count represents 20% poverty in County A but only 1% in County B.
Dividing by population changes the predictor from a mixture of county size and
poverty into a comparable rate. A predictor and response do not need matching
units for regression: the units determine how we interpret the slope.
:::

The response and predictor need not share units. Adjustment matters because it
changes the predictor from county size mixed with poverty into the proportion
of residents in poverty.

### Demonstration: read a draftsman plot

The diagonal shows distributions, the lower panels show scatterplots, and the
upper panels show correlations.

```{r}
#| label: draftsman-plot
#| fig-cap: "Distributions and pairwise relationships among poverty rate, median income and lung-cancer mortality."
#| fig-alt: "A three by three matrix with distributions on the diagonal, scatterplots below and correlations above."
#| fig-width: 9
#| fig-height: 9
#| out-width: "100%"
draftsman_data <- cancer_clean |>
  transmute(
    `Poverty (%)` = poverty_percent,
    `Median income ($)` = med_income,
    `Mortality (deaths per 100,000)` = Mortality_Rate_num
  )

draftsman_plot <- ggpairs(
  draftsman_data,
  progress = FALSE,
  diag = list(continuous = wrap("densityDiag", alpha = 0.5)),
  lower = list(continuous = wrap("points", alpha = 0.25, size = 0.55)),
  upper = list(continuous = wrap("cor", size = 4, stars = FALSE))
) + theme_minimal(base_size = 11)

print(draftsman_plot)
```

::: {.callout-note title="What to look for"}
The diagonal panels show distributions, the lower panels show scatterplots and
the upper panels show correlations. In the lower-left panel, mortality is the
vertical axis and poverty percentage is the horizontal axis.
:::

::: {.callout-important title="Discuss"}
What direction and strength do you see in the poverty–mortality panel? Does the
cloud look tight enough for poverty alone to predict mortality precisely?
:::

::: {#wk11-answer-draftsman .callout-tip collapse="true" title="Worked interpretation — open after discussion"}
The lower-left panel plots mortality against poverty percentage. It shows a
moderate positive association: counties with higher poverty tend to have higher
mortality, but the cloud is broad. The upper panel reports approximately
\(r=0.398\), confirming positive but not strong linear association. The
correlation summarises direction and linear strength; the plot shows spread,
possible curvature and unusual observations that a single number hides.
:::

Describe the direction, form and strength of the poverty–mortality
relationship. What does its correlation add to the scatterplot?

### Your turn: explore median income

Describe the income–mortality relationship from the matrix. Then make a
full-size scatterplot with a LOESS smooth and calculate its correlation.

```{r}
#| label: income-exploration
#| fig-cap: "Lung-cancer mortality and median household income."
#| fig-alt: "Scatterplot of income and mortality with a smooth curve."
#| out-width: "100%"
# Use med_income and Mortality_Rate_num.
# Add points, a LOESS smooth, labels with units and theme_minimal().
# Your code here
```

```{r}
#| label: income-correlation
# Calculate the correlation between med_income and Mortality_Rate_num.
# Your code here
```

How does the full-size plot change or strengthen your description?

## 2. Fit and interpret the models (20 minutes)

### Demonstration: poverty-rate model

The coefficient table has one row for the intercept and one for the predictor.
Estimate is the fitted change in mortality per one predictor unit (or the
fitted value when the predictor is zero); Std. Error measures the sampling
uncertainty of that estimate. The t statistic and p-value use these columns.

The slope is tiny per one unit of `pov_norm`, so interpret a change of 10,000
per 100,000: **10 percentage points**.

```{r}
#| label: fit-poverty-model
poverty_model <- lm(Mortality_Rate_num ~ pov_norm, data = cancer_clean)
tidy(poverty_model)
```

```{r}
#| label: interpret-poverty-slope
poverty_slope <- coef(poverty_model)[["pov_norm"]]
tibble(
  change_in_poverty_percentage_points = 10,
  fitted_change_in_mortality = poverty_slope * 10000
)
```

::: {.callout-note title="What to look for"}
Interpret the slope using a meaningful 10-percentage-point change, not one
person per 100,000. Check whether zero poverty is represented before using the
intercept practically.
:::

::: {.callout-important title="Discuss"}
What mortality difference does the slope imply for 10 percentage points? Is the
intercept a realistic county comparison? Does this show causation?
:::

::: {#wk11-answer-poverty-coefficients .callout-tip collapse="true" title="Worked interpretation — open after discussion"}
The poverty slope implies about 9.13 more deaths per 100,000 for a 10
percentage-point higher poverty rate. Zero poverty is outside the observed
range, so the intercept is mainly a mathematical anchor. Because these are
observational county data, the fitted association can reflect confounding and
does not show that changing poverty would cause mortality to change.
:::

Interpret the sign and size in context. Why is this an association rather
than a causal effect? Is the intercept useful for counties in these data?

### Your turn: income model

Fit `Mortality_Rate_num ~ med_income` as `income_model`, show its coefficient
table, and calculate the fitted mortality difference for $10,000 of income.

```{r}
#| label: fit-income-model
# Use fit-poverty-model as your template.
# Your code here
```

```{r}
#| label: interpret-income-slope
# Extract the med_income slope and multiply it by 10000.
# Your code here
```

Interpret the result in deaths per 100,000. Why would comparing the two raw
slope coefficients be misleading?

## 3. Confidence intervals and hypothesis tests (30 minutes)

### Demonstration: poverty-rate slope

A 95% confidence interval uses the slope, its standard error and a t critical
value. Every value below comes from the fitted model. The calculation is

\[
\widehat\beta_1 \pm t_{0.975,n-2}\operatorname{SE}(\widehat\beta_1).
\]

Here \(\widehat\beta_1\) is the estimated slope, SE is its standard error,
the critical value \(t_{0.975,n-2}\) leaves 2.5% in each tail, and
df.residual() supplies \(n-2\) because the intercept and slope are estimated.

```{r}
#| label: poverty-manual-confidence-interval
poverty_slope_row <- tidy(poverty_model) |> filter(term == "pov_norm")
critical_t <- qt(0.975, df = df.residual(poverty_model))

poverty_ci_manual <- poverty_slope_row |>
  transmute(
    lower = estimate - critical_t * std.error,
    estimate,
    upper = estimate + critical_t * std.error
  )

poverty_ci_manual
confint(poverty_model, "pov_norm", level = 0.95)
```

::: {.callout-note title="What to look for"}
The interval is estimate ± a t critical value times its standard error. Use
`df.residual()` because the fitted intercept and slope consume two degrees of
freedom.
:::

::: {.callout-important title="Discuss"}
Does the interval include zero? What range does it give for a 10-percentage-point
difference? What does 95% refer to over repeated samples?
:::

::: {#wk11-answer-poverty-confidence-interval .callout-tip collapse="true" title="Worked interpretation — open after discussion"}
The slope interval is approximately 0.000835 to 0.000991 per poverty unit.
Multiplying by 10,000 gives 8.35 to 9.91 deaths per 100,000 for a
10-percentage-point difference. Zero is not in the interval. In repeated samples
from the same process, approximately 95% of intervals constructed this way
would contain the population slope; this is not a 95% probability that this
particular fixed slope lies inside the interval.
:::

Multiply all three values by 10,000 to interpret them for a
10-percentage-point difference. Does the interval include zero?

```{r}
#| label: poverty-manual-p-value
poverty_t <- poverty_slope_row$estimate / poverty_slope_row$std.error
poverty_p_manual <- 2 * pt(
  -abs(poverty_t),
  df = df.residual(poverty_model)
)
tibble(
  t_statistic = poverty_t,
  manual_p_value = poverty_p_manual,
  tidy_p_value = poverty_slope_row$p.value
)
```

::: {.callout-important title="Discuss"}
What null hypothesis produces this t statistic? What does the two-sided
p-value measure, and what does it not establish?
:::

::: {#wk11-answer-poverty-p-value .callout-tip collapse="true" title="Worked interpretation — open after discussion"}
The test is \(H_0:\beta_1=0\). The t statistic compares the estimated slope with
its standard error, and the two-sided p-value is the probability, under the
null and model assumptions, of a result at least this far from zero in either
direction. Its extremely small value is strong evidence against a zero slope
under the model. It does not establish causation, practical importance or that
the model assumptions are correct.
:::

The p-value tests \(H_0: \beta_1=0\). State what it says about evidence
against a zero slope. It does not measure effect size or practical importance.

### Your turn: income slope

Construct the 95% confidence interval and two-sided p-value for the income
slope. Check the results against `confint()` and `tidy()`.

```{r}
#| label: income-manual-confidence-interval
# Extract the med_income row and calculate estimate +/- critical_t * SE.
# Your code here
```

```{r}
#| label: income-manual-p-value
# Calculate the t statistic and its two-sided p-value.
# Your code here
```

Interpret the interval for a $10,000 income difference. What does the p-value
establish, and what does it leave unanswered?

## 4. Connect residual variance and R-squared (15 minutes)

### Demonstration: poverty-rate model

With an intercept, the fraction of sample variation explained is
\(1-\operatorname{Var}(e)/\operatorname{Var}(y)\).

```{r}
#| label: poverty-r-squared
poverty_results <- augment(poverty_model)
poverty_results |>
  summarise(
    response_variance = var(Mortality_Rate_num),
    residual_variance = var(.resid),
    variance_explained = 1 - residual_variance / response_variance
  ) |>
  mutate(model_r_squared = glance(poverty_model)$r.squared)
```

::: {.callout-important title="Discuss"}
How do residual variance and response variance combine to form R-squared? What
percentage remains unexplained?
:::

::: {#wk11-answer-poverty-r-squared .callout-tip collapse="true" title="Worked interpretation — open after discussion"}
The ratio \(1-\operatorname{Var}(e)/\operatorname{Var}(y)\) agrees with the
reported \(R^2\) because least squares decomposes response variation into fitted
and residual parts. Here the poverty model explains about 15.8% of sample
mortality variation, leaving about 84.2% associated with other measured or
unmeasured differences and random variation. Explained variation is descriptive,
not a causal percentage.
:::

Why do the two versions agree apart from rounding? What remains unexplained?

### Your turn: compare the models

Repeat the calculation for `income_model`. Both models use the same response
and the same retained counties.

```{r}
#| label: income-r-squared
# Augment income_model and calculate both versions of R-squared.
# Your code here
```

Which model accounts for more observed variation? Does that identify a cause?

## 5. Prediction and uncertainty (25 minutes)

Before predicting, inspect the predictor support so you can distinguish
interpolation from extrapolation.

```{r}
#| label: observed-ranges
cancer_clean |>
  summarise(
    minimum_poverty_percent = min(poverty_percent),
    maximum_poverty_percent = max(poverty_percent),
    minimum_income = min(med_income),
    maximum_income = max(med_income)
  )
```

### Demonstration: predict at 20% poverty

Twenty percent poverty equals 20,000 people per 100,000.

```{r}
#| label: poverty-point-prediction
poverty_20 <- tibble(pov_norm = 20000)
poverty_coef <- coef(poverty_model)
manual_prediction <- poverty_coef[["(Intercept)"]] +
  poverty_coef[["pov_norm"]] * poverty_20$pov_norm

tibble(manual_prediction)
predict(poverty_model, newdata = poverty_20)
```

::: {.callout-important title="Discuss"}
Why is 20% represented by 20,000 in this model? What quantity is predicted?
:::

::: {#wk11-answer-poverty-point-prediction .callout-tip collapse="true" title="Worked interpretation — open after discussion"}
Twenty percent of 100,000 is 20,000 people per 100,000, matching the units of
`pov_norm`. The manual fitted equation and `predict()` agree at about 56.61
deaths per 100,000. This is a predicted county mortality rate, not a prediction
about one individual person.
:::

A confidence interval estimates mean mortality among counties at this
predictor value. A prediction interval is for one additional county and also
includes county-to-county variation, so it is wider.

```{r}
#| label: poverty-intervals
predict(poverty_model, poverty_20, interval = "confidence", level = 0.95)
predict(poverty_model, poverty_20, interval = "prediction", level = 0.95)
```

```{r}
#| label: poverty-confidence-levels
map_dfr(c(0.90, 0.95, 0.99), \(confidence_level) {
  as_tibble(predict(
    poverty_model, poverty_20,
    interval = "prediction", level = confidence_level
  )) |>
    mutate(level = confidence_level, .before = 1)
})
```

::: {.callout-important title="Discuss"}
Which interval estimates the mean mortality and which covers one additional
county? Which is wider? How does width change from 90% to 99%?
:::

::: {#wk11-answer-poverty-intervals .callout-tip collapse="true" title="Worked interpretation — open after discussion"}
The confidence interval describes the mean mortality for counties at 20%
poverty. The prediction interval describes one additional county and includes
individual county variation, so it is wider. Raising confidence from 90% to
95% to 99% increases the critical value and therefore widens both kinds of
interval. Wider intervals provide more coverage at the cost of precision.
:::

What happens to interval width as confidence increases?

### Your turn: predict at $50,000 income

Calculate the point prediction manually and with `predict()`. Then obtain 95%
confidence and prediction intervals and 90%, 95% and 99% prediction intervals.

```{r}
#| label: income-point-prediction
# Create income_50000, calculate manually, and check with predict().
# Your code here
```

```{r}
#| label: income-intervals
# Calculate 95% confidence and prediction intervals.
# Your code here
```

```{r}
#| label: income-confidence-levels
# Use poverty-confidence-levels as your template.
# Your code here
```

Explain which 95% interval is wider and why.

### Demonstration: check extrapolation

```{r}
#| label: poverty-extrapolation
predict(
  poverty_model,
  newdata = tibble(pov_norm = c(0, 75000)),
  interval = "prediction"
)
```

::: {.callout-important title="Discuss"}
Compare 0% and 75% with the observed poverty range. Why can R return a number
without that number being well supported by the data?
:::

::: {#wk11-answer-poverty-extrapolation .callout-tip collapse="true" title="Worked interpretation — open after discussion"}
Observed poverty ranges from about 2.60% to 46.49%. Both 0% and 75% are outside
that range, so these are extrapolations. R applies the fitted equation beyond
the observed data, but the straight-line relationship and its uncertainty have
not been checked there. Such predictions should be treated as model scenarios,
not evidence-based estimates for actual counties.
:::

Both 0% and 75% poverty lie outside the observed range. R calculates results,
but the fitted relationship has not been checked there. Why should these
predictions be treated cautiously?

## 6. Independent interpretation (10 minutes)

Write a short conclusion about the income model including:

1. the association per $10,000 and its 95% confidence interval;
2. the p-value and what it means;
3. the prediction and 95% prediction interval at $50,000;
4. the model's \(R^2\); and
5. one limitation of these observational county data.

Avoid causal language and distinguish statistical evidence from practical
importance.

## Key takeaways

- Population adjustment turns a poverty count into a comparable county rate.
- Prediction intervals include individual variation and are wider than
  confidence intervals for the mean.
- A p-value addresses evidence against a zero slope; \(R^2\) describes
  explained sample variation.
- Predictions outside the observed range rely on unchecked assumptions.
