---
title: "Workshop Week 12: Inference and Assumptions"
subtitle: "Student workbook: uncertainty, intervals and diagnostics"
author:
  - name: "Nick Knowlton"
  - name: "297.101 Statistical Data Science"
date-meta: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

In Weeks 10 and 11 we used vehicle data to build simple and multiple regression
models, interpret coefficients, standard errors, p-values and $R^2$, and make
fitted predictions. This week we apply the same ideas to a new dataset and ask
a harder question: how much can we trust the model's estimates and predictions?

We will use county-level lung-cancer mortality data to introduce confidence
intervals, prediction intervals and regression diagnostics. We use a new
dataset this week so you can apply the same regression workflow in a different
setting.

The data come from a broader project that combined county socioeconomic and
cancer information. Noah Rippner's
[Python data-preparation code](https://github.com/nrippner/can_clin_trails_prep_r/blob/master/extract_zips_10.py)
shows how data from different sources were cleaned and joined by place. Those
joins let us examine the same county's socioeconomic conditions and mortality,
although the resulting associations do not establish causes or describe
individual people.

This workshop asks how uncertain the model estimates are and whether the
model's assumptions are reasonable.

## Today's question

How well can we trust our model's estimates and predictions of county
lung-cancer mortality?

## Learning objectives

By the end of this workshop you should be able to:

- use standard errors to interpret uncertainty in regression coefficients;
- interpret confidence intervals and hypothesis tests;
- distinguish confidence intervals for an average from prediction intervals for
  an individual county;
- interpret supplied LINE diagnostic plots and use study design to assess a
  model; and
- evaluate one transformed alternative without relying on R-squared alone.

## How to use this workbook

Work through the sections in order. Run each supplied demonstration and inspect
its output or plot before moving on. In each **Your turn** section, write your
own code in the blank chunk, then record your interpretation beside the prompt.
Hints appear when a function or idea is first introduced; later tasks expect
you to reuse it. Open this file from the `labs` folder so the supplied data file
can be found.

## 1. Fit and understand one primary cancer model

**What to do:** Fit the multiple regression model and identify the response and
predictors. Interpret each slope while holding the other predictor constant.
Check that you can distinguish the coefficient estimate from its standard error
and residual variation.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(broom)
library(performance)
library(see)

cancer <- read_csv("regression/cancer_data.csv", show_col_types = FALSE)

cancer_model_data <- cancer |>
  select(
    area_name, Mortality_Rate_num, all_poverty, popestimate2015,
    Incidence_Rate_num
  ) |>
  filter(
    is.finite(Mortality_Rate_num),
    is.finite(all_poverty),
    is.finite(popestimate2015),
    is.finite(Incidence_Rate_num),
    popestimate2015 > 0
  ) |>
  mutate(
    poverty_percent = 100 * all_poverty / popestimate2015,
    incidence_rate = Incidence_Rate_num
  ) |>
  select(area_name, Mortality_Rate_num, poverty_percent, incidence_rate)

glimpse(cancer_model_data)
```

### Demonstration: fit the primary model

```{r}
cancer_model <- lm(
  Mortality_Rate_num ~ poverty_percent + incidence_rate,
  data = cancer_model_data
)

cancer_model
tidy(cancer_model)
```

From the output, name the response and both predictors. Find each slope and
interpret its units while holding the other predictor fixed. Compare each
estimate with its standard error: which slope is estimated more precisely
relative to its size?

The poverty coefficient is the estimated difference in mortality for a
one-percentage-point difference in poverty among counties with the same
incidence rate. The incidence coefficient is interpreted similarly while
holding poverty constant. These are associations, not causal effects.

### Your turn: identify what the coefficients mean

**Your turn:** Identify the response variable and both predictors. Interpret each
slope in the units of the problem, remembering to hold the other predictor
constant. Which coefficient appears to be estimated more precisely relative to
its size?

### A quick return to $R^2$ and adjusted $R^2$

**What to do:** Compare ordinary $R^2$ and adjusted $R^2$ for the fitted
model. Recall that ordinary $R^2$ cannot decrease when predictors are added,
even if the new predictor contributes little. Adjusted $R^2$ applies a
penalty for additional model terms, so it can decrease when extra complexity
is not justified by improved fit.

```{r}
glance(cancer_model) |>
  select(r.squared, adj.r.squared)
```

**Your turn:** Why can ordinary $R^2$ increase when a useless predictor is
added, while adjusted $R^2$ can decrease? What would it suggest if $R^2$
increased but adjusted $R^2$ decreased?

**Hint:** Compare what the two measures reward when a model gains another term.

## 2. Confidence intervals and coefficient p-values

**What to do:** Use the coefficient table to interpret the 95% confidence
interval and p-value for each coefficient. State the null hypothesis being
tested, decide whether the data provide evidence against a zero coefficient,
and keep statistical evidence separate from effect size and causality.

Last week we used standard errors and p-values to describe uncertainty and
evidence for adjusted coefficients. We can now extend that idea by attaching a
confidence interval to each coefficient.

```{r}
tidy(cancer_model, conf.int = TRUE)
```

The table contains a two-sided t test for each coefficient. The rows correspond
to these null hypotheses:

| Row | Null hypothesis | What a nonzero coefficient represents |
|:--|:--|:--|
| (Intercept) | $H_0:\beta_0=0$ | fitted mortality when both predictors are zero |
| poverty_percent | $H_0:\beta_{\mathrm{poverty}}=0$ | change per poverty percentage point, at fixed incidence |
| incidence_rate | $H_0:\beta_{\mathrm{incidence}}=0$ | change per incidence-rate unit, at fixed poverty |

A p-value is calculated under the stated null and model. It is not the
probability that the null is true. Use reject or fail to reject; never say that
a large p-value proves a null.

### Demonstration: test the poverty slope

```{r}
poverty_row <- tidy(cancer_model, conf.int = TRUE) |>
  filter(term == "poverty_percent")
poverty_row
```

For this test, $H_0:\beta_{\mathrm{poverty}}=0$ and
$H_A:\beta_{\mathrm{poverty}}\ne0$. A small p-value is evidence against a
zero adjusted poverty slope. The confidence interval gives the range of
coefficient values compatible with the model and sample under repeated-sampling
reasoning. Statistical evidence does not by itself establish a causal effect
or practical importance.

### Your turn: test the incidence slope and the intercept

**Your turn:** Interpret the incidence coefficient, its 95% confidence interval
and p-value. State the null hypothesis and your conclusion. Then do the same
for the intercept. Is the intercept test statistically significant, and is it
substantively useful?

**Hint:** For each term, compare the confidence interval with zero and read the
p-value against the 5% level. Then check whether the predictor values in that
interpretation occur in the observed data.

## 3. Confidence versus prediction intervals

**What to do:** Make a prediction for the specified county characteristics
using both a confidence interval and a prediction interval. Compare their
widths and explain why the prediction interval for one county is wider than the
confidence interval for the average response.

```{r}
new_county <- tibble(
  poverty_percent = 15,
  incidence_rate = 60
)
```

### Demonstration: confidence interval for an average

```{r}
confidence_interval <- as_tibble(
  predict(cancer_model, newdata = new_county, interval = "confidence")
) |>
  mutate(interval = "Confidence: average county")

confidence_interval |>
  select(interval, fit, lwr, upr)
```

A confidence interval describes uncertainty about the average mortality for all
counties with these predictor values.

### Your turn: prediction interval for one county

**Your turn:** Calculate a prediction interval for one county with these same
predictor values. Compare it with the confidence interval above. Which is
wider, and why?

```{r}
# Your code here
```

## 4. Interpret the LINE diagnostics

**What to do:** Work through the LINE assumptions one at a time. For each
assumption, state what it means, identify what evidence the supplied diagnostic
provides, and decide whether there is a meaningful concern. Remember that
violations of assumptions happen in degrees; they are not an automatic
pass/fail rule.

### Demonstration: L: what linearity asks

The linearity plot shows the model residual pattern against fitted values. A
roughly horizontal reference line supports a straight-line mean
structure; a visible curve suggests that the conditional mean is not adequately
represented by straight-line terms.

```{r}
check_model(cancer_model, check = "linearity")
```

### Your turn: interpret the linearity plot

**Your turn:** Does the reference line show clear curvature, or is a
straight-line mean structure reasonably supported?

**Hint:** Look for a systematic curve in the residual trend, rather than
isolated points far from the line.

### Demonstration: I: what independence asks

There is no valid row-order plot for independence. Counties may share
geography, services, environmental exposures or regional smoking patterns. So
this one is handled by study design and knowing your data and its source.

### Your turn: assess independence from study design

**Your turn:** Can independence be assessed from these diagnostic plots alone?
What features of the study design could make counties non-independent?

**Hint:** Think about nearby counties sharing geography, exposures or services.

### Demonstration: N: what normality asks

`check_normality()` creates what is needed for a residual Q-Q diagnostic. With
a large sample the curve can diverge at the tails.

```{r}
normality_check <- check_normality(cancer_model)
normality_check
plot(normality_check, type = "qq")
```

### Your turn: interpret the normality plot

**Your turn:** How closely do the residuals follow the reference line? Pay
particular attention to the centre and tails.

**Hint:** A Q-Q plot compares the residual quantiles with what a normal
distribution would produce; check whether departures are mainly in the tails.

### Demonstration: E: what equal variance asks

`check_heteroscedasticity()` is the same plot as linearity, that is fitted
values versus residuals.

```{r}
heteroscedasticity_check <- check_heteroscedasticity(cancer_model)
plot(heteroscedasticity_check)
```

### Your turn: interpret the equal-variance plot

**Your turn:** Does the residual spread stay approximately constant as the
fitted values increase, or does it change systematically?

**Hint:** Look for a widening or narrowing vertical cloud across fitted values.

## 5. Influence and one transformation

**What to do:** Identify observations that may have unusual influence on the
fitted model, then compare the original model with one transformed alternative.
Do not delete observations automatically or choose a transformation using
$R^2$ alone. Consider the residual pattern, influence, interpretability and
purpose of the model.

### Demonstration: screen for influential observations

```{r}
influence_check <- check_outliers(
  cancer_model,
  method = "cook"
)

influence_check
plot(influence_check)

influence_data <- augment(cancer_model, data = cancer_model_data) |>
  mutate(observation = row_number())

influence_data |>
  slice_max(.cooksd, n = 5) |>
  select(observation, area_name, .cooksd, .hat, .std.resid)
```

### Your turn: investigate influence

**Your turn:** Identify the counties with the largest Cook's distances. Which
observations would you investigate further, and what would you check before
deciding whether any observation should be excluded?

**Hint:** Inspect the largest values in the supplied table. Treat a large
Cook's distance as a reason to investigate the source record, not as an
automatic deletion rule.

### Demonstration: compare a log-incidence alternative

```{r}
cancer_log_model <- lm(
  Mortality_Rate_num ~ poverty_percent + log(incidence_rate),
  data = cancer_model_data
)

tibble(
  model = c("Original incidence", "Log incidence"),
  r_squared = c(
    glance(cancer_model)$r.squared,
    glance(cancer_log_model)$r.squared
  ),
  residual_sd = c(
    glance(cancer_model)$sigma,
    glance(cancer_log_model)$sigma
  )
)
```

```{r}
check_model(cancer_model, check = "linearity")
check_model(cancer_log_model, check = "linearity")

plot(check_heteroscedasticity(cancer_model))
plot(check_heteroscedasticity(cancer_log_model))
```

### Your turn: decide about the transformation

**Your turn:** Compare the original and log-incidence models. Look at $R^2$,
residual SD, residual curvature and spread. Does the transformation improve the
model enough to justify making the coefficient harder to interpret?

**Hint:** Compare the supplied fit table and paired diagnostic plots. Base the
decision on the size of the improvement and the model's purpose, not on $R^2$
alone.

## 6. Independent workflow: income and insurance

**Your turn: complete the full workflow using income and insurance.**

1. Keep complete rows for the response, income, uninsured count and population.
   Calculate income in thousands of dollars and uninsured percentage, then fit
   mortality using both predictors.
2. Interpret each adjusted coefficient in its original units. Inspect standard
   errors, 95% intervals and p-values, and compare $R^2$ with adjusted $R^2$.
3. Predict mortality when income is 50 thousand dollars and uninsured
   percentage is 10. Calculate both an interval for the average county and an
   interval for one county; compare their widths.
4. Interpret the supplied LINE plots and influence output. State the main
   concern for each assumption and name any counties you would investigate.
5. Fit a second model using log income. Compare fit and diagnostic plots, then
   recommend which model to retain and explain the trade-off in interpretation.

### Your turn: prepare and fit the independent model

Fit the model and interpret both adjusted slopes, confidence intervals and
p-values. Also state what the intercept represents and whether that comparison
is meaningful for these counties.

Create income_data with complete rows for area_name, Mortality_Rate_num,
med_income, all_without_insur and popestimate2015. Derive:

- income_thousands = med_income / 1000; and
- uninsured_percent = 100 * all_without_insur / popestimate2015.

Then fit and print income_model:

```{r}
# Your code here
```

Use tidy(income_model, conf.int = TRUE) and answer:

1. What is the adjusted mortality difference for a $1,000 income difference?
2. What is the adjusted mortality difference for a one-percentage-point
   uninsured-rate difference?
3. Are the slopes or the intercept plausibly zero according to their tests?
4. Is the intercept at zero income and zero uninsured percentage a meaningful
   county comparison?

Also report ordinary and adjusted $R^2$.

### Your turn: predict with uncertainty

Calculate both intervals and explain why they differ in width. Use a county with
income_thousands = 50 and uninsured_percent = 10, and calculate both a
confidence interval for the average and a prediction interval for one county:

```{r}
# Your code here
```

### Your turn: interpret the independent diagnostics

Work through LINE and influence just as you did for the primary model. Which
assumptions cause the most concern?

For each plot, write what pattern you see, what it suggests about the
assumption, and whether it concerns you. For influence, name any counties in
the largest Cook's-distance rows and say what you would check.

This supplied diagnostics chunk uses `income_model` from your previous task.
After you have created and fitted that model, run the chunk in RStudio. It is
skipped during rendering while the model code is still blank.

```{r}
#| eval: false
check_model(income_model, check = "linearity")

income_normality_check <- check_normality(income_model)
plot(income_normality_check, type = "qq")

income_heteroscedasticity_check <- check_heteroscedasticity(income_model)
plot(income_heteroscedasticity_check)

income_influence_check <- check_outliers(
  income_model,
  method = "cook",
  threshold = list(cook = 4 / nobs(income_model))
)

income_influence_check
plot(income_influence_check)

income_results <- augment(income_model, data = income_data)

income_results |>
  mutate(observation = row_number()) |>
  slice_max(.cooksd, n = 5) |>
  select(observation, area_name, .cooksd, .hat, .std.resid)
```

Write one paragraph covering L, I, N and E, and identify any influential
observations that require checking. Do not claim that a plot proves independence.

### Your turn: compare a transformed independent model

Compare the original and log-income models. Decide whether the transformation
produces a worthwhile improvement in fit or diagnostics.

Fit a second model with log(income_thousands) in place of
income_thousands. Compare R-squared, residual standard deviation and the
performance diagnostic plots. Finish with a recommendation that distinguishes
association from causation and qualifies inference when LINE concerns remain.

```{r}
# Your code here: fit income_log_model and compare fit.
# Use check_model() for linearity and plot(check_heteroscedasticity())
# for both the original and transformed models.
```

## Conclusion

Summarise the primary model's adjusted associations, what each p-value tests,
the confidence-versus-prediction distinction, the most important diagnostic
concern and whether the transformation is worth retaining. Explain what extra
data or study design would make the independence assumption more credible.

[View the worked solutions online](https://knowlton.co.nz/297101/labs/Workshop_Week12_Assumptions_guide.html).
