---
title: "Workshop Week 10: Simple Regression"
subtitle: "Student workbook: relationships, fitted lines and prediction"
author:
  - name: "Nick Knowlton"
  - name: "297.101 Statistical Data Science"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

For the first part of this course we focused on working with data in R and RStudio,
including using `ggplot2` to explore and visualise relationships. More recently we
have looked at probability distributions, sampling variation and the central limit
theorem, which give us a way to think about uncertainty in estimates from data.

In this workshop we start bringing those ideas together. We will use plots to explore
relationships between vehicle characteristics and fuel use, then fit our first
regression models to describe those relationships numerically. Along the way we will
interpret the fitted line, coefficients, standard errors, p-values, $R^2$, predictions
and residuals.

[Waka Kotahi's vehicle-fleet data](https://www.nzta.govt.nz/resources/new-zealand-motor-vehicle-register-statistics/new-zealand-vehicle-fleet-open-data-sets)
provide a snapshot of vehicles recorded on the Motor Vehicle Register, while
[EECA's guide to vehicle energy-economy labels](https://www.eeca.govt.nz/regulations/vehicle-emissions-and-energy-economy-labelling-programme/about-the-vehicle-emissions-and-energy-economy-labelling-programme/)
explains the standard fuel-economy measure in litres per 100 km. The supplied
extract contains 467 unique vehicle specifications, prepared from Waka Kotahi's
vehicle-fleet data as at **31 July 2026** and containing model-year 2025 petrol
passenger cars and vans with complete measurements.

## Today's question

How much can one vehicle specification tell us about fuel use?

## Learning objectives

By the end of this workshop you should be able to:

- recognise an approximately linear association from a scatterplot;
- fit and interpret a simple linear regression model;
- interpret the slope, intercept, standard error, p-value and $R^2$;
- calculate and interpret a fitted value; and
- define and interpret a residual.

## How to use this workbook

Work through the sections in order. Run the supplied demonstrations and look at
their output before moving on. In each **Your turn** section, write your own
code in the blank chunk, then answer the question in words. Hints appear when a
function or idea is first introduced; later tasks expect you to reuse it.
Open this file from the `labs` folder so the supplied data file can be found.

## 1. Explore engine size and power

**What to do:** Start by looking at the data before fitting a model. For each
scatterplot, describe the direction, strength and shape of the relationship and
look for anything unusual. Then calculate the correlation and ask whether a
straight line looks like a reasonable summary.

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(broom)

vehicles <- read_csv("nz_vehicles_2025.csv", show_col_types = FALSE)
glimpse(vehicles)
```

Each row is one unique 2025 petrol vehicle specification. Fuel use is the
combined-cycle test value in litres per 100 km, so the plots describe
associations among specifications rather than causal effects or real-world
consumption.

### Demonstration: engine size

Let's start with the flexible smoother we learned earlier in the course to get a feel
for the relationship between engine size and fuel use. A reminder that this
lets us look at the *shape* of the relationship first and decide whether a linear
model is a reasonable summary.

```{r}
ggplot(vehicles, aes(engine_l, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Do larger engines tend to use more fuel?"
  ) +
  theme_minimal(base_size = 13)

vehicles |>
  summarise(engine_correlation = cor(engine_l, fuel_use_l_per_100km))
```

Look at the scatterplot first. Describe the direction, shape and strength of
the relationship, and note anything unusual. Then use the correlation to
summarise its linear strength. Does that number support what you saw? Remember
that correlation has no units and is not a slope.

### Your turn: engine power

Repeat the same process for engine power. Before looking at the correlation, use
the plot to decide whether the relationship is positive or negative, whether it
looks approximately linear, and whether it appears stronger or weaker than the
engine-size relationship.

**Hint:** Use `ggplot()` with a LOESS smooth, then calculate the correlation
with `cor()`.

```{r}
# Your code here
```

Which relationship appears more nearly linear? Which has the stronger linear
association according to correlation? Use the plot and correlation as evidence;
do not interpret either correlation as a slope.

## 2. Fit simple linear models

**What to do:** Fit a simple linear regression model with fuel use as the
response and engine size as the explanatory variable. Identify the intercept and
slope, then add the fitted regression line to the scatterplot. Repeat the same
process using engine power.

The relationship looks approximately linear, so we can now summarise it using
a straight-line model.

A simple linear model describes the average response as a straight-line
function of one explanatory variable:

$$
\widehat{y} = \widehat{\beta}_0 + \widehat{\beta}_1 x.
$$

```{r}
engine_model <- lm(
  fuel_use_l_per_100km ~ engine_l,
  data = vehicles
)

engine_model
```

```{r}
ggplot(vehicles, aes(engine_l, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "A straight-line model for engine size and fuel use"
  ) +
  theme_minimal(base_size = 13)
```

There are many possible straight lines we could draw through these data.
`lm()` uses least squares: it chooses the intercept and slope that minimise the
sum of the squared vertical distances between the observed values and the fitted
line. This is the same idea as minimising the mean squared error we looked at
previously.

What is `lm()` changing when it searches for the best-fitting line?

The line captures the increasing average trend, but individual vehicles remain
above and below it. It is a summary of the conditional mean, not an exact rule.

From the printed model, identify the intercept and slope. On the plot, check
whether the fitted line follows the main pattern in the points. What does the
line represent for vehicles with the same engine size?

### Your turn: power model

Fit the power model yourself before looking at the answer. Compare its fitted
line with the LOESS smooth from Section 1. Does a straight line still seem like
a reasonable summary?

**Hint:** Fit fuel use as the response and `power_kw` as the explanatory
variable with `lm()`. Add a straight fitted line to a scatterplot with
`geom_smooth(method = "lm")`.

```{r}
# Your code here
```

```{r}
# Your code here
```

## 3. Coefficients and R²

**What to do:** Use `tidy()` to examine the fitted coefficients. For each
coefficient, identify the estimate, standard error and p-value. Interpret the
slope in the units of the problem rather than just reporting the number. Then
use $R^2$ to describe how much of the observed variation in fuel use is
accounted for by the model.

```{r}
engine_coefficients <- tidy(engine_model)
engine_coefficients
```

Write the fitted equation. Identify the intercept and slope, then interpret the
slope in L/100 km per litre of engine size. Before giving the intercept a
practical meaning, check whether zero litres is represented in these data.

Use the `estimate`, `std.error`, `statistic`, and `p.value` columns to explain
what each part of the coefficient table tells us. A small p-value is evidence
against a zero population coefficient under the model; it does not by itself
show that an association is large, important or causal.

We already calculated the correlation between engine size and fuel use. Square
it and compare it with the model $R^2$. What do you notice? Write one sentence
about how much variation the engine-size model explains.

```{r}
# Your code here: square the correlation and compare it with model R^2
```

**Hint:** Square the engine-size correlation and compare it with
`glance(engine_model)$r.squared`.

### Power coefficients and fit

Now interpret the power model yourself. What does the slope mean in L/100 km
per kW? Is that a useful scale, or would reporting the change per 100 kW be
easier to understand? Compare its standard error, p-value and $R^2$ with the
engine-size model. State which model accounts for more of the observed
variation. Since the predictors use different units, do not compare their raw
slope standard errors as if they were on the same scale.

```{r}
# Your code here
```

## 4. Make a prediction

**What to do:** Use the fitted equation to calculate the fitted fuel use for a
vehicle with a 2.0 L engine by hand. Then use `predict()` to obtain the same
value in R. The two answers should agree. Explain what the fitted value
represents and whether it describes every individual vehicle with a 2.0 L
engine.

**Hint:** Use the intercept and engine-size slope from `coef(engine_model)`;
for the R prediction, supply a one-row tibble with `engine_l = 2`.

```{r}
# Your code here
```

What does your fitted value represent? Does the model say that every 2.0 L
vehicle will have this fuel use?

## 5. What is a residual?

For one vehicle, the residual is observed fuel use minus fitted fuel use:

$$
e_i = y_i-\hat{y}_i
$$

A positive residual means the observed fuel use was above the fitted value; a
negative residual means it was below.

**What to do:** Use `augment()` to compare the observed fuel use, fitted value
and residual for several vehicles. Check that

$$
\text{residual}=\text{observed}-\text{fitted}.
$$

Find one positive residual and one negative residual and explain what each means
in the context of fuel use.

**Hint:** In the augmented results, the observed response is
`fuel_use_l_per_100km`, the fitted value is `.fitted`, and the residual is
`.resid`.

```{r}
# Your code here
```

On the plot, the residual is the vertical distance from each observed value to
the fitted line. Before looking at the residual column, use the graph to decide
whether a point should have a positive or negative residual.

Use the plot to find one point above the fitted line and one below it. Predict
the sign of each residual from the graph, then check those signs in the table.
What are the units of a residual?

**Hint:** The residual is the observed value minus the fitted value.

```{r}
# Your code here
```

A near-zero average residual is expected from least squares, but it does not by
itself establish that the model is adequate. Formal residual diagnostics belong
to Week 12.

## 6. Independent workflow: gross vehicle mass

**Your turn: complete the full workflow using gross vehicle mass.**

1. Plot gross mass against fuel use with a smooth. Describe the direction,
   strength and shape, and calculate the correlation.
2. Fit and print a simple model called `mass_model`. Interpret its slope per
   1000 kg, then inspect the slope's standard error and p-value.
3. Report $R^2$ and explain how much variation the model accounts for.
4. Predict the fitted mean fuel use at 2000 kg.
5. Compare this model's $R^2$ with the engine-size and power models. Which
   single-predictor model accounts for the most variation in this dataset?

```{r}
# Your code here
```

## Looking ahead

Next week we return to these models and add body type as another predictor. This
introduces multiple regression, reference categories and adjusted coefficients.
The coefficient estimates, standard errors and p-values are interpreted in the
same general way as today, but the meaning of each coefficient changes because
it is now conditional on the other variables in the model.

[View the worked solutions online](https://knowlton.co.nz/297101/labs/Workshop_Week10_Relationships_guide.html).
