---
title: "Workshop Week 10: Simple Regression"
subtitle: "Answer guide: relationships, fitted lines and prediction"
author:
  - name: "Nick Knowlton"
  - name: "297.101 Statistical Data Science"
date-meta: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

This guide follows [`Workshop_Week10_Relationships.qmd`](Workshop_Week10_Relationships.qmd).
It contains the completed code and interpretations for the workbook.

For the first part of this course we focused on working with data in R and RStudio,
including using `ggplot2` to explore and visualise relationships. More recently we
have looked at probability distributions, sampling variation and the central limit
theorem, which give us a way to think about uncertainty in estimates from data.

In this workshop we start bringing those ideas together. We will use plots to explore
relationships between vehicle characteristics and fuel use, then fit our first regression
models to describe those relationships numerically. Along the way we will interpret
the fitted line, coefficients, standard errors, p-values, $R^2$, predictions and residuals.

[Waka Kotahi's vehicle-fleet data](https://www.nzta.govt.nz/resources/new-zealand-motor-vehicle-register-statistics/new-zealand-vehicle-fleet-open-data-sets)
provide a snapshot of vehicles recorded on the Motor Vehicle Register, while
[EECA's guide to vehicle energy-economy labels](https://www.eeca.govt.nz/regulations/vehicle-emissions-and-energy-economy-labelling-programme/about-the-vehicle-emissions-and-energy-economy-labelling-programme/)
explains the standard fuel-economy measure in litres per 100 km. The supplied
extract contains 467 unique vehicle specifications, prepared from Waka Kotahi's
vehicle-fleet data as at **31 July 2026** and containing model-year 2025 petrol
passenger cars and vans with complete measurements.

## Today's question

How much can one vehicle specification tell us about fuel use?

## Learning objectives

By the end of this workshop you should be able to:

- recognise an approximately linear association from a scatterplot;
- fit and interpret a simple linear regression model;
- interpret the slope, intercept, standard error, p-value and $R^2$;
- calculate and interpret a fitted value; and
- define and interpret a residual.

## 1. Explore engine size and power

**What to do:** Start by looking at the data before fitting a model. For each
scatterplot, describe the direction, strength and shape of the relationship and
look for anything unusual. Then calculate the correlation and ask whether a
straight line looks like a reasonable summary.

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(broom)

vehicles <- read_csv("nz_vehicles_2025.csv", show_col_types = FALSE)
glimpse(vehicles)
```

Each row is one unique 2025 petrol vehicle specification. Fuel use is the
combined-cycle test value in litres per 100 km, so the plots describe
associations among specifications rather than causal effects or real-world
consumption.

### Demonstration: engine size

Let's start with the flexible smoother we learned earlier in the course to get a feel
for the relationship between engine size and fuel use. A reminder that this
lets us look at the *shape* of the relationship first and decide whether a linear
model is a reasonable summary.

```{r}
ggplot(vehicles, aes(engine_l, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Do larger engines tend to use more fuel?"
  ) +
  theme_minimal(base_size = 13)

vehicles |>
  summarise(engine_correlation = cor(engine_l, fuel_use_l_per_100km))
```

What does the correlation add to the plot? Remember that it describes linear
strength, has no units, and is not a slope.

The relationship is positive, strong and approximately linear. The correlation
is about 0.90, so larger engines tend to be associated with higher fuel use,
although there is still visible vehicle-to-vehicle variation. Correlation is
unit-free and measures linear strength; it is not the slope and does not imply
that changing engine size causes the difference. Correlation only summarises
linear association, so always look at the scatterplot as well. Non-linearity,
unusual observations or other structure can be hidden by a single correlation
value.

### Your turn: engine power

Repeat the same process for engine power. Before looking at the correlation, use
the plot to decide whether the relationship is positive or negative, whether it
looks approximately linear, and whether it appears stronger or weaker than the
engine-size relationship.

```{r}
ggplot(vehicles, aes(power_kw, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Engine power (kW)",
    y = "Combined fuel use (L/100 km)",
    title = "Do more powerful vehicles tend to use more fuel?"
  ) +
  theme_minimal(base_size = 13)

vehicles |>
  summarise(power_correlation = cor(power_kw, fuel_use_l_per_100km))
```

Power also has a positive relationship with fuel use. Its correlation is about
0.884, slightly smaller than the engine-size correlation in this dataset. Again,
the correlation should be interpreted alongside the scatterplot because it only
summarises the linear part of the relationship.

Both relationships are positive and approximately linear. The engine-size
relationship appears slightly more nearly linear in this dataset, and it has the
larger correlation (0.901 versus 0.884). These correlations measure unit-free
linear strength; they are not slopes and should not be compared as changes in
fuel use per unit of the predictor.

Which relationship appears more nearly linear? Which has the larger correlation?
Do not compare the correlations as if they were slopes.

## 2. Fit simple linear models

**What to do:** Fit a simple linear regression model with fuel use as the
response and engine size as the explanatory variable. Identify the intercept and
slope, then add the fitted regression line to the scatterplot. Repeat the same
process using engine power.

The relationship looks approximately linear, so we can now summarise it using
a straight-line model.

A simple linear model describes the average response as a straight-line
function of one explanatory variable:

$$
\widehat{y} = \widehat{\beta}_0 + \widehat{\beta}_1 x.
$$

```{r}
engine_model <- lm(
  fuel_use_l_per_100km ~ engine_l,
  data = vehicles
)

engine_model
```

```{r}
ggplot(vehicles, aes(engine_l, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "A straight-line model for engine size and fuel use"
  ) +
  theme_minimal(base_size = 13)
```

There are many possible straight lines we could draw through these data.
`lm()` uses least squares: it chooses the intercept and slope that minimise the
sum of the squared vertical distances between the observed values and the fitted
line. This is the same idea as minimising the mean squared error we looked at
previously.

What is `lm()` changing when it searches for the best-fitting line?

It changes the intercept and slope. The least-squares search chooses the values
that minimise the sum of the squared residuals, so the resulting line is the
closest straight-line summary under that criterion.

The line captures the increasing average trend, but individual vehicles remain
above and below it. It is a summary of the conditional mean, not an exact rule.

### Your turn: power model

Fit the power model yourself before looking at the answer. Compare its fitted
line with the LOESS smooth from Section 1. Does a straight line still seem like
a reasonable summary?

```{r}
power_model <- lm(
  fuel_use_l_per_100km ~ power_kw,
  data = vehicles
)

power_model
```

```{r}
ggplot(vehicles, aes(power_kw, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Engine power (kW)",
    y = "Combined fuel use (L/100 km)",
    title = "A straight-line model for engine power and fuel use"
  ) +
  theme_minimal(base_size = 13)
```

Both lines capture a positive average association and leave meaningful
vehicle-specific variation unexplained. The straight line is a reasonable
summary for the power relationship, although it is not an exact rule for every
vehicle.

## 3. Coefficients and R²

**What to do:** Use `tidy()` to examine the fitted coefficients. For each
coefficient, identify the estimate, standard error and p-value. Interpret the
slope in the units of the problem rather than just reporting the number. Then
use $R^2$ to describe how much of the observed variation in fuel use is
accounted for by the model.

```{r}
engine_coefficients <- tidy(engine_model)
engine_coefficients
```

The `estimate` column gives the fitted coefficients from our sample. For engine
size, the estimated slope is 1.886 L/100 km per litre, so a one-litre difference
in engine size is associated with an average difference of about 1.89 L/100 km
in fuel use.

The `std.error` describes the sampling uncertainty in each coefficient estimate.
Smaller standard errors mean the coefficient has been estimated more precisely.

The `statistic` is the coefficient estimate divided by its standard error. The
`p.value` asks how compatible the observed estimate is with a null hypothesis
that the corresponding population coefficient is zero, assuming the regression
model and its assumptions are appropriate.

For engine size the p-value is extremely small, providing strong evidence that
the population slope is not zero. The estimated effect size is still given by
the coefficient itself. A small p-value does not tell us that an association is
large, important or causal.

The fitted equation is approximately

$$
\widehat{\text{fuel use}} = 4.070 + 1.886(\text{engine size})
$$

Write the fitted equation. Which number is the intercept, and which number is
the slope?

A one-litre difference in engine size is associated with about 1.886 L/100 km
higher average fuel use. Zero litres is outside the observed range (0.998 to
6.749 L), so the intercept is a mathematical line value rather than a practical
vehicle prediction. Notice that the intercept also has an extremely small
p-value. That **does not** make it practically useful: a zero-litre engine is
outside the range of these data. Statistical evidence and practical
interpretation are different questions!

We already calculated the correlation between engine size and fuel use. Square
it and compare it with the model $R^2$. What do you notice? Write one sentence
about how much variation the engine-size model explains.

```{r}
engine_r <- cor(vehicles$engine_l, vehicles$fuel_use_l_per_100km)
tibble(
  correlation = engine_r,
  correlation_squared = engine_r^2,
  r_squared = glance(engine_model)$r.squared,
  difference = correlation_squared - r_squared
)
```

The correlation is about 0.901 and the simple-model R-squared is about 0.812.
Thus the engine-size model accounts for roughly 81% of the observed variation
in fuel use in this dataset. For simple linear regression with an intercept,
$R^2 = r^2$. Here $0.901^2\approx0.812$, which is exactly the $R^2$ from
the fitted model. This equivalence holds for simple linear regression with an
intercept. Once we move to multiple regression next week, there is no single
correlation that we can simply square to obtain the model $R^2$.

### Power coefficients and fit

```{r}
power_coefficients <- tidy(power_model)
power_fit <- glance(power_model)

power_coefficients
power_fit |>
  select(r.squared)
```

Now interpret the power model yourself. What does the slope mean in L/100 km
per kW? Is that a useful scale, or would reporting the change per 100 kW be
easier to understand? Compare its standard error, p-value and $R^2$ with the
engine-size model.

Read this output in the same way. The power coefficient is about 0.0186 L/100
km per kW, with a very small standard error and p-value. The coefficient tells
us the size of the estimated association; the standard error tells us its
precision; and the p-value provides evidence about whether the population slope
could reasonably be zero under the model.

The power slope is about 0.0186 L/100 km per kW, or about 1.86 L/100 km per
100 kW. The fitted equation is approximately

$$
\widehat{\text{fuel use}} = 4.705 + 0.0186(\text{power}) .
$$

The power slope standard error is about 0.000456 L/100 km per kW and its
p-value is extremely small. The engine-size slope standard error is about
0.0421 L/100 km per litre and its p-value is also extremely small. Because the
predictors use different units, the raw standard errors should not be compared
as if they measured the same quantity; both slopes are estimated precisely
relative to their scales. Power has an R-squared of about 0.782, compared with
about 0.812 for engine size. In these data, the engine-size model therefore
accounts for slightly more of the observed variation in fuel use.

## 4. Make a prediction

**What to do:** Use the fitted equation to calculate the fitted fuel use for a
vehicle with a 2.0 L engine by hand. Then use `predict()` to obtain the same
value in R. The two answers should agree.

```{r}
engine_prediction_manual <- coef(engine_model)[["(Intercept)"]] +
  coef(engine_model)[["engine_l"]] * 2

engine_prediction_r <- predict(
  engine_model,
  newdata = tibble(engine_l = 2)
)

tibble(
  manual = engine_prediction_manual,
  using_predict = as.numeric(engine_prediction_r),
  difference = manual - using_predict
)
```

The predicted average fuel use for a 2.0 L vehicle is about 7.84 L/100 km. The
manual and `predict()` values agree. This is the fitted mean fuel use for
vehicles with a 2.0 L engine according to this model. Note that individual 2.0 L
vehicles can sit well above or below this value, so it should
not be interpreted as *truth* or a guarantee for an individual vehicle.

What does 7.84 L/100 km represent? Does the model say that every 2.0 L vehicle
will have this fuel use?

It is the model’s estimated mean fuel use for vehicles with a 2.0 L engine. It
does not mean every 2.0 L vehicle will use 7.84 L/100 km; individual vehicles
can be above or below this fitted value.

## 5. What is a residual?

A residual is the difference between what we observed and what the model fitted:

$$
e_i = y_i-\hat{y}_i
$$

A positive residual means the observed fuel use was above the fitted value; a
negative residual means it was below.

**What to do:** Use `augment()` to compare the observed fuel use, fitted value
and residual for several vehicles. Check that

$$
\text{residual}=\text{observed}-\text{fitted}.
$$

Find one positive residual and one negative residual and explain what each means
in the context of fuel use.

```{r}
engine_results <- augment(engine_model)
engine_results |>
  select(engine_l, fuel_use_l_per_100km, .fitted, .resid) |>
  slice_head(n = 8)
```

On the plot, the residual is the vertical distance from each observed value to
the fitted line. Before looking at the residual column, use the graph to decide
whether a point should have a positive or negative residual.

```{r}
set.seed(297101)
engine_gap_data <- engine_results |>
  slice_sample(n = min(30, nrow(engine_results)))

ggplot(vehicles, aes(engine_l, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.35) +
  geom_line(
    data = engine_results |> arrange(engine_l),
    aes(y = .fitted),
    colour = "#2C7FB8",
    linewidth = 1
  ) +
  geom_segment(
    data = engine_gap_data,
    aes(xend = engine_l, yend = .fitted),
    colour = "#D95F02",
    alpha = 0.65
  ) +
  labs(
    x = "Engine size (litres)",
    y = "Combined fuel use (L/100 km)",
    title = "Observed values and their fitted values"
  ) +
  theme_minimal(base_size = 13)
```

In the displayed table, the first vehicle has a residual of about -1.73 L/100
km: it uses less fuel than the fitted value for its engine size. The fourth
vehicle has a residual of about +0.62 L/100 km: it uses more fuel than the
fitted value. The sign comes from observed minus fitted, not from the size of
the engine itself.

What are the units of a residual? A near-zero average residual is expected from
least squares, but it does not by itself establish that the model is adequate.
Formal residual diagnostics belong to Week 12.

Each residual is in L/100 km. A point above the line has a positive residual;
one below has a negative residual. The residual is vertical because the model
predicts the response at the observed explanatory value. When an intercept is
included in the model, least squares makes the mean residual essentially zero,
but that algebraic fact does not validate the model.

## 6. Independent workflow: gross vehicle mass

**Your turn: complete the full workflow using gross vehicle mass.**

1. Plot gross mass against fuel use and use a smooth to examine the shape of the relationship.
2. Describe the direction, strength and approximate form of the relationship.
3. Calculate the correlation.
4. Fit a simple linear regression model and save it as `mass_model`.
5. Interpret the slope. Report it per 1000 kg rather than per kilogram if that makes the effect easier to understand.
6. Examine the standard error and p-value for the slope.
7. Report and interpret $R^2$.
8. Predict the fitted mean fuel use for a 2000 kg vehicle.
9. Compare the gross-mass model with the engine-size and power models. Which has the largest in-sample $R^2$?

```{r}
ggplot(vehicles, aes(gross_mass_kg, fuel_use_l_per_100km)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Gross mass (kg)",
    y = "Combined fuel use (L/100 km)",
    title = "Gross mass and fuel use"
  ) +
  theme_minimal(base_size = 13)

mass_correlation <- cor(vehicles$gross_mass_kg, vehicles$fuel_use_l_per_100km)
mass_model <- lm(
  fuel_use_l_per_100km ~ gross_mass_kg,
  data = vehicles
)
mass_coefficients <- tidy(mass_model)
mass_fit <- glance(mass_model)
mass_prediction <- predict(
  mass_model,
  newdata = tibble(gross_mass_kg = 2000)
)

mass_slope_row <- mass_coefficients |>
  filter(term == "gross_mass_kg")

tibble(
  correlation = mass_correlation,
  r_squared = mass_fit$r.squared,
  slope_per_1000_kg = 1000 * mass_slope_row$estimate,
  std_error_per_1000_kg = 1000 * mass_slope_row$std.error,
  p_value = mass_slope_row$p.value,
  predicted_fuel_use_at_2000_kg = as.numeric(mass_prediction)
)
```

Gross mass has a positive association with fuel use, but its single-predictor
R-squared is only about 0.331. The slope is about 3.29 L/100 km per 1000 kg,
with a small standard error on that scale and a very small p-value, so the data
provide strong evidence of a positive linear association under this model. The
model predicts about 8.25 L/100 km at 2000 kg. Of the three single-predictor
models considered here, engine size has the highest in-sample R-squared. A
larger R-squared does not make a relationship causal. It only tells us that the
fitted model accounts for a larger proportion of the observed variation in the
response in this dataset.

Taken together, one numerical vehicle specification can explain a substantial
part of the variation in fuel use, but none of these single-predictor models
predicts every vehicle exactly. Engine size is the strongest of the three
single-predictor summaries here by in-sample $R^2$, followed by power and then
gross mass.


## Looking ahead

Next week we return to these models and add body type as another predictor. This
introduces multiple regression, reference categories and adjusted coefficients.
The coefficient estimates, standard errors and p-values are interpreted in the
same general way as today, but the meaning of each coefficient changes because
it is now conditional on the other variables in the model.
