---
title: "Workshop Week 12"
subtitle: "Answer guide — regression assumptions and diagnostic plots"
author: "297.101 Statistical Data Science"
date: "Spring 2026"
format:
  html:
    toc: true
    code-tools: true
    code-copy: true
    code-block-bg: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
resources:
  - regression/cancer_data.csv.gz
---

Week 11 found county-level associations between poverty, median income and
lung-cancer mortality. Those fitted lines were useful summaries, but they did
not tell us whether a straight line represents the pattern well, whether its
uncertainty calculations are dependable, or whether a few unusual counties
strongly affect the result.

The data come from a broader project that combined county socioeconomic and
cancer information. Noah Rippner's
[original data-preparation code](https://github.com/nrippner/can_clin_trails_prep_r/blob/master/extract_zips_10.py)
shows how data from different sources were cleaned and joined by place. Those
joins let us examine the same county's socioeconomic conditions and mortality,
although the resulting associations do not establish causes or describe
individual people.

::: {.callout-note}
## Today's question

How well can we trust our models' estimates and predictions of county
lung-cancer mortality?
:::

This guide shows how prediction bands, residual patterns and influential
counties affect the Week 11 conclusions, then compares straight-line and
log-predictor models. Each diagnostic answers a different question about the
support for inference and prediction.

## 1. What are we comparing?

::: {.callout-important title="Question"}
Are the poverty and income models fitted to the same counties and expressed in
units that allow a fair comparison?
:::

::: {.callout-note title="What to look for"}
Check that the three row counts agree. Both models must use the same counties
and response for their fit and diagnostics to be comparable.
:::

We rebuild both Week 11 models and check their observation counts.

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(broom)

cancer <- read_csv("regression/cancer_data.csv.gz", show_col_types = FALSE)
cancer_clean <- cancer |>
  select(Mortality_Rate_num, all_poverty, med_income, popestimate2015) |>
  filter(if_all(everything(), is.finite), popestimate2015 > 0) |>
  mutate(
    pov_norm = all_poverty / popestimate2015 * 100000,
    poverty_percent = pov_norm / 1000
  )
poverty_model <- lm(Mortality_Rate_num ~ pov_norm, data = cancer_clean)
income_model <- lm(Mortality_Rate_num ~ med_income, data = cancer_clean)
poverty_results <- augment(poverty_model)
income_results <- augment(income_model)
tibble(
  rows_retained = nrow(cancer_clean),
  poverty_model_rows = nobs(poverty_model),
  income_model_rows = nobs(income_model)
)
```

::: {.callout-tip title="Worked interpretation"}
All counts are 2,809. The poverty and income models therefore use identical
counties, and both predict mortality in deaths per 100,000. Their predictor
units differ: poverty is people per 100,000 and income is US dollars.
:::

## 2. How precise are predictions for another county?

::: {.callout-important title="Question"}
How closely can either model predict mortality for an individual additional
county, and what does observed coverage tell us about future counties?
:::

A prediction band shows intervals for individual additional counties across a
range of predictor values. Its width includes the county-to-county variation
left around the fitted mean.

### Poverty-rate prediction band

::: {.callout-note title="What to look for"}
The line estimates mean mortality at each poverty value. Look at the vertical
scatter around it, then compare the band's width near the middle of the poverty
range with its width at sparse extremes.
:::

```{r}
#| label: poverty-prediction-band-data
poverty_grid <- tibble(
  pov_norm = seq(min(cancer_clean$pov_norm),
                 max(cancer_clean$pov_norm), length.out = 200)
)
poverty_band <- bind_cols(
  poverty_grid,
  as_tibble(predict(
    poverty_model, poverty_grid,
    interval = "prediction", level = 0.95
  ))
)
stopifnot(all(poverty_band$lwr <= poverty_band$fit))
stopifnot(all(poverty_band$fit <= poverty_band$upr))
```

```{r}
#| label: poverty-prediction-band
#| fig-cap: "Fitted mortality and a 95% prediction band across the observed poverty range."
#| fig-alt: "Poverty and mortality scatterplot with a fitted line and shaded prediction band."
#| out-width: "100%"
ggplot(cancer_clean, aes(x = pov_norm, y = Mortality_Rate_num)) +
  geom_ribbon(
    data = poverty_band, aes(x = pov_norm, ymin = lwr, ymax = upr),
    inherit.aes = FALSE, alpha = 0.2
  ) +
  geom_line(data = poverty_band, aes(y = fit), linewidth = 1) +
  geom_point(alpha = 0.25, size = 0.7) +
  labs(x = "Poverty (people per 100,000)",
       y = "Mortality (deaths per 100,000)",
       title = "Individual county outcomes vary widely around the fitted line") +
  theme_minimal(base_size = 13)
```

::: {.callout-tip title="Worked interpretation"}
The fitted line estimates mean mortality for counties at each poverty rate.
The band is much wider because it also includes remaining county-to-county
variation. It is narrowest near the mean poverty value and widens toward the
edges; the broad scatter shows that poverty alone does not predict an
individual county precisely.
:::

::: {.callout-important title="Question"}
What percentage of the fitting observations falls inside the nominal 95%
prediction intervals, and why is that not validation on future counties?
:::

::: {.callout-note title="What to look for"}
The coverage calculation counts observations that were used to fit the model.
Compare the observed percentage with 95%, while remembering that this is an
in-sample description rather than a test on new counties.
:::

```{r}
#| label: poverty-observed-coverage
poverty_observed_intervals <- bind_cols(
  cancer_clean,
  as_tibble(predict(poverty_model, interval = "prediction"))
)
poverty_observed_intervals |>
  summarise(
    observations_inside = sum(between(Mortality_Rate_num, lwr, upr)),
    observations = n(),
    observed_percentage = mean(between(Mortality_Rate_num, lwr, upr)) * 100
  )
```

::: {.callout-tip title="Worked interpretation"}
About 94.66% of the fitting observations lie inside their nominal 95%
prediction intervals. The same data estimated the line and its spread, so this
does not validate coverage for future counties. New-county coverage could be
poorer if the model form, variance or independence assumptions are unsuitable.
:::

### Income prediction band

::: {.callout-important title="Question"}
Where is the income prediction band narrowest, and does its width suggest
precise prediction for an individual county?
:::

::: {.callout-note title="What to look for"}
Compare the width of the income band near typical incomes with its width at
sparse extremes. Keep the fitted line's mean response distinct from an
individual county prediction.
:::

```{r}
#| label: income-prediction-band-data
income_grid <- tibble(
  med_income = seq(min(cancer_clean$med_income),
                   max(cancer_clean$med_income), length.out = 200)
)
income_band <- bind_cols(
  income_grid,
  as_tibble(predict(
    income_model, income_grid,
    interval = "prediction", level = 0.95
  ))
)
stopifnot(all(income_band$lwr <= income_band$fit))
stopifnot(all(income_band$fit <= income_band$upr))
```

```{r}
#| label: income-prediction-band
#| fig-cap: "Fitted mortality and a 95% prediction band across the observed income range."
#| fig-alt: "Income and mortality scatterplot with a fitted line and shaded prediction band."
#| out-width: "100%"
ggplot(cancer_clean, aes(x = med_income, y = Mortality_Rate_num)) +
  geom_ribbon(
    data = income_band, aes(x = med_income, ymin = lwr, ymax = upr),
    inherit.aes = FALSE, alpha = 0.2
  ) +
  geom_line(data = income_band, aes(y = fit), linewidth = 1) +
  geom_point(alpha = 0.25, size = 0.7) +
  labs(x = "Median household income (US dollars)",
       y = "Mortality (deaths per 100,000)",
       title = "Prediction is most precise near typical incomes") +
  theme_minimal(base_size = 13)
```

```{r}
#| label: income-observed-coverage
income_observed_intervals <- bind_cols(
  cancer_clean,
  as_tibble(predict(income_model, interval = "prediction"))
)
income_observed_intervals |>
  summarise(
    observations_inside = sum(between(Mortality_Rate_num, lwr, upr)),
    observations = n(),
    observed_percentage = mean(between(Mortality_Rate_num, lwr, upr)) * 100
  )
```

::: {.callout-tip title="Worked interpretation"}
The income band is narrowest near mean income and widens towards sparse
extremes, although the large individual-county variation makes the change in
width visually subtle. Its in-sample coverage describes these fitted counties;
it does not establish coverage for new ones.
:::

## 3. What patterns does the straight line miss?

::: {.callout-important title="Question"}
Do the residuals support a straight-line mean with independent, approximately
normal errors and reasonably constant variance?
:::

Use this reference table while discussing each diagnostic. The plots provide
evidence to weigh; none is a mechanical pass/fail certificate.

| Assumption | What it means | Evidence to inspect | Why it matters |
|:--|:--|:--|:--|
| Linearity | The conditional mean is represented adequately by a straight line. | Residuals versus fitted and the smooth. | Curvature can bias fitted means and predictions. |
| Independence | Errors across observations are independent; consider design and spatial dependence. | County sampling information and geographic residual patterns. | Dependence can make standard errors and intervals too optimistic. |
| Normality | Conditional errors are approximately normal; residuals are an imperfect diagnostic. | Centre and both tails of the Q-Q plot. | Tail departures can affect small-sample inference and prediction intervals. |
| Equal variance | Conditional error variance is reasonably constant. | Scale-location spread and smooth. | Changing spread makes interval precision vary across predictors. |

### L — Linearity

::: {.callout-important title="Question"}
Does the fitted line leave a curved pattern in the residuals, and where would
that cause the model to overpredict or underpredict mean mortality?
:::

Residuals versus fitted values should form a shapeless cloud around zero. A
curved smooth suggests that a straight-line mean misses structure.

#### Poverty-rate model

::: {.callout-note title="What to look for"}
Look for a cloud centred around zero and a roughly horizontal smooth. A bend or
region where residuals share a sign indicates that the straight line misses
part of the conditional mean.
:::

```{r}
#| label: poverty-linearity
ggplot(poverty_results, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.3, size = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Fitted mortality", y = "Residual",
       title = "Does the poverty model leave a curved pattern?") +
  theme_minimal(base_size = 13)
```

::: {.callout-tip title="Worked interpretation"}
The smooth is slightly below zero at low fitted values, rises through the
middle and falls below zero at the high end. This curvature means the line does
not represent mean mortality equally well across the range. The sparse high
end makes its departure less certain, but it remains a pattern to investigate.
:::

#### Income model

::: {.callout-important title="Question"}
Is the income model's curvature similar to the poverty model's, and what would
it imply about using a straight-line mean?
:::

::: {.callout-note title="What to look for"}
Compare the smooth's direction and its position relative to zero with the
poverty model.
:::

```{r}
#| label: income-linearity
ggplot(income_results, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.3, size = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Fitted mortality", y = "Residual",
       title = "Does the income model leave a curved pattern?") +
  theme_minimal(base_size = 13)
```

::: {.callout-tip title="Worked interpretation"}
The income residual smooth also departs systematically from zero: it is above
zero at low fitted mortality and below zero through part of the middle. The
straight-line income model therefore misses some structure in mean mortality,
even though its overall association remains clear.
:::

### E — Equal variance

::: {.callout-important title="Question"}
Does residual variation stay reasonably constant across fitted mortality, or
does prediction precision change across the range?
:::

#### Poverty-rate model

::: {.callout-note title="What to look for"}
The scale-location plot shows residual magnitude without its sign. Look for a
level band and flat smooth; the smooth need not be at zero.
:::

```{r}
#| label: poverty-equal-variance
poverty_results |>
  mutate(root_abs_standardized_residual = sqrt(abs(.std.resid))) |>
  ggplot(aes(x = .fitted, y = root_abs_standardized_residual)) +
  geom_point(alpha = 0.3, size = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Fitted mortality", y = "Sqrt(|standardized residual|)",
       title = "Is residual spread stable for the poverty model?") +
  theme_minimal(base_size = 13)
```

::: {.callout-tip title="Worked interpretation"}
The poverty-model scale-location smooth rises with fitted mortality and the
points spread more at the high end. This suggests increasing conditional error
variation, so prediction intervals are likely to be less precise for counties
with high fitted mortality.
:::

#### Income model

::: {.callout-important title="Question"}
Does income residual magnitude rise, fall or form a U-shape, and where would
prediction intervals be least precise?
:::

::: {.callout-note title="What to look for"}
Use residual magnitude without its sign. A non-horizontal smooth indicates
that prediction precision changes across fitted mortality.
:::

```{r}
#| label: income-equal-variance
income_results |>
  mutate(root_abs_standardized_residual = sqrt(abs(.std.resid))) |>
  ggplot(aes(x = .fitted, y = root_abs_standardized_residual)) +
  geom_point(alpha = 0.3, size = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Fitted mortality", y = "Sqrt(|standardized residual|)",
       title = "Is residual spread stable for the income model?") +
  theme_minimal(base_size = 13)
```

::: {.callout-tip title="Worked interpretation"}
The income-model spread changes across fitted values rather than remaining
level. Its smooth shows a U-shaped pattern with a rise at the high end, so a
single constant-variance description does not capture prediction precision
equally well throughout the range.
:::

### N — Normality

::: {.callout-important title="Question"}
How closely do the standardised residuals follow a normal distribution,
especially in the tails?
:::

#### Poverty-rate model

::: {.callout-note title="What to look for"}
Compare theoretical normal quantiles with the ordered standardised residuals.
Inspect the centre and the lower and upper tails separately.
:::

```{r}
#| label: poverty-normality
ggplot(poverty_results, aes(sample = .std.resid)) +
  geom_qq(alpha = 0.35, size = 0.8) +
  geom_qq_line() +
  labs(x = "Theoretical normal quantiles",
       y = "Standardized residual quantiles",
       title = "Poverty-model residuals") +
  theme_minimal(base_size = 13)
```

::: {.callout-tip title="Worked interpretation"}
The centre is fairly close to the reference line, while both tails depart,
especially the upper tail. Central residuals are roughly compatible with
normality, but unusually large residuals occur more often than the normal-error
model predicts, which matters particularly for individual prediction.
:::

#### Income model

::: {.callout-important title="Question"}
How close is the income model's centre to the reference line, and how does each
tail depart?
:::

::: {.callout-note title="What to look for"}
Describe the centre and both tails separately, then connect any departures to
coefficient inference and individual-county prediction.
:::

```{r}
#| label: income-normality
ggplot(income_results, aes(sample = .std.resid)) +
  geom_qq(alpha = 0.35, size = 0.8) +
  geom_qq_line() +
  labs(x = "Theoretical normal quantiles",
       y = "Standardized residual quantiles",
       title = "Income-model residuals") +
  theme_minimal(base_size = 13)
```

::: {.callout-tip title="Worked interpretation"}
The income-model centre follows the reference line more closely than its tails.
Both tails depart, so the normal-error approximation remains imperfect. This
does not mean that income or mortality itself must be normally distributed;
the plot concerns conditional errors around the fitted line.
:::

### I — Independence

::: {.callout-important title="Question"}
Could nearby counties have related errors, and can these data investigate that
possibility?
:::

#### Poverty-rate model

::: {.callout-note title="What to look for"}
Independence depends on the study design and spatial relationships rather than
the ordinary row order. Look for county identifiers, locations or adjacency
information that would let us compare neighbouring residuals.
:::

Nearby counties can share air quality, smoking patterns, health services,
policy and population composition. Assessing this needs county identifiers and
spatial adjacency or coordinates.

::: {.callout-tip title="Worked interpretation"}
The data supplied here cannot establish independence. Neighbouring counties may
share unmeasured conditions and therefore related errors, while CSV row order
does not encode geographic distance. Without spatial information, standard
errors and intervals should be interpreted cautiously.
:::

#### Income model

Changing the predictor does not change how county observations were sampled
or their geography. Possible spatial dependence therefore remains.

## 4. Do a few counties strongly affect the fitted line?

::: {.callout-important title="Question"}
Do a small number of counties have enough influence to change the fitted
association appreciably?
:::

Influence is separate from independence. Cook's distance measures how strongly
one observation affects the fitted coefficients through its predictor value
and residual.

### Poverty-rate model

::: {.callout-note title="What to look for"}
Cook's distance combines an unusual predictor position with a large residual.
The screening guide used in the plot is

$$
\frac{4}{n}.
$$

Values above it prompt investigation rather than automatic deletion.
:::

```{r}
#| label: poverty-influence
poverty_influence <- poverty_results |> mutate(observation = row_number())
ggplot(poverty_influence, aes(x = observation, y = .cooksd)) +
  geom_col() +
  geom_hline(yintercept = 4 / nrow(poverty_influence),
             linetype = "dashed") +
  labs(x = "Observation", y = "Cook's distance",
       title = "Which counties most influence the poverty-rate model?",
       subtitle = "The dashed 4/n line is a screening guide") +
  theme_minimal(base_size = 13)
```

```{r}
#| label: poverty-most-influential
poverty_influence |>
  arrange(desc(.cooksd)) |>
  select(observation, pov_norm, Mortality_Rate_num, .fitted, .resid, .cooksd) |>
  slice_head(n = 5)
```

::: {.callout-tip title="Worked interpretation"}
The largest poverty-model Cook's distances are around 0.043, with a small
number of clear spikes. Inspect those county records, check their data quality
and refit the model as a sensitivity check before considering any removal.
:::

### Income model

::: {.callout-important title="Question"}
Which income observations have the largest Cook's distances, and do their
values reflect unusual income, large residuals or both?
:::

::: {.callout-note title="What to look for"}
Compare the largest income values with the counties highlighted by the poverty
model, then consider a sensitivity refit before drawing a conclusion.
:::

```{r}
#| label: income-influence
income_influence <- income_results |> mutate(observation = row_number())
ggplot(income_influence, aes(x = observation, y = .cooksd)) +
  geom_col() +
  geom_hline(yintercept = 4 / nrow(income_influence),
             linetype = "dashed") +
  labs(x = "Observation", y = "Cook's distance",
       title = "Which counties most influence the income model?",
       subtitle = "The dashed 4/n line is a screening guide") +
  theme_minimal(base_size = 13)
```

```{r}
#| label: income-most-influential
income_influence |>
  arrange(desc(.cooksd)) |>
  select(observation, med_income, Mortality_Rate_num,
         .fitted, .resid, .cooksd) |>
  slice_head(n = 5)
```

::: {.callout-tip title="Worked interpretation"}
The income model also has a few Cook's-distance spikes worth checking. The
counties need not be the same as those flagged by poverty because leverage
depends on the predictor. The screening threshold does not direct us to delete
any observation; it identifies cases for contextual and sensitivity checks.
:::

## 5. Does logging the predictor improve the model?

::: {.callout-important title="Question"}
Does logging poverty or income reduce the diagnostic problems enough to make a
more useful model?
:::

Logging a predictor can straighten a curved relationship while leaving the
response on its original mortality scale. In the model

$$
y=a+b\log(x),
$$

a 1% increase in the predictor is associated approximately with

$$
0.01b
$$

units of the response.

### Log-poverty model

::: {.callout-note title="What to look for"}
Compare the linear and log-predictor models on the same counties. Inspect
curvature, residual magnitude, tail behaviour and influential observations
before choosing on the basis of R-squared alone.
:::

```{r}
#| label: fit-log-poverty
poverty_comparison_data <- cancer_clean |> filter(pov_norm > 0)
poverty_linear_same_rows <- lm(
  Mortality_Rate_num ~ pov_norm, poverty_comparison_data
)
poverty_log_model <- lm(
  Mortality_Rate_num ~ log(pov_norm), poverty_comparison_data
)
stopifnot(nobs(poverty_linear_same_rows) == nobs(poverty_log_model))
tibble(
  model = c("Linear predictor", "Log predictor"),
  observations = c(nobs(poverty_linear_same_rows), nobs(poverty_log_model)),
  r_squared = c(glance(poverty_linear_same_rows)$r.squared,
                glance(poverty_log_model)$r.squared),
  residual_sd = c(glance(poverty_linear_same_rows)$sigma,
                  glance(poverty_log_model)$sigma)
)
```

```{r}
#| label: log-poverty-interpretation
log_poverty_slope <- coef(poverty_log_model)[["log(pov_norm)"]]
tibble(
  approximate_change_for_one_percent_more_poverty =
    0.01 * log_poverty_slope
)
```

```{r}
#| label: log-poverty-diagnostics
augment(poverty_log_model) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.3, size = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Fitted mortality", y = "Residual (deaths per 100,000)",
       title = "Residual pattern after logging poverty") +
  theme_minimal(base_size = 13)
```

::: {.callout-tip title="Worked interpretation"}
Logging poverty raises R-squared modestly, from about 0.158 to 0.171, and makes
the residual smooth somewhat closer to zero over the central range. Changing
spread, tail departures and influential observations remain. In

$$
y=a+b\log(x),
$$

multiplying poverty by 1.01 changes fitted mortality by exactly

$$
b\log(1.01),
$$

which is approximately

$$
0.01b
$$

deaths per 100,000. Spatial dependence and performance on new counties remain
unresolved.
:::

### Log-income model

::: {.callout-important title="Question"}
Does the log-income model improve the residual pattern and fit on the same
counties, and how would we interpret a 1% relative income increase?
:::

::: {.callout-note title="What to look for"}
Check that both models use identical observations. Compare curvature, changing
spread, Q-Q tails, Cook's-distance spikes and overall fit rather than relying
on one diagnostic.
:::

```{r}
#| label: fit-log-income
income_comparison_data <- cancer_clean |> filter(med_income > 0)
income_linear_same_rows <- lm(
  Mortality_Rate_num ~ med_income, income_comparison_data
)
income_log_model <- lm(
  Mortality_Rate_num ~ log(med_income), income_comparison_data
)
stopifnot(nobs(income_linear_same_rows) == nobs(income_log_model))
tibble(
  model = c("Linear predictor", "Log predictor"),
  observations = c(nobs(income_linear_same_rows), nobs(income_log_model)),
  r_squared = c(glance(income_linear_same_rows)$r.squared,
                glance(income_log_model)$r.squared),
  residual_sd = c(glance(income_linear_same_rows)$sigma,
                  glance(income_log_model)$sigma)
)
```

```{r}
#| label: log-income-interpretation
log_income_slope <- coef(income_log_model)[["log(med_income)"]]
tibble(
  approximate_change_for_one_percent_more_income =
    0.01 * log_income_slope
)
```

```{r}
#| label: log-income-diagnostics
augment(income_log_model) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.3, size = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Fitted mortality", y = "Residual (deaths per 100,000)",
       title = "Residual pattern after logging income") +
  theme_minimal(base_size = 13)
```

The comparison uses the same positive-income counties in both models. The
ordered model label keeps the four diagnostic figures directly comparable.

```{r}
#| label: log-income-comparison-prep
income_comparison_results <- bind_rows(
  augment(income_linear_same_rows) |> mutate(model = "Linear predictor"),
  augment(income_log_model) |> mutate(model = "Log predictor")
) |>
  mutate(
    model = factor(model, levels = c("Linear predictor", "Log predictor")),
    observation = rep(seq_len(nrow(income_comparison_data)), 2)
  )
```

```{r}
#| label: log-income-diagnostics-comparison
#| fig-cap: "Residuals versus fitted values for the linear and log-income models."
#| fig-alt: "Two faceted residual plots comparing linear and log income predictors."
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "100%"
ggplot(income_comparison_results, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.2, size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ model, nrow = 1) +
  labs(x = "Fitted mortality", y = "Residual (deaths per 100,000)") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: log-income-scale-location
#| fig-cap: "Scale-location diagnostics for the two income models."
#| fig-alt: "Faceted scale-location plots showing residual magnitude against fitted mortality."
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "100%"
ggplot(income_comparison_results,
       aes(x = .fitted, y = sqrt(abs(.std.resid)))) +
  geom_point(alpha = 0.2, size = 0.5) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ model, nrow = 1) +
  labs(x = "Fitted mortality", y = "Square root of absolute standardized residual") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: log-income-qq
#| fig-cap: "Normal Q-Q diagnostics for the two income models."
#| fig-alt: "Faceted Q-Q plots comparing theoretical normal quantiles with standardized residuals."
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "100%"
ggplot(income_comparison_results, aes(sample = .std.resid)) +
  stat_qq(alpha = 0.2, size = 0.6) +
  stat_qq_line() +
  facet_wrap(~ model, nrow = 1) +
  labs(x = "Theoretical normal quantiles", y = "Ordered standardized residuals") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: log-income-influence
#| fig-cap: "Cook's distances for the two income models."
#| fig-alt: "Faceted Cook's-distance plots with a 4 over n screening line."
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "100%"
ggplot(income_comparison_results, aes(x = observation, y = .cooksd)) +
  geom_col(width = 1, alpha = 0.55) +
  geom_hline(yintercept = 4 / nrow(income_comparison_data),
             linetype = "dashed") +
  facet_wrap(~ model, nrow = 1) +
  labs(x = "Observation index (same counties in both panels)",
       y = "Cook's distance") +
  theme_minimal(base_size = 12)
```

| Evidence | Linear predictor | Log predictor | Interpretation |
|:--|:--|:--|:--|
| Residual curvature | Smooth above zero at low fitted values and below in the middle | Closer to zero over much of the range | Logging reduces, but does not eliminate, curvature. |
| Changing residual magnitude | Spread changes across fitted values | U-shape and high-end rise remain | Interval precision still varies. |
| Q-Q tail departures | Both tails depart | Both tails depart | Normal-error approximation remains imperfect. |
| Influential observations | A few Cook's-distance spikes | A few spikes remain | Inspect records and sensitivity before deletion. |
| R-squared and residual SD | 0.200; larger SD | 0.217; smaller SD | Log model improves fit modestly on identical rows. |

::: {.callout-tip title="Worked interpretation"}
The log-income model raises R-squared from about 0.200 to 0.217 and reduces the
residual standard deviation. Its residual smooth is closer to zero across much
of the range, although changing spread, tail departures and Cook's-distance
spikes remain. A 1% income increase is associated approximately with
-0.263 deaths per 100,000; that is a change in
mortality units rather than a percentage change in mortality.
:::

## 6. Which model would we use, and with what limitations?

::: {.callout-important title="Question"}
Which income model would we use, and what limitations must accompany that
recommendation?
:::

::: {.callout-note title="What to look for"}
Weigh linearity, equal variance, normality, influence and relative fit on the
same counties. Include the unresolved independence concern and the lack of
out-of-sample validation.
:::

::: {.callout-tip title="Worked interpretation"}
The log-income model is a reasonable recommendation: its R-squared improves from
about 0.200 to 0.217 and its residual smooth is closer to zero over much of the
observed range. The scale-location plot still shows changing spread, both Q-Q
tails depart from the reference line, and Cook's-distance spikes deserve
investigation. Possible spatial dependence and the lack of validation on new
counties remain unresolved. The simpler linear model could also be defended if
those modest improvements do not outweigh its more direct interpretation.
:::

## Conclusion

The Week 11 associations remain informative, but the diagnostics qualify how
confidently the models support inference and prediction. Neither model gives a
mechanical pass or fail: the evidence points to modest curvature, changing
spread, non-normal tails, influential counties and possible spatial dependence.
Logging a predictor improves some features without resolving them all.

::: {.callout-tip title="Optional question"}
If a transformation improves the diagnostics on these counties, does that
guarantee better predictions for new counties?

No. Better in-sample diagnostics show that the transformed model represents
these counties more suitably in some respects. Prediction for new counties
must be assessed using counties that did not determine the fitted model, and it
can still be weakened by population differences, changing relationships or
unresolved spatial dependence.
:::
